### 2017-07-26 cover letter/cover.tex
initial

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{vmargin}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{url}
\usepackage{hyperref}

\setmarginsrb{2.25cm}{0.5cm}{2.25cm}{0.5cm}
{12pt}{20pt}{12pt}{36pt}

%----------------------------------------------------
% Sender info
%----------------------------------------------------
\def\Who{Dr. Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m} 
\def\What{}
\def\Where{Embedded Software Group}
\def\Affiliation{Delft University of Technology}
\def\Address{Mekelweg 4, 2628 CD, Delft, The Netherlands}
\def\Email{E-mail:  k.s.yildirim@tudelft.nl}
\def\Telephone{Phone: +31 15 278 9375}
\def\Webpage{\url{http://www.st.ewi.tudelft.nl/~sinan}}

\def\Sender{

	\begin{flushright}
		\Who \What \\
		\Where \\
		\Affiliation \\
		\Address \\
		\Email \\
		\Telephone \\
		\Webpage
	\end{flushright}	
}

\begin{document}

	\thispagestyle{empty}

	\Sender

	\bigskip

	Dear Editor,

	With this letter we are submitting an article titled \emph{On the 
		Synchronization 
		of Intermittently Powered Wireless Embedded Systems}.
	The objective of this article is to provide initial observations pertaining to 
	the synchronization of battery-free computational RFID platforms. In 
	particular, we design and implement the first synchronization protocol for the 
	WISP (Wireless 
	Identification and Sensing Platform) and provide real-world experiments, as 
	well as mathematical analysis to evaluate its performance. We also 
	provide an overview of new 
	research opportunities in this emerging domain.

	A preliminary version of this article was recently presented at the \emph{HLPC 
		2016--Hilariously 
		Low-Power Computing Pushing the Boundaries of Intermittently Powered Devices 
		Workshop}, 
	in Atlanta, GA, USA, 2 April, 2016; which is available online at 
	\href{https://arxiv.org/abs/1606.01719}{https://arxiv.org/abs/1606.01719} and 
	also attached to this submission. This submission extends the above mentioned 
	paper by providing 
	\begin{itemize}
		\item a more detailed related work in Section II;
		\item extended descriptions of the hardware platform and testbed in Section 
		III;
		\item a mathametical framework in Section IV to analyze the proposed 
		synchronization algorithms;
		\item extended descriptions of the proposed synchronization algorithms in 
		Section V and VI;
		\item a rigorous theoretical analysis of the proof of convergence and 
		steady-state error performance of the proposed algorithm in Appendix;
		\item an experimental work to evaluate the synchronization
		performance of the proposed solutions under the loss of computational 
		state in Section VII-C;	
		\item an experimental work to evaluate the energy overhead of the proposed solutions in Section VII-D;	
		\item implementation and evaluation of the synchronization of 
		many WISP tags in Section VIII.
	\end{itemize}

	We believe that \emph{IEEE Transactions on Communications} is 
	the best 
	venue to present our work, as its scope 
	perfectly matches with the scope of our contribution. Thank you in advance for 
	considering our submission.

	\vspace{0.3cm}

	Sincerely,

	\Who \\
	Corresponding Author

	encl: \\
	- Article for review \\
	- The preliminary version of our article which has been presented at HLPC 	
	2016 Workshop\\

\end{document}

### 2017-07-26 figures/Event-Based.tex
initial

\usetikzlibrary{arrows}
\usetikzlibrary{decorations.pathreplacing}

\begin{tikzpicture}[font=\sffamily,
>=stealth',
commentl/.style={text width=3cm},
scale=1.0, 
every node/.style={scale=1.0}]
\tikzstyle{bl} = [draw,rotate=90,minimum height=5mm,minimum width=21.5mm, very thick, draw=black, top color=white]
\tikzstyle{BL} = [draw,rotate=90,minimum height=10mm,minimum width=21.5mm, very thick, draw=black, top color=white,text width=1.8cm,align=center]

\node(host) at (0.5,3.5cm){\textbf{Host}};
\node(reader) at (0.5,2cm){\textbf{Reader}};
\node(tag) at (0.5,0cm){\textbf{Tag}};
\node(h) at (0,3.5cm){};
\node(r) at (0,2cm){};
\node(t) at (0,0cm){};

\foreach \x in {0,2cm,3.5cm}{
	\node (v1) at (10mm,\x) {};	
	\node (v2) at (110mm,\x) {};
	\draw [dotted] (v1) edge (v2);
}

\draw[->] ([xshift=13mm]h.east)  --
([xshift=18mm]r.east)node[midway, below, sloped]     { AccessSpec};

\node[bl](x) at (2.3cm,1cm){\scriptsize{Handshake (EPC)}};
\node[BL](b0) at ([xshift=9mm]x){\scriptsize{BlockWrite (start frame)}};
\node[BL](x) at ([xshift=11.5mm]b0){\scriptsize{BlockWrite (frame 1)}};

\node[BL](b7) at ([xshift=20mm]x){\scriptsize{BlockWrite (frame 7)}};
\draw[loosely dotted,ultra thick] ([xshift=2mm]x.south) --([xshift=-2mm]b7.north);

\draw[->] ([xshift=70mm]r.east)--
([xshift=75mm,yshift=1.5cm]r.east) node[midway, above, sloped] { ASReport};

\draw[->] ([xshift=80mm]h.east)  --
([xshift=85mm]r.east) coordinate(as2) node[midway, below, sloped]     { AccessSpec};
\node[bl](x) at (90mm,1cm){\scriptsize{Handshake (EPC)}};
\node[bl](x) at ([xshift=6.5mm]x){\scriptsize{Read( frames )}};
\draw[->] ([xshift=15mm]as2.east)--
([xshift=20mm,yshift=1.5cm]as2.east) node[midway, above, sloped] { ASReport};

%\tikzstyle{br} = [decorate, decoration={brace, amplitude=4pt}]
%\draw[br] ([xshift=19mm]t.east) -- ([xshift=13mm]t.east) node[midway, , sloped,yshift=-6mm] {Other slots};
%\draw[br] ([xshift=49mm]t.east) -- ([xshift=43mm]t.east) node[midway, , sloped,yshift=-6mm] {Other slots};
% Timestamp stuff:
\node(tTag) at ([yshift=-6mm,xshift=4mm]b0.west){TimeStamp};
\draw[->] ([yshift=2.5mm,xshift=-10mm]tTag.east)--([yshift=6mm,xshift=-10mm]tTag.east) node[near start]     {};

\node(tTag) at ([yshift=-6mm,xshift=4mm]b7.west){TimeStamp};
\draw[->] ([yshift=2.5mm,xshift=-10mm]tTag.east)--([yshift=6mm,xshift=-10mm]tTag.east) node[near start]     {};

\end{tikzpicture}

### 2017-07-26 figures/Sender-Receiver.tex
initial

\usetikzlibrary{arrows}
\usetikzlibrary{decorations.pathreplacing}

\begin{tikzpicture}[font=\sffamily,
>=stealth',
commentl/.style={text width=3cm},
scale=1.0, 
every node/.style={scale=0.75}]
\tikzstyle{myedgestyle} = [dotted]
\tikzstyle{ar} = [->]
\node(host) at (-0.5,3.5cm){\textbf{Host}};
\draw [myedgestyle] (host) edge ([xshift=8.8cm]host);
\node(reader) at (-0.5,2cm){\textbf{Reader}};
\draw [myedgestyle] (reader) edge ([xshift=8.8cm]reader);
\node(tag) at (-0.5,0cm){\textbf{Tag}};
\draw [myedgestyle] (tag) edge ([xshift=8.8cm]tag);
\node(h) at (-0.4,3.5cm){};
\node(r) at (-0.4,2cm){};
\node(t) at (-0.4,0cm){};

\draw[ar] ([xshift=5mm]h.east) coordinate(as1) --
([xshift=10mm]r.east) coordinate(as2) node[midway, below, sloped]     { AccessSpec};
\draw[ar] ([xshift=1mm]as2.east) coordinate(Q1)--
([xshift=16mm]t.east) coordinate(Q2) node[midway, below, sloped]     { Query};
\draw[ar] ([xshift=1mm]Q2.east) coordinate(R1)--
([xshift=11mm]Q1.east) coordinate(R2) node[midway, above, sloped]     { Reply};
\draw[ar] ([xshift=1mm]R2.east) coordinate(A1)--
([xshift=11mm]R1.east) coordinate(A2)node[midway, below, sloped]     { ACK};
\draw[ar] ([xshift=1mm]A2.east) coordinate(E1)--
([xshift=11mm]A1.east)coordinate(E2) node[midway, above, sloped]     { EPC};
\draw[ar] ([xshift=4mm]E2.east)coordinate(Rq1)--
([xshift=14mm]E1.east)coordinate(Rq2) node[midway, below, sloped]     { Req\_rn};
\draw[ar] ([xshift=1mm]Rq2.east)coordinate(nR1)--
([xshift=11mm]Rq1.east)coordinate(nR2) node[midway, above, sloped]     { newRN};
\draw[ar] ([xshift=1mm]nR2.east)node (A1) {}--
([xshift=11mm]nR1.east)node (A2) {} node[midway, below, sloped]     { ACK};
\draw[ar] ([xshift=1mm]A2.east)coordinate(Er1)--
([xshift=11mm]A1.east)coordinate(Er2) node[midway, above, sloped]     { EPC};
\draw[ar] ([xshift=4mm]Er2.east)coordinate(c1)--
([xshift=14mm]Er1.east)coordinate(c2) node[midway, below, sloped]     { Read('timer')};
\draw[ar] ([xshift=2mm]c2.east)coordinate(r1)--
([xshift=12mm]c1.east)coordinate(r2) node[midway, above, sloped]     { Reply(timer)};
\draw[ar] ([xshift=1mm]r2.east)coordinate(AR1)--
([xshift=6mm,yshift=1.5cm]r2.east)coordinate(AR2) node[midway, above, sloped] { ASReport};

\tikzstyle{br} = [decorate, decoration={brace, amplitude=8pt}]
\draw[br] ([xshift=5mm]E1) -- ([xshift=-5mm]Q2) node[midway, , sloped,yshift=-6mm] {Handshake};
\draw[br] ([xshift=5mm]Er1) -- ([xshift=-5mm]Rq2) node[midway, , sloped,yshift=-6mm] {Cmd init};
%\draw[br] ([xshift=5mm]r1) -- ([xshift=-5mm]c2) node[midway, , sloped,yshift=-6mm] {Cmd};
%\draw[br] ([xshift=5mm,yshift=-6mm]r1) -- ([xshift=-5mm,yshift=-6mm]Rq2) node[midway, , sloped,yshift=-6mm] {Optional};

% Timestamp stuff:
\node(tTag) at ([yshift=-5mm,xshift=0mm]c2.east){Timestamp at $t_2$};
\draw[ar] ([yshift=-15mm,xshift=2mm]tTag)--
([xshift=0mm]c2.east) node[near start]     {};

\node(tReader) at ([xshift=0mm,yshift=7mm]E2.east) {Timestamp at $t_1$};
\draw[ar] (tReader)--
([xshift=0mm]E2.east) node[near start]     {};

\end{tikzpicture}

### 2017-07-26 manuscript.tex
initial

\documentclass[journal,draftcls,onecolumn,12pt,twoside]{IEEEtranTCOM}

\normalsize
%\documentclass[english,journal]{IEEEtran}

% Metadata Information
%\acmVolume{9}
%\acmNumber{4}
%\acmArticle{39}
%\acmYear{2016}
%\acmMonth{3}

% Copyright
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
%\doi{0000001.0000001}

%ISSN
%\issn{1234-56789}

%\usepackage[latin9]{inputenc}
%\usepackage{hyphenat}
%\usepackage{epstopdf}
%\usepackage{flushend}
\usepackage{algorithm,setspace}
\usepackage[noend]{algpseudocode} % needed for: algorithm pseudocode
\usepackage{color}
\usepackage{array}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{url}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{soul}
\usepackage{cancel}
\usepackage{listings}
\usepackage{tikz} % needed for: tikz pics
\usetikzlibrary{arrows}
\usetikzlibrary{decorations.pathreplacing}

\definecolor{aliceblue}{rgb}{0.94, 0.97, 1.0}
\definecolor{applegreen}{rgb}{0.55, 0.71, 0.0}
\definecolor{amber}{rgb}{1.0, 0.75, 0.0}

\lstdefinestyle{mystyle}{
	tabsize          = 1,     % tab space width
	numbers          = left,  % display line numbers on the left
	%stepnumber		 = 2,
	numberfirstline=false,
	%framesep         = 1pt,   % expand outward
	framerule        = 0.2pt, % expand outward 
	commentstyle     = \color{applegreen},      % comment color
	keywordstyle     = \color{blue},       % keyword color
	stringstyle      = \color{amber},    % string color
	backgroundcolor  = \color{aliceblue}, % backgroundcolor color
	showstringspaces = false,              % do not mark spaces in strings
	otherkeywords	 ={signal,inos_task,uint16_t},
	language = C,
	%numbers = none, 
	escapechar = !,
	basicstyle = \ttfamily\bfseries\scriptsize, 
	xleftmargin=0.5ex,
	linewidth = .75\linewidth,
	basicstyle=\tiny,
}

\newcommand{\expected}[1]{E\left\llbracket #1 \right\rrbracket}
\newcommand{\revised}[1]{\textcolor{blue}{#1}}
\newcommand{\TODO}[1]{\textcolor{blue}{\textbf{TODO: }}\textcolor{red}{#1}}
\newtheorem{fact}{Fact}

% text revision commands
%\newcommand{\add}[1]{\textcolor{blue}{#1}}
%\newcommand{\remove}[1]{\textcolor{red}{\st{#1}}}
%\newcommand{\replace}[2]{\remove{#1} \add{#2}}

%\makeatother

\setcounter{secnumdepth}{6}

% Document starts
\begin{document}

% Title portion
\title{On the Synchronization of Intermittently Powered Wireless Embedded Systems}

\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, Henko~Aantjes, Przemys{\l}a~Pawe{\l}czak and Amjad~Yousef~Majid

\thanks{A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}.}
\thanks{The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).}}

\markboth{IEEE Transactions on Communications}%
{Submitted paper}

\maketitle

\vspace{-1cm}

\begin{abstract}
Battery-free computational RFID platforms, such as WISP (Wireless Identification and Sensing Platform), are intermittently-powered devices designed for replacing existing sensor networks. Accordingly, synchronization appears as one of the  crucial building blocks for collaborative and coordinated actions in these platforms. However, intermittent power leads to frequent loss of computational state and short-term clock frequency instability that makes synchronization challenging. In this article, we introduce WISP-Sync protocol that provides synchronization among WISP tags in the communication range of an RFID reader. WISP-Sync overcomes the aforementioned challenges by employing a Proportional-Integral (PI) controller-inspired algorithm which (i) is adaptive---reactive to short-term clock instabilities; (ii) requires only a few computation steps---suitable for limited harvested energy; and (iii) keeps a few variables to hold the synchronization state---minimum overhead to recover from power interrupts.	Evaluations in our testbed showed that WISP-Sync ensured an average synchronization error of approximately 1\,ms among the tags with an average energy overhead of 1.85\,$\mu$J per synchronization round.
\end{abstract}

\begin{IEEEkeywords}Time Synchronization, Computational RFIDs, Wireless Identification and Sensing Platform (WISP).
\end{IEEEkeywords}

\IEEEpeerreviewmaketitle

\section{Introduction}
\label{sec:Intro}

\bstctlcite{BSTcontrol}\IEEEPARstart{L}{ow-power} wireless embedded systems consist of tiny, low-cost and spatially separated computers that communicate with each other by exchanging radio packets. Powering these small-scale embedded systems, e.g. wireless sensor networks (WSNs), is still a crucial problem~\cite{emergence}. Replacing or recharging their batteries is impractical and inhibits long-term operation. Moreover, batteries increase the size and cost of the hardware. Fortunately, the energy efficiency of these systems has improved considerably such that their power requirements are in the order of a few $\mu$W \cite{booksmith2013}. Furthermore, recent advancements in microelectronics technology enabled harvesting power from radio frequency (RF) sources that is sufficient to operate low-power embedded systems in practice~\cite{60ghz,emergence,bi:commag:2015,xie_wcom_2013,garnica_procieee_2013}. Nowadays, the growth of the RF-powered computing paradigm is bringing new research opportunities and challenges~\cite{emergence}, leading to a promising class of low-power embedded systems, the so-called Intermittently Powered Devices (IPDs).

By taking existing RFID (Radio Frequency IDentification) technology as a foundation, computational RFIDs (CRFIDs) are emerging IPDs that allow sensing, computation and communication without batteries---replacing existing battery-powered sensor networks~\cite{booksmith2013}. CRFIDs are equipped with a backscatter radio~\cite[Chapter 4]{rfid_handbook} composed of a simple circuitry that modulates the carrier wave generated by a reader to transmit information. This allows communication to come almost for free. This is a fundamental difference from sensor networks where the transceiver circuit is the most energy-hungry component. In the CRFID domain, the bottleneck in terms of power consumption has shifted from communication to computation and sensing~\cite{ekhonet}. A typical example of CRFIDs is the WISP (Wireless Identification and Sensing Platform)~\cite{wisp}. Commercial RFID readers implementing the EPC Gen2 standard~\cite{epc_gen2} are used to power WISP tags. Apart from low data rate sensing~\cite{neuralWISP}, these maintenance-free devices are evolving to support high data rate and more complex sensing applications such as continuous sensor-data streaming, e.g. capture and transfer of images via battery-free cameras, i.e. WISPCam~\cite{wispcam_2015,cameralocalize_2015}.

 As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}. However, recent advancements in the wireless communication enable tag-to-tag communication, e.g.~\cite{liu2013ambient}. We anticipate that these multi-hop networks will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts. In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences.

%\add{\textbf{Distributed Sensing.} As an example, consider applications of monitoring inaccessible areas or general surveillance in which multiple battery-less cameras, such as WISPCam~\cite{wispcam_2015}, capture images from different locations simultaneously and forward the collected data to the RFID reader. In such applications, the WISPCams can coordinate their image capture actions by means of the almost simultaneously received periodic ``\texttt{perfrom image snapshot}'' commands from the RFID reader. However, the reception and processing of the commands, and in turn the coordination, can be interrupted since WISPCams might exhibit different \emph{charge/discharge} patterns. In principle, WISPCams operate in a power cycling mode, in which their storage capacitor is fully-charged in the charging phase and then discharged until a lower threshold (no operation) voltage is attained. Due to their different distances to the RFID reader leading to different harvested energy levels, it can happen that a WISPCam would have to wait until it is fully-charged to capture the image. It has been observed that ``the charging time for voltage to reach 1.8\,V to power a WISP equipped with a 100\,$\mu$F capacitor can be as large as 155 seconds, when the RFID reader is 10.0 meters away''~\cite{fu2013minimizing}. This extra delay might result in the loss of coordination and the capture of images at different time instants. To prevent this situation, the reader can send ``\texttt{perform image snapshot $x$ seconds later}' command and continue charging the WISPCams. Then, WISPCams can activate themselves and capture the image simultaneously, just after $x$ seconds after the reception of the reader command. Since each WISPCam has its own clock hardware whose oscillator generate pulses at slightly different speeds, a \emph{synchronization} service that allows WISPCams to wake-up simultaneously and facilitates such collaborative and coordinated actions becomes a fundamental requirement for these systems.}

%that the charging time for voltage to reach 1.8V to power a WISP equipped with a 100uF capacitor can be as large as 155 seconds, when the RFID reader is 10.0 meters away

%It should be noted that there can also be a degree of synchronization on the 
%reader side by timestamping the received 
%information. However, the WISPCams that experience a power 
%loss might resume transmission of the captured image to the reader after a 
%relatively long delay due to harvesting sufficient energy. In this case, the 
%timestamps do not make sense since they do not reflect the actual capture time 
%of the images. Therefore, the establishment of a ``common time'' notion

\label{sec:Motivation}

The characteristics of CRFID systems expose fundamentally different challenges than sensor networks to implement the synchronization service. The reason is mainly twofold:

\begin{itemize}
\item \textbf{Challenge I:} CRFID systems should perform computations in an energy efficient manner despite intermittent RF power that leads to frequent loss of computational state, e.g. when the RFID reader moves away from the CRFID tag \cite{ransford:2008}.

\item \textbf{Challenge II:} The continuously varying voltage supply introduces severe hardware instability, e.g. varying oscillator frequencies in short-term affects the stability of the clocks and degrades the accuracy sensing.	
\end{itemize}

The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered wireless embedded systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices. In particular, the main contributions and findings of 
this article are:
%
\begin{itemize}
\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems. Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides explicit synchronization of individual tags in the communication range of an RFID reader. 
\item By addressing the aforementioned challenges, we design and implement\emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015}. WISP-Sync (i) 
requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
%
\item We provide theoretical analysis to prove that this primitive establishes synchronization and to reveal the factors affecting its synchronization performance.
%
\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
\end{itemize}
%
We note that WISP-side implementations throughout the article are done using C and assembly language. All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\href{https://goo.gl/RuWIZ2}{https://goo.gl/RuWIZ2}.

The remainder of this article is organized as follows. In Section~\ref{sec:Related-Work} we present the fundamental challenges of synchronizing IPDs. Section~\ref{sec:Clock_Hardware} presents the hardware, in particular clock system of the WISP platform briefly, while Section~\ref{sec:clock_model} provides the mathematical notation to be used throughout the article. In 
Section~\ref{sec:sender_receiver} we propose a sender-receiver synchronization approach between an RFID reader and a single tag. Sections~\ref{sec:event_based} and~\ref{sec:eval-wispsync} present the WISP-Sync protocol and its evaluation in our testbed, respectively. We also present the synchronization of multiple tags in Section~\ref{sec:multi-wisp}. Finally, we provide our conclusions and future research directions in Section~\ref{sec:Conclusion}.

\section{Synchronizing Battery-Free Wireless Embedded Systems}
\label{sec:Related-Work}

In this section, we present a brief  state of the art on synchronization issues in conventional WSNs. Then, we emphasize the 
main challenges of synchronizing wireless battery-free embedded systems by focusing on a particular IPD platform: computational RFIDs.

\subsection{Synchronization in Conventional Wireless Sensor Networks}
\label{sec:WSN-Sync}

The instability of the clock hardware, delays during communication among sensor nodes, and software methods to establish synchronization are the main factors affecting the synchronization in conventional WSNs.

\subsubsection{Clock Hardware}

In WSNs, each sensor node is equipped with a built-in clock that is implemented as a counter register clocked by a low-cost external crystal oscillator. At each oscillator pulse, i.e. \emph{tick} of the clock, the counter register is incremented. The duration between two consecutive ticks is the \emph{rate} of the built-in clock. Environmental factors such as temperature, supply voltage and aging of the crystal prevent built-in clocks to generate ticks at the exact speed of real-time, leading to bounded \emph{clock drift}. The prominent environmental factor affecting the frequency of the built-in clocks is the temperature~\cite{Schmid:2009,Schmid:2010}. Moreover, \emph{quantization errors} occur with low-frequency built-in clocks, which prevents precise timing measurements.

\subsubsection{Wireless Communication}

Sensor nodes exchange their clock information periodically to compute a \emph{software clock} that represents the synchronized notion of time. A software clock is composed of an offset and a frequency that hold the value and speed difference between the corresponding built-in clock and the reference time, respectively. The difference between the reference time and the software clock is the \emph{synchronization error}. In WSNs, the synchronization error is affected by several sources of errors. The \emph{transmission delay}, defined as the time that passes between the start of the broadcast and the receipt by the receiver node, is the major error source and composed of deterministic and non-deterministic components~\cite{Maroti2004}. Assigning timestamps at the MAC layer removes the deterministic delay components and improves synchronization accuracy. This obligates the use of transceivers that allow assignment of time information to a radio packet just before transmission and reception, e.g. Chipcon CC2420~\cite{chipcon:cc2420}. Since the transceiver is the most power-hungry circuit, nodes in WSNs should reduce re-synchronization frequency to decrease communication overhead and save power.

\subsubsection{Computation Methods}

The most common synchronization mechanism is to propagate the time information of a particular \emph{reference node} to let receiver nodes synchronize themselves to the received reference time information. Commonly, \emph{least-squares regression} is employed by several practical synchronization protocols to adjust the offset and the frequency of the software clocks ~\cite{Maroti2004,Lenzen:2015,TATS:2016}. On the other hand \emph{iterative} computation methods, e.g. in ~\cite{Yildirim:AVTS:2014,pi2015} and \cite{Yildirim:Gradient:2016}, perform identical computation steps at each re-synchronization round. With iterative methods, sensor nodes establish synchronization after a finite number of rounds in a lightweight manner in terms of computation and memory requirements. There are also \emph{fully distributed} approaches in which sensor nodes interact only with and synchronize to their direct neighbors in a peer-to-peer fashion without the requirement of a dedicated reference node. In these approaches, nodes employ computation methods based on \emph{distributed consensus}~\cite{Sommer2009Gradient,SchenatoFiorentin:2011,Yildirim:External:2014}.

\subsection{Fundamental Challenges of Synchronizing IPDs: The CRFID Case}
\label{sec:CRFIDs}

We now delve into the synchronization characteristics of RF-powered wireless embedded systems. To this end, we will consider the WISP, the de facto CRFID platform. The main aspects pertaining to synchronization in the WISP platform can be stated as follows:

\paragraph{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication. Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}. Since WISP tags are unable to communicate with their neighboring nodes directly using backscatter communication as of now, \emph{tag-tag synchronization} is impossible.

\begin{figure}
	\centering
	\includegraphics[scale=0.35]{figures/clock_frequency_dc_rf.eps}
	\caption{\label{fig:frequency_dc_rf} In our testbed, see Section~\ref{sec:testbed} for the details, we programmed a single WISP tag so that it toggles one of its output port whenever its 32~KHz built-in timer progresses 32000 clock ticks. In the ideal case where the crystal oscillator of the tag does not drift, the tag would toggle its output port at each 1 second, i.e. with a frequency of 1\,Hz. During our experiment under a stable environmental temperature, we connected the output port of the tag to a logic analyzer and measured the frequency of the toggle event. We plotted our measurements when the tag is powered by the Flash Emulation Tool (FET) with a stable DC power and when the tag is powered by the RFID Reader at a distance of 5~cm. Since harvested RF energy was different and unstable than the constant DC case, the frequency of the oscillator and in turn the toggle event was quite different. We measured the mean and standard deviation of the frequency in the DC case as 1.1653 and 0.00028553, and these values for the RF case were 1.1793 and 0.00026049, respectively. }
\end{figure}

\paragraph{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term. On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift. Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level rather than the temperature of the WISP tags, see Fig.~\ref{fig:frequency_dc_rf}.

\paragraph{Frequent loss of synchronization state} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g. the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up. However, saving computational state is also an energy consuming task~\cite{blisp}.

\paragraph{Computation and memory overhead sensitivity} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p. 44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}. Due to the intermittent power, lightweight methods in terms of computation and memory are required for the synchronization. Since methods like least-squares regression are computationally heavy and require considerable amount of memory~\cite{pi2015} they should be avoided.

\paragraph{Limitations of the EPC Gen 2 standard} The WISP firmware implements the EPC Gen 2 standard~\cite{epc_gen2} which increases the compatibility with the existing RFID systems. However, the standard introduces limitations, e.g. currently it does not assign timestamps to the radio packets, which is a fundamental requirement to establish synchronization. Moreover, communication delays between the reader and tag are quite dependent on the implementation of this standard by RFID readers. Unfortunately, these issues lead to less accurate synchronization as compared to existing WSN solutions, as shown by our measurements presented in the following sections.

\section{WISP Hardware and CRFID Networks: A Brief Look}
\label{sec:Clock_Hardware}

In this section we present brief information about the WISP hardware, in particular its clock system. For the sake of clarity, we 
also give details about a typical CRFID network, which we used to implement and evaluate our synchronization methods in this article.

\subsection{Low Power Operation Modes and WISP Clock System}
\label{subsec:clock_system}

The WISP 5.0 platform comes with the MSP430FR5969 \cite{msp430fr5969_data_sheet} microcontroller (later called as MSP430 in this article) with FRAM non-volatile memory. The MSP430 clock system includes (i) a 32\,kHz external crystal oscillator, (ii) an internal very-low-power low-frequency oscillator, and (iii) an integrated internal digitally controlled oscillator (DCO). The five built-in 16-bit timers (TA0--TA3, TB0) in this system can be clocked with the auxiliary clock (ACLK) signal that is sourced from the relatively stable external 32\,kHz oscillator. The MSP430 has one active mode and seven software selectable low-power operation modes. Low Power operation Mode 3 (LPM3) is the standby mode where CPU, FRAM and high-frequency peripherals are off; the external 32\,kHz oscillator and the ACKL signal are active. In our target implementations, we configured TB0 to be clocked  with ACKL so that it has 16-bit 32\,kHz precision and runs continuously in LPM3 mode.

\subsection{\label{sec:testbed}An Experimental CRFID System}

\begin{figure}
\centering
	\begin{tikzpicture}
	\node[anchor=south west, inner sep=0] at 
	(0,0){\includegraphics[height=.35\columnwidth]{figures/testbed_resized.jpg}};

	\node [fill= white,circle] at (9,2) { };
	\node [color=black ] at (9,2) { 1 };

	\node [fill= white,circle] at (9,4) {  };
	\node [color=black ] at (9,4) { 2 };

	\node [fill= white,circle] at (4,2) {  };
	\node [color=black ] at (4,2) { 3 };

	\node [fill= white,circle] at (5.5,2) {  };
	\node [color=black ] at (5.5,2) { 4 };

	\node [fill= white,circle] at (8,5) {  };
	\node [color=black ] at (8,5) { 5 };

	\end{tikzpicture}
\caption{\label{fig:crfid_testbed} A single-hop CRFID testbed: (1) an Impinj Speedway R1000 RFID reader, (2) a single WISP 5.1 tag, (3) a host computer, (4) FET programmer and (5) a Laird S9028PCR antenna. The same testbed setup has been used previously in~\cite{wisent:2016} by us to evaluate downstream data transmission in CRFIDs.}
\end{figure}

As presented in Fig.~\ref{fig:crfid_testbed}, a typical CRFID system is composed of an RFID reader, a single WISP tag placed inside the communication range of this reader and a host computer to control this reader. In our testbed, we used a 915\,MHz Impinj Speedway R1000 RFID reader with firmware version 3.2.4 connected to a Laird S9028PCR 8.5\,dBic gain antenna. We placed the WISP tag at the line-of-sight from the reader antenna. For the host-reader control operations, we used sllurp~\cite{sllrp_github}, a LLRP (Low-Level Reader Protocol)~\cite{llrp}  control library written in Python.  To program the WISP tag, we used MSP430 Flash Emulation Tool (FET), in combination with TI Code Composer Studio (CCS), attached to the host.  All of our experimental evaluation presented in the next sections were conducted using this setup at a university office with human presence.

\subsection{\label{sec:sniffer}A Sniffer for the CRFID System}

Communication delays are one of the main sources of errors in synchronization~\cite{Maroti2004} and they should be kept as small as possible. In order to characterize the communication delays between the RFID reader and the WISP tags during backscatter communication and understand the effect of these delays on the synchronization error, we used USRP 210 software-defined radio, another Laird antenna placed at 50\,cm from the tag and the GNU Radio toolkit~\cite{gnuradio} with the custom software to sniff the radio packets. The sniffer was presented first in~\cite{stork:2017} and we refer the reader to this study for more details. Implementing sniffing mechanism allowed us to store the timings of the transmission and reception of the EPC Gen 2 packets, which we later processed offline to measure the actual transmission delays. We will present these measurements in the next sections. 
\section{System Model, Notation and Preliminaries}
\label{sec:clock_model}

In this section, we provide a mathematical framework that models the built-in clocks and in particular the clock drifts of the WISP tags and the RFID reader. We will use these models to prove the correctness of our synchronization algorithms and to derive mathematical expressions that represent their synchronization performance.
\begin{figure*}
\centering
\includegraphics[scale=0.35] {figures/relative_clock_frequency.eps}
\caption{\label{fig:frequency}The frequencies of the toggle events (obtained using the same steps presented in Fig.\ref{fig:frequency_dc_rf}) of two RF-powered WISP tags, placed at 5 cm distance from the RFID reader. On the left the individual event frequencies ($f_1$ and $f_2$) are presented, and their relative clock frequency ($\frac{f_1}{f_2}-1$) is presented on the right. We measured the mean and standard deviation of $f_1$ as 1.1793 and 0.00026049 and those of $f_2$ as 1.1705 and  0.00068239, respectively.}
\end{figure*}
In this framework, we denote a single WISP tag by $w$ and the RFID reader by $r$. The Timer B0 mentioned in Section~\ref{subsec:clock_system} can be considered as the built-in clock, i.e. \emph{local clock}, of the tag $w$, 
which we denote by $C_{w}$. The value of $C_{w}$ at time $t$ can be modeled as
%
\begin{eqnarray}
C_{w}(t) & \triangleq & C_w(t_0)+ \int_{t_0}^{t}f_{w}(u)du, 
\label{eq:local_clock}
\end{eqnarray}
%
where $t_0$ represents the time at which tag $w$ is powered on and  $f_{w}(u)$ represents the \emph{instantaneous oscillator frequency} of $C_{w}$ at time $u$. Since the frequency of the oscillators are affected by environmental factors such as temperature and voltage level, they are not stable~\cite{Lenzen2009Optimal} and $f_{w}$ is not a constant but it is \emph{time varying}, see Fig.~\ref{fig:frequency}. From \eqref{eq:local_clock}, it can be noticed that $C_w$ is \emph{unitless}. Indeed, it can be seen as a counter of oscillation events that are produced at a frequency of $f_w$.

The RFID reader $r$ can be considered as a natural time reference for the tag $w$. We denote the local clock of the RFID reader at time $t$ by $C_{r}(t)$ and its frequency by $f_r(t)$. By collecting (time) information from the RFID reader, the tag $w$ runs a \emph{clock synchronization algorithm} to calculate a \emph{software clock}, denoted by $S_{w}(t)$, whose value represents the synchronized notion of time. The objective of the synchronization algorithm is minimizing the \emph{synchronization error} with respect to the reader $r$ at any time $t$, which is defined as 
%
\begin{eqnarray}
\gamma(t)&\triangleq&S_w(t)-C_r(t). \label{eq:gamma_error}
\end{eqnarray}
%
Considering the individual clock frequencies $f_w$ and $f_r$, we denote the \emph{instantaneous relative clock 
frequency} of the WISP tag 
with respect to the RFID reader at time $t$ as
%
\begin{eqnarray}
f_w^r(t)&\triangleq&\frac{f_w(t)}{f_r(t)}-1. \label{eq:relative_frequency}	
\end{eqnarray}
\section{Sender-Receiver Based Reader-Tag Synchronization in CRFID Systems}
\label{sec:sender_receiver}

In this section we provide initial observations, design and implementation of a \emph{sender-receiver} based synchronization approach that synchronizes a single WISP tag to the RFID reader and we reveal its limitations in current CRFID networks. In sender-receiver based synchronization mechanisms, receiver devices synchronize to the clock of a reference sender device. In order to synchronize itself to the RFID reader with such a mechanism, the WISP tag should obtain several $(C_w(t),C_r(t))$ synchronization points to establish a \emph{relationship} between its local clock $C_w$ and the reader clock $C_r $, represented by its software clock $S_w$. The value $S_w(t)$ will provide an estimate of the reference clock $C_r(t)$ at any time instant $t$.

\begin{figure}
\centering
\includegraphics[scale=0.8]{figures/Sender-Receiver.pdf}
\caption{\label{fig:sender_receiver} The message exchange among the host computer, the RFID reader and the tag for sender-receiver synchronization. The host machine, i.e. a PC,  sends the high level commands to the reader via LLRP \emph{AccessSpec} message and receives the results through an \emph{ASReport}. The RFID reader follows the steps defined in the EPC Gen2 standard: (i) performs a \emph{Handshake} to initialize the communication with the active tag; (ii) performs a \emph{Command Initialization} (Cmd Init) by requesting a random number from the tag (Req\_n) and receiving the random number (newRN); (iii)  performs a \emph{Read} command by sending its request and receiving the timer value. The reader assigns the \emph{FirstSeenTimestamp} to the tag at time $t_1$ and the tag timestamps the command reception event at time $t_2$ with its local clock reading.}
\end{figure}

We explored the EPC Gen2 standard and the LLRP protocol and found out that LLRP assigns a \emph{FirstSeenTimestamp} in UTC (Coordinated Universal Time), which is defined as \emph{``The Reader SHALL set it to the time of the first observation amongst the tag reports that get accumulated in the TagReportData''} \cite[p. 87]{llrp}. From this definition, we assume that this timestamp is assigned by the reader when it receives the EPC during the handshake operation with the corresponding tag, shown in Fig. \ref{fig:sender_receiver} as the timestamp assigned at time $t_1$. Therefore, the \emph{FirstSeenTimestamp} can be considered as $C_r(t_1)$. In order to obtain the corresponding local time $C_w(t_1)$, one strategy is to force the reader to send a special ``synchronization'' command after the handshake so that the tag timestamps the command reception event using its local clock, shown in Fig.\,\ref{fig:sender_receiver} as time $t_2$. As we emphasized in Section~\ref{sec:WSN-Sync}, transmission delays are the main source of errors in time synchronization and they should be kept as deterministic as possible. We denote the \emph{transmission delay} in this case by $\Delta t= t_2-t_1$ and it is desirable to keep the \emph{variation} of $\Delta t$, i.e. \emph{jitter}, as small as possible~\cite{Maroti2004}.

\paragraph{Transmission Delay and Jitter Measurements}

\begin{figure}
\centering
\includegraphics[scale=0.35]{figures/sender_receiver_uncertainty.eps}
\caption{\label{fig:transmission_delay} The normalized number of occurrences of the transmission delays measured by sniffing the communication between the RFID reader and the WISP tag. We calculated the mean transmission delay and its standard deviation as 1.89\,ms and 0.0164\,ms with a 99\% confidence interval of [1.8874,1.8925] and [0.0148,0.0184], respectively.}
\end{figure}

In order to characterize the transmission delay $\Delta t$, we placed the WISP tag 20 cm away from the antenna using the setup presented in Fig.~\ref{fig:crfid_testbed} and we sniffed the communication between the RFID reader and the WISP tag during the communication scenario of Fig.~\ref{fig:sender_receiver}. Fig.~\ref{fig:transmission_delay} presents a summary of our measurements (with 300 samples gathered). We observed that the transmission delay is distributed with a mean of 1.89\,ms and standard deviation of 0.0164\,ms. Moreover, we observed some outliers, presented at the right hand side of the figure, which we attribute to the EPC Gen2 implementation of the Impinj reader.

\paragraph{Collection of Clock Values for Offline Processing}

After characterizing the variation of the transmission delay, the next step was the collection of the clock values from the RFID Reader and the WISP tag in order to try synchronization mechanisms offline, which allowed to evaluate the performance of sender-receiver synchronization mechanism in CRFID networks. We collected $(C_w(t_2),C_r(t_1))$ pairs for offline processing by controlling the RFID reader to send a \emph{Read} command to the tag and by programming the WISP tag so that it backscatters $C_w(t_2)$ upon receiving this command. The received pairs $(C_w(t_2),C_r(t_1))$ were logged by the host computer.

\subsection{Synchronization Approach 1: Software Clock Computation Using  Least-Squares Regression}

Following studies~\cite{Maroti2004,Schmid:2010,Lenzen:2015,TATS:2016} that use a linear clock model, we assume a \emph{linear} relationship between $C_r$ and $C_w$ and we model the software clock of the WISP tag as $S_w(C_w(t)) = a + b C_w(t)$ where $S_w$ represents an estimator of the reader's clock $C_r$, $a$ is the \emph{offset} and $b$ is the \emph{relative speed} with respect to the local clock $C_w$. To establish such a relationship, we performed \emph{least-squares regression} offline in MATLAB using the collected pairs in the previous subsection. As in \cite{Maroti2004}, since the WISP tag has limited memory, computation capability and energy, at each step only the most recent $N$ pairs are used to estimate the regression line.

\begin{figure}
\centering
\includegraphics[scale=0.35]{figures/least_squares.eps}
\caption{\label{fig:least-squares}Synchronization error by employing least-squares regression on the collected timestamps. We observed a maximum synchronization error of 10 clock ticks between the RFID reader and the WISP tag, leading to 3.2\,ms synchronization accuracy.}
\end{figure}

Formally, let $[C_w(t_k),C_r(t_k)]$ denote the $k$th pair in the log file where $t_k$ denotes the real-time at which 
\emph{FirstSeenTimestamp} has been assigned during the collection of this pair. At each $k$th step, the pairs $\{[C_w(t_k),C_r(t_k)],$\ldots$,[(C_w(t_{k+N-1}),C_r(t_{k+N-1})]\}$ are used to calculate the parameters $a$ and $b$, i.e. the \emph{intercept} and the \emph{slope} of the estimated regression line \cite{book-SheldonRoss}, using \textsf{polyfit} function of MATLAB. We calculated the synchronization error as 
$\gamma(t_{k+N})\triangleq S_w(C_w(t_{k+N}))-C_r(t_{k+N})$
%
that represents the difference between the predicted reference time and the received reference time. In our implementation we used $N=8$ as in~\cite{Maroti2004} and Fig.~\ref{fig:least-squares} presents the synchronization error at each step. We observed a maximum synchronization error of 0.32\,ms in this one-hop network, which is more than one order of magnitude larger than the synchronization performance of the de facto WSN solution \cite{Maroti2004}, which was reported as approximately 10\,$\mu$s.

\paragraph{Limitations of the Approach 1}

In addition to the challenges listed in Section \ref{sec:CRFIDs}, we observed two crucial limitations for the WISP platform, which prevents to build up a sender-receiver synchronization building block:

\emph{Host computer computation:} We are unaware of any EPC Gen2 command that will allow to send the \emph{FistSeenTimestamp} to the tag. Hence, even though we were able to collect $(C_w(t),C_r(t))$ pairs for offline processing, it is not possible for the tag to collect $C_r(t)$ and synchronize itself to the reader. Therefore, with this limitation, only a host computer can collect and log the timestamps, calculate the relationship between the clock of the tag and the clock of the reader and send the data that represents this relationship to the tag for synchronization, making this method only of theoretical nature as of now.

\emph{Lack of broadcast primitive:} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader. Unfortunately, in the current EPC Gen2 and LLRP standard, we were unable to find any mechanism to send a global timestamp that is received by all tags inside the range of the RFID reader to establish synchronization in one step.

\emph{Big and reader dependent transmission delays:} As depicted previously, we observed that transmission delays in our testbed were on the order of milliseconds, see Fig.~\ref{fig:least-squares}, whereas they are reported as a few microseconds in conventional WSN platforms. This is one of the crucial limitations since the smaller the transmission delays are, the better the synchronization.

\section{Event-Based Reader-Tag Synchronization In CRFID Systems}\label{sec:event_based}

In this section, we introduce a simple but novel \emph{event-based} reader-tag synchronization inspired by simple PI controllers and provide its advantages over Approach 1. In event-based synchronization, a common event which is observable by all receiver devices simultaneously is generated by the reference device. Upon receiving events generated at regular intervals, receiver devices can predict the occurrence time of future events. In order to synchronize the tag with such a mechanism, the RFID reader does not send explicit timestamp values as in sender-receiver based synchronization but instead generates events at regular intervals. Upon observing these events, the receiver tag adjusts the rate of its software clock so that it predicts the occurrence of the next event precisely. 
\begin{figure}
\centering
\includegraphics[scale=0.8]{figures/Event-Based.pdf}
\caption{\label{fig:event_based}Event-Based synchronization steps: The tag timestamps successive \emph{BlockWrite} events and adjusts its software clock.}
\end{figure}
We explored the EPC Gen2 standard to see how to generate events at regular intervals and realized that the \emph{BlockWrite} operation allows this feature.\footnote{It is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0, i.e. Miller 1, modulation) in order to get the timing measurements of the regular patterns in the signal. However, this requires using a high-precision clock to obtain microsecond-precision timing measurements. In this study, we used the higher-level properties of the EPC Gen2 standard, rather than exploiting its low-level signal properties. By means of this, we could  use the external 32\,KHz clock that runs even in low-power standby mode (see Section \ref{sec:Clock_Hardware}) to obtain timing measurements.} Fig.~\ref{fig:event_based} presents the steps of event-based synchronization. During the command phase, the EPC Gen2 standard allows a maximum of eight successive \emph{BlockWrite} operations. It is desirable to use the first and the last \emph{BlockWrite} events for synchronization since it is better to compensate frequency differences observed in longer time intervals to adjust the software clock.

\paragraph{Event Period Measurements}
\label{subsec:event_period_uncertainty}

\begin{figure}
\centering
\includegraphics[scale=0.35]{figures/blockwrite_delay.eps}
\caption{\label{fig:block_write_delay}The delay between the first and the last \emph{BlockWrite} event, i.e. \emph{event period}, by considering 20 samples during the communication scenario in Fig.~\ref{fig:event_based}. We measured its mean and standard deviation as 226.7667\,ms and 0.4097\,ms with a 99\% confidence interval of [226.4961,227.0372] and [0.2852,0.6945], respectively.}
\end{figure}

The real-time length between the first and the last \emph{BlockWrite} operation is the \emph{event period}, we denote by $\tau$, and its variation is the main error source. Therefore, it is important to explore its characteristics. We sniffed the communication protocol between the RFID reader and a tag presented in Fig. \ref{fig:event_based}. We took 20 sample measurements about the \emph{event period}, which is summarized in Fig.~\ref{fig:block_write_delay}. According to our measurements, the \emph{event period} was distributed with a mean of 226.76\,ms and standard deviation of 0.41\,ms; respectively. We would like to mention that these values are dependent on the RFID Reader implementation of the EPC Gen2 protocol.

\subsection{Synchronization Approach 2: Event-Based Synchronization Inspired By PI Controllers }

For the computation of the software clock, \emph{adaptive} and \emph{lightweight} solutions are crucial due to the following important requirements:

\begin{itemize}
\item Since the voltage level is not stable and the harvested power is limited, the computations should demand little amount of energy to calculate the software clock, i.e. marginal number of steps---\emph{Addressing Challenge I in Section~\ref{sec:Motivation}}.

\item Since power is intermittent, the number of variables pertaining to the software clock should be marginal so that saving the state of the synchronization to non-volatile memory will demand little time and less energy. Therefore, lightweight computation methods in terms of main memory overhead are required---\emph{Addressing Challenge I in Section~\ref{sec:Motivation}}.

\item Unstable voltage level leads to frequently varying clock frequencies. Therefore, the computation methods should adapt such dynamic conditions in a very fast manner to keep synchronization accuracy stable---\emph{Addressing Challenge II in Section~\ref{sec:Motivation}}
\end{itemize}

Considering these facts, we designed a lightweight and adaptive clock synchronization approach inspired by the PI-controller based solution introduced in~\cite{pi2015}. Let $t_0$ denote the first event reception time at the tag. Using the pre-measured event period $\tau$ of the RFID reader, the tag can estimate the next event reception time at time $t_1$ as
\begin{eqnarray}
\hat{C}_w(t_1) &=& C_w(t_0)+\tau(1+\hat{f}_w^r(t_0)), \label{eq:core}
\end{eqnarray}
where $\hat{f}_w^r(t_0)$ denotes the \emph{estimated relative frequency} at time $t_0$, which is an estimate for \eqref{eq:relative_frequency}. The intuition behind \eqref{eq:core} is that $1+\hat{f}_w^r(t_0)$ gives the estimated value of $f_r(t_0)/f_w(t_0)$, $\tau f_r(t_0)$ gives the estimated local time passed on the reader side and $\tau f_r(t_0)/f_w(t_0)$ gives the estimated number of clock ticks on the tag side that will pass until the next packet reception. Redefining \eqref{eq:gamma_error}, the \emph{prediction error} at time $t_1$ can be written as
\begin{eqnarray}
\gamma(t_1) & = & 
C_{w}(t_{1})-\hat{C}_{w}(t_{1}) = 
C_{w}(t_{1})-C_{w}(t_{0})-\tau(1+\hat{f}_{w}^{r}(t_0)). \label{eq:gamma_t_1}
\end{eqnarray}
Apparently, the objective is to minimize the prediction error so that the event reception time is estimated correctly. As can be observed from \eqref{eq:gamma_t_1}, the estimation error on the relative clock frequency at time $t_0$, i.e. $\hat{f}_w^r(t_0)$, depends on the mean event period $\tau$, and its variance (since they affect $C_w(t_1)-C_w(t_0)$). Inspired by the integral controller update rule in \cite[Section III, (5)]{pi2015}, we propose to update the estimated relative frequency at time $t_1$ as
\begin{eqnarray}
\hat{f}_{w}^{r}(t_1) &=& \hat{f}_{w}^{r}(t_0)+\beta\gamma(t_1), 
\label{eq:update_rule}
\end{eqnarray}
where $\beta$ is the \emph{integral gain}. Assume that the uncertainty of the event period is $\varepsilon\sim\mathcal{N}(0,\sigma^2_{\tau})$ and the instantaneous change on the relative clock frequency is $\frac{df_{w}^{r}(t)}{dt}\sim\mathcal{N}(0,\sigma_{\eta}^{2})$. Under these assumptions, it can be proven that if $0<\beta<\frac{2}{\tau}$ is satisfied the synchronization will hold eventually; i.e. $\hat{f}_{w}^{r}=f_{w}^{r}$. Moreover, the mean and the variance ($\sigma^2_\gamma$) of the steady-state (asymptotic) estimation error of WISP-Sync can be given by 
\begin{eqnarray}
\gamma & \sim & \mathcal{N} \left(0,\sigma^2_\gamma=\frac{\sigma^2_{\eta}\left(\tau^2-\beta\tau^3+\frac{\beta^2\tau^4}{3}\right)+\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)}+\frac{\sigma^2_{\eta}\tau^3}{3}+  \sigma^2_{\tau} \right),
\label{eq:sync_performance}
\end{eqnarray}
which is derived in Section~\ref{sec:app_b} of the Appendix.

\begin{figure*}
	\centering
	\includegraphics[scale=0.35]{figures/sim-theory.eps}
	\caption{\label{fig:theory-vs-simulations} The plot of the cumulative distribution function (CDF) of $\gamma$ in \eqref{eq:sync_performance} and the CDF of the steady-state $\gamma$ obtained during numerical simulations of \eqref{eq:update_rule} (by implementing \eqref{eq:theta} and \eqref{eq:delta_recursive} in Appendix). We used $\tau=1$, $\beta= 0.0001$, $\sigma_{\tau}=0.00005$ and our plots depict the theoretical and simulation results with respect to $\sigma_\eta = 0.0002$ and $\sigma_\eta = 0.0008.$ It can be observed that \eqref{eq:sync_performance} is consistent with the simulations and bigger uncertainty leads to bigger synchronization error.}
\end{figure*}
In Fig.~\ref{fig:theory-vs-simulations}, we emphasize that our theoretical derivation in \eqref{eq:sync_performance} is consistent with the simulations of \eqref{eq:update_rule}. We refer again the reader to the Appendix for the detailed proofs of the convergence and the asymptotic estimation error. Moreover, in Section~\ref{sec:adaptibility}, we discuss the impact of the parameters $\tau$ and $\beta$ on the synchronization accuracy.

\subsubsection{The Synchronization Algorithm}

\begin{algorithm}[t]
			\setstretch{1.35}
	\caption{\label{alg:sync} \emph{WISP-Sync}: An integral controller-based 
		algorithm at WISP tag side.}
	\begin{algorithmic}[1]	
		\scriptsize

		\Statex\textbf{Definitions:}
		\Statex$t_f$ \Comment{local time of the first \emph{BlockWrite}} 
		\Statex $C_w$ \Comment{bult-in clock of the WISP tag}
		\Statex $\hat{f}_w^r\leftarrow0$ \Comment{estimated relative clock 
			frequency}	
		\Statex $\tau\leftarrow$measured value of a particular RFID reader from 
		testbed 
		\Comment{mean \emph{event period}}	
		\Statex $\beta\leftarrow$ choose a value satisfying 
		$0<\beta<\frac{2}{\tau}$ \Comment{integral gain}	
		\Statex $V_{\min}, V_{\text{op}}, V_{\text{cur}}$ \Comment{minimum voltage threshold, operation voltage threshold ($V_{\text{op}}>V_{\min}$) and current voltage}

		\State $\square$ \textbf{Upon receiving the first \emph{BlockWrite}} 
		\State \quad $ t_f = C_w$ \Comment{store the local time in $t_f$}
		\State $\square$ \textbf{Upon receiving the last \emph{BlockWrite}} 
		\State \quad $\gamma = C_w- (t_f+\tau(1+\hat{f}_w^r))$ 
		\Comment{calculate estimation error $\gamma$} 
		\State \quad $\hat{f}_w^r = \hat{f}_w^r +\beta\gamma$ \Comment{apply 
			integral control 
			to update relative clock 
			frequency}
		\State $\square$ \textbf{Upon} $V_{\text{cur}}=V_{\min}$
		\State \quad store $\hat{f}_w^r$ into FRAM \Comment{store the relative 
			clock 
			frequency in non-volatile memory}
		\State $\square$ \textbf{Upon} $V_{\text{cur}} < V_{\text{op}}$
		\State \quad restore $\hat{f}_w^r$ from FRAM \Comment{restore the 
			relative 
			clock frequency}
	\end{algorithmic}	
\end{algorithm}

Based on this model we introduce Algorithm \ref{alg:sync}, namely \emph{WISP-Sync}, that synchronizes the speed of the software clock to the reference clock's speed using the update rule defined in \eqref{eq:update_rule}. The steps of this algorithm can be explained as follows. Initially, the tag lets its software clock run at the same speed of its local clock by setting its relative clock rate to zero. Upon receiving the first \emph{BlockWrite} event (Line 1), the tag stores its local clock value at the variable $t_f$ (Line 2). After receiving the last \emph{BlockWrite} event (Line 3), first it calculates the estimation error (Line 4). Then, it applies the correction on its estimated relative rate $\hat{f}_w^r$ by multiplying the estimation error with the integral gain $\beta$ and adding it to $\hat{f}_w^r$ (Line 5). After receiving successive events, $\hat{f}_w^r$ will converge to its desired value eventually, as proven in the Appendix.

\subsubsection{State Recovery}

Due to intermittent RF power, the tag is subject to frequent power losses, which will lead to loss of the synchronization state. In voltage-aware systems~\cite{Buettner:2011:Dewdrop,Ransford:2011:Mementos,Lucia:2015:Execution,Colin:2016:Debugger}, the general idea is to save the system state when the input voltage is below a pre-defined threshold $V_{\min}$ and then recover it back when the input voltage is above a pre-defined threshold $V_{\text{op}}$. As shown in Algorithm~\ref{alg:sync}, the only part of the system state pertaining to synchronization is the variable $\hat{f}_w^r$. Therefore, if this variable is also saved to FRAM (when there is a power loss and recovered when there is sufficient voltage) the synchronization will resume from the point of power loss, that will prevent loss of synchronization and lead to improved synchronization accuracy. These steps are summarized in Lines 6--9 of Algorithm~\ref{alg:sync}. It should be noted that these steps require hardware support so that the voltage level is monitored continuously and the corresponding voltage level indicator events are generated. Another approach can be keeping $\hat{f}_w^r$ in non-volatile memory as a \emph{persistent variable} so that the corresponding FRAM memory location of $\hat{f}_w^r$ is updated immediately when this variable is updated---eliminating the need for Lines 6--9 of Algorithm~\ref{alg:sync}, and in turn extra voltage measurement hardware, but introducing more FRAM access overhead.

\subsubsection{Computation Complexity\label{subsec:complexity}}

It can be observed that WISP-Sync employs only very simple arithmetic operations at each step, e.g. Lines 3--5 are composed of only two subtraction, two addition and two multiplication operations. Moreover, the \emph{state of synchronization}, i.e. the parameters that are required to be saved in the non-volatile memory, is represented by only the variable $\hat{f}_w^r$. As a consequence, WISP-Sync addresses \emph{Challenge I} in Section~\ref{sec:Motivation} in terms of computation and memory overhead.

\subsubsection{Adaptability} 
\label{sec:adaptibility}

\begin{figure*}
\centering
\includegraphics[scale=0.35]{figures/alpha-period-vs-variance.eps}
\caption{\label{fig:simulations} Numerical results showing the relations between communication period, integral gain and the steady-state variance. First, we set $\sigma_{\tau}=0.4097$, $\sigma_\eta = 0.0008$, $\beta=1$ and plotted $\sigma^2_{\gamma}$ using different $\tau$ (left hand side). Observe that increasing $\tau$ increases $\sigma^2_{\gamma}$. Then, we set $\sigma_{\tau}=0.4097$,	$\tau=1$ and $\beta \in (0,2)$ due to \eqref{eq:step_size_bounds-pi}. Using different values for $\sigma_\eta$, we plotted how $\sigma^2_{\gamma}$ changes with respect to the $\beta$ (right hand side). Observe that for each value of $\sigma_\eta$, there is a particular $\beta$ that minimizes $\sigma^2_{\gamma}$.}
\end{figure*}

In Section \ref{subsec:event_period_uncertainty} we presented our measurements pertaining to the event period and showed 
that $\sigma_{\tau}=0.4097$. Moreover, from Fig.~\ref{fig:frequency}, we calculated $\sigma_{\eta} \approxeq 0.0008$ using the \textsf{normfit} function of MATLAB.  Substituting these values in \eqref{eq:sync_performance}, we performed MATLAB simulations to calculate $\sigma^2_\gamma$, which are depicted in Fig.~\ref{fig:simulations}. These simulations led us to reveal the following points:

\begin{itemize}
\item Considering our previous measurements, it is apparent that $\sigma^2_{\tau}$ is the dominant factor effecting the synchronization performance. Therefore, it is desirable to keep transmission delay variations and also $\tau$, as small as possible. Apparently, the bigger the resynchronization period $\tau$, the worse the synchronization error is (see left hand side of Fig.~\ref{fig:simulations}).

\item Parameter $\beta$ has significant impact on the performance of the WISP-Sync. From 	\eqref{eq:sync_performance}, it can be concluded that the synchronization error can be minimized using a proper value of $\beta$. However, the $\beta$ value that minimizes 	\eqref{eq:sync_performance} depends on the parameters $\sigma_{\eta}$ and $\sigma_{\tau}$ which represent randomness of the system. In Fig.~\ref{fig:simulations} (right hand side), we fixed $\tau$ and $\sigma_{\tau}$, we used different values of $\sigma_{\eta}$ and we plotted $\sigma^2_{\gamma}$ with respect to the values of the $\beta$ that will lead to convergence. Observe that a different $\beta$ value minimizes $\sigma^2_{\gamma}$ for each different $\sigma_{\eta}$.

\item Due to varying RF power levels, $\sigma_{\eta}$ is also a time-varying value. Therefore, $\beta$ should be selected to compensate for also \emph{dynamic} $\sigma_{\eta}$ in a \emph{fast manner} to keep synchronization error within desired bounds---which addresses \emph{Challenge II} in Section~\ref{sec:Motivation}. However, the $\beta$ that minimizes $\sigma^2_{\gamma}$ may lead to longer convergence time (see Appendix). Therefore, $\beta$ also defines how \emph{reactive} WISP-Sync is against the dynamic environment, i.e. the \emph{adaptivity}.
\end{itemize}

\section{Evaluation of WISP-Sync}
\label{sec:eval-wispsync}

In order to evaluate WISP-Sync, we considered the value of $\gamma$ in Algorithm~\ref{alg:sync} (the estimation error) as an evaluation metric. We collected the local clock readings of the tag at the first and last \emph{BlockWrite} events using our testbed setup and sniffer, and used MATLAB to implement WISP-Sync algorithm which processes the collected clock values to output the software clock. This gave us the flexibility to try different approaches without reprogramming the tag.

\paragraph{Selection of the Integral Gain $\beta$}

\begin{figure}
\centering
\includegraphics[scale=0.35]{figures/integral_gain.eps}
\caption{\label{fig:beta} The synchronization error tends to get smaller with smaller integral gains, but after some point decreasing it has no significant effect due to the precision of 32\,kHz clock. Hence, we selected a fixed value of $\beta=0.0001$ for the next steps. }
\end{figure}

Since we measured the event period as 226.76\,ms on average (see Fig.~\ref{fig:block_write_delay}), we set $\tau=7086$ 
clock ticks since the local clock of the tag is operating at 32\,kHz and each clock tick occurs every 32 microseconds. First, we explored how $\beta$ affects the performance of the algorithm. By substituting $\tau$ into \eqref{eq:step_size_bounds-pi}, the convergence condition becomes $0<\beta<0.000284$ in our case. We present the synchronization error with different $\beta$ values in Fig.~\ref{fig:beta}. Observe that the synchronization error tends to get smaller with smaller $\beta$ values but the convergence time increases negligibly. Moreover, due to the low-precision 32\,kHz clock, decreasing $\beta$ has no significant effect after some point. We mention that  in order to make WISP-Sync adapt faster, WISP-Sync can start with a big value of $\beta$ to converge fast and then decrease $\beta$ gradually to decrease the synchronization error, i.e. the adaptive strategy proposed in~\cite{pi2015}. However, from  Fig.~\ref{fig:beta}, we consider a fixed value of $\beta=0.0001$ for the next evaluation steps.

\paragraph{Synchronization Performance}

\begin{figure}
\centering
\includegraphics[scale=0.35]{figures/synchronization_accuracy.eps}
\caption{\label{fig:sync_error}Event-based synchronization performance when the tag is powered through a constant voltage source and through RF power. Mean synchronization errors without and with WISP-Sync were almost 127 and 16 clock ticks under stable voltage, and 175 and 22 clock ticks with RF power harvesting (represented by the straight lines in this figure), respectively.}
\end{figure}

Fig.~\ref{fig:sync_error} presents the synchronization error measurements when the tag is powered by using a constant voltage source, and powered only by an RFID reader. Measurements under constant and stable voltage allowed us to observe synchronization under stable clock frequency. We obtained a significant synchronization performance with WISP-Sync---almost a factor of eight less synchronization error as compared to the case where we did not perform any synchronization. It should be noted that even in this case, we observed quite fluctuating synchronization errors due to the varying transmission delays, that led a peak error to appear between samples 50 and 60. Apart from experiments with constant voltage input, measurements under highly varying RF power led us to observe the behavior of synchronization under highly varying clock frequencies. In this case, there is a considerable amount of increase on the error, i.e. approximately twice as much, as compared to the stable voltage case. However, we also obtained considerable improvements with our approach with unstable clock frequencies. WISP-Sync was quite reactive to the frequently varying clock frequencies and provided almost a factor of eight better accuracy.

\paragraph{Power Loss and Recovery}

\begin{figure}
\centering
\includegraphics[scale=0.35]{figures/power_loss.eps}
\caption{\label{fig:power_loss}Impact of the state recovery on the synchronization performance. During the power loss periods, the tag is assumed to be turned off. With state recovery, presented by the solid line, the relative clock frequency is assumed to be saved in the FRAM and recovered back when sufficient power is supplied. On the other hand, the dashed line represents the case of starting synchronization from scratch. It can be observed that with state recovery, the tag adapts itself quickly and synchronization performance is considerably better, i.e. more than 100 clock ticks less synchronization error on power loss.} 
\end{figure}

In order to evaluate the performance of WISP-Sync under power losses, we simulated power loss in MATLAB by omitting the collected clock readings during predefined time intervals. As presented in Fig.~\ref{fig:power_loss}, after samples 10 and 30 WISP-Sync is disabled and before the beginning of the sample times 20 and 40 the algorithm is enabled. The solid line in this figure indicates enabling algorithm with state recovery, where the algorithm starts to execute using the value of $\hat{f}_w^r$  previously saved in the FRAM before power loss. It is apparent that with state recovery, the tag quickly adapts itself and re-synchronization is quickly established. On the other hand, starting the algorithm from scratch, as presented by the dashed lines, is problematic since it takes several steps to re-synchronize the tag.
%
%\begin{figure}
%		\begin{lstlisting}[style = mystyle,escapechar=|] 
%		_attribute_((section(".infoA"))) float f_w = 0.0f;
%		uint16_t t_f = 0;
%		
%		void wisp_sync(uint16_t c_w) {
%			float error = (float)c_w - ((float)t_f + TAU * (1.0f + f_w));
%			f_w += BETA * error;
%		}
%		\end{lstlisting}
%\caption{The actual implementation of  lines 3--5 of Algorithm~\ref{alg:sync} in C language for TI MSP430FR5969 platform. Constants \texttt{BETA} and \texttt{TAU} are defined outside of this function in the remaining part of the C program implementing Algorithm~\ref{alg:sync}.}
%		\label{fig:wisp_implementation}
%\end{figure}
%
\paragraph{Energy Overhead}
%
We now present the real energy measurements of the actual WISP tag-side C language implementation of lines 3--5 of Algorithm~\ref{alg:sync} which are the main computation steps to calculate the new values of the variables pertaining to the synchronization. We performed our measurements using the Energy-interference-free Debugger (EDB)~\cite{colin2016energy} to measure the output voltage without affecting the energy state of the tag. EDB charges the capacitor of a device to a certain level and then disconnects itself. Using this feature of EDB we charge the capacitor of WISP to 2.45\,V, i.e. just above the operational level of WISP. While the program is being executed, the value of the capacitor is measured at specific code locations---in this case before Line 3 and after Line 5 of Algorithm~\ref{alg:sync}. The difference between these two results represents the energy overhead of WISP-Sync. We repeated this experiment ten times to eliminate quantization effects during measurements. We observed an average energy of $\approx1.85 \mu$J with a standard deviation of 0.038\,$\mu$J is spent by the tag in order to perform the aforementioned computations.

The main source of energy consumption in our implementation is the \emph{floating point arithmetic}: WISP-Sync requires two multiplications, three additions and one subtraction on float variables. Since MSP430 microcontroller does not have any floating point coprocessor, the float operations are handled using a software library, that makes them relatively slow. We used the standard gcc compiler (specifically, msp430-gcc version 6.2.1.16) math library for the sake of the portability of float operations. However, we expect that a more efficient architecture-specific implementation of the math library in software or an architecture with a math coprocessor will lead to less computation time and in turn relatively less energy overhead for WISP-Sync~\cite{mathlib_website}. Another point we would like to mention is that non-volatile memory access consumes more power as compared to SRAM access~\cite[Section 3.3]{ti_fram}. The only non-volatile variable in WISP-Sync is $\hat{f}_w^r$ in Algorithm~\ref{alg:sync}. This variable is read twice but modified once in our implementation. On the other hand, WISP-Sync maintains only one variable in SRAM, i.e. $t_f$, which is read only once. We conclude that WISP-Sync performs three FRAM and one SRAM access operations---low energy overhead in terms of memory access.

\paragraph{Limitations of WISP-Sync}

In conclusion, the experimental results so far indicated that a maximum synchronization error of approximately 1.5 ms can be ensured between an RFID reader and a single WISP tag when employing event-based synchronization. The main factors that limited the WISP-Sync synchronization accuracy were the clock precision of the WISP platform and the non-deterministic transmission delays during reader-tag communication. As we indicated, we used the external 32\,kHz clock of the platform since it is also running when the tag is in very low-power operation mode. Therefore, in the best case when there are no transmission delays, the quantization effect due to the timing measurements using the low precision clock will not allow a better synchronization accuracy than 32\,$\mu$s. On the other hand, to obtain this synchronization accuracy with WISP-Sync, the reader and the tags shoud support a high-precision and accurate packet timestamping as in sensor networks~\cite{Maroti2004}. As indicated in Section \ref{sec:CRFIDs}, the effect of the non-deterministic delays should be reduced for a better synchronization accuracy.

Even though the event-based approach is relatively lightweight and adaptive as compared to the sender-receiver synchronization, it has some limitations. First, the tag is unable to obtain an \emph{explicit reference clock value}. The only variable it tunes is the $\hat{f}_w^r$ that represents the relative frequency of the software clock with respect to the clock of the reader. Therefore, we require additional steps that will allow tags to obtain the actual reference clock value. However, to the best of our knowledge, there is no such command in the EPC Gen 2 standard that allows to send the explicit RFID reader time to the tag. Second, WISP-Sync requires knowledge about $\tau$, which can be RFID reader dependent and requires a sniffer setup described in Section~\ref{sec:sniffer} to measure. This limitation can be addressed by employing an iterative estimation as an initial step to the WISP-Sync: the tag can start with an initial estimated $\tau$ value, and improve its estimation as it receives successive packets from the reader. After some iterations, the tag can have a rough estimate $\hat{\tau}$ that can be used on behalf of $\tau$. However, estimation errors might also perturb the synchronization accuracy. Last, the steps in Fig.~\ref{fig:event_based} present the communication scenario between one WISP and the reader. To allow other tags sniff this communication and synchronize, a broadcast primitive is crucial---in the next section we show how to overcome this limitation.

\section{The Synchronization of Multiple WISP Tags Using the WISP-Sync}
\label{sec:multi-wisp}

In this section, we turn our attention to the synchronization of  multiple WISP tags using the WISP-Sync. To this end, we have done the \emph{real implementation} of the WISP-Sync using the C programming language and TI's CCS~\cite{ti_css}, instead of using MATLAB to simulate our protocol using the collected the clock values from our testbed. Moreover, we extended our experimental testbed in Section~\ref{sec:testbed} by introducing two additional tags, as presented in Fig.~\ref{fig:testbed-new}.

\begin{figure}
	\centering
	\begin{tikzpicture}
	\node[anchor=south west, inner sep=0] at 
	(0,0){\includegraphics[height=.3\columnwidth]{figures/3-wisps_resized.jpg}};

	\node [fill= white,circle] at (0.8,2.5) { };
	\node [color=black ] at (0.8,2.5) { 1 };

	\node [fill= white,circle] at (1,4.5) {  };
	\node [color=black ] at (1,4.5) { 2 };

	\node [fill= white,circle] at (2.3,2.1) {  };
	\node [color=black ] at (2.3,2.1) { 3 };

	\node [fill= white,circle] at (5.5,1) {  };
	\node [color=black ] at (5.5,1) { 4 };

	\node [fill= white,circle] at (5.5,2.8) {  };
	\node [color=black ] at (5.5,2.8) { 5 };

	\end{tikzpicture}
\caption{\label{fig:testbed-new} A testbed of multiple tags: (1,2,3) tags, (4) logic analyzer, (5) breadboard. Tags 1 and 3 were at 8\,cm while tag 2 was at 20\,cm distance to the reader. The breadboard was used to make connections between the logic analyzer and the tags.} 
\end{figure}

We programmed the RFID Reader to send synchronization points via \emph{BlockWrite} command with a period of one second and we programmed two tags using our WISP-Sync implementation. On the other hand, we disabled clock synchronization for the last tag---which was a baseline for our evaluation.  In order to allow tags to simultaneously receive the information sent by the RFID Reader, we exploited an overhearing mechanism presented first in ~\cite[Section III-B]{stork:2017}. Therefore, during the experiments in our testbed (i) all of the tags received the periodic synchronization points sent by the RFID Reader; (ii) only two tags process this information and run Algorithm~\ref{alg:sync} to update their software clocks; (iii) the last tag did not participate in synchronization.

In order to measure the synchronization accuracies of the tags, we configured one of the ports of the MSP430 microcontroller as an output port and let tags (i) to set this port when they receive the first \emph{BlockWrite} command in Algorithm~\ref{alg:sync} (Line 1); (ii) to set a timer that will fire at their predicted time of the reception of the last \emph{BlockWrite} command from the RFID Reader; (iii) to clear this port when their timer fires. We connected the output ports to logic analyzer~\cite{saleae} and measured the time difference between the times at which the output ports are cleared. This allowed us to measure the prediction errors among the tags. We observed that all the tags set their output ports when they overhear the first \emph{BlockWrite} command, however their clear times are quite different and depend on their prediction error---the port clear times of the synchronizing two tags were closer to each other while they were quite different than the unsynchronized tag.

\begin{figure}
\centering
\includegraphics[scale=0.35]{figures/multiwisp-experiments.eps}
\caption{\label{fig:multi-sync-results}The synchronization error calculated during the experiments in our testbed. The blue dotted line denotes the maximum synchronization error among all of the tags (including the one which did not participate in synchronization) whereas the red line denotes the maximum error just between the synchronized tags.} 
\end{figure}

In Fig.~\ref{fig:multi-sync-results}, we plotted two different results: (i) the blue line depicts the maximum synchronization error (i.e. the maximum pairwise difference between the port clear times) when we considered all of the tags; (ii) the red line depicts when only synchronized tags are considered. From this figure it can be concluded that synchronization led to a factor of approximately four times better prediction on average among the tags: we observed an average synchronization error of 1\,ms while it was 4\,ms when one of the tags was unsynchronized.

\section{Conclusions and Future Research Directions}
\label{sec:Conclusion}

In this article we explored synchronization scenarios between an RFID reader and WISP tags. We studied sender-receiver and event-based synchronization mechanisms in this setting and provided initial designs that will guide future explicit synchronization mechanisms among individual WISPs that reside inside the communication range of a common RFID reader. We provided implementation and evaluation of these designs in our testbed and identified their limitations and drawbacks. Our main finding is that with lightweight mechanisms, as of now, an average synchronization error of approximately 1\,ms can be ensured among tags. We provide the following issues for future studies in this domain:

\begin{itemize}
\item \textbf{Network-wide synchronization:} We studied event-based synchronization among many tags. However, synchronization of these tags where they can access the explicit clock of the RFID reader and also synchronization of the whole CRFID network composed of several RFID readers and tags is still an issue, due to the single-hop nature of backscatter communication. Since tags can only communicate with the reader, not with their neighboring tags, it is interesting to explore \emph{WISP to WISP} synchronization.

\item \textbf{Power loss and recovery:} It is crucial to save the synchronization status to non-volatile memory just before the power loss, but writing to non-volatile memory also consumes energy. Hence, when to save the synchronization status without extra hardware support is worth exploring~\cite{Buettner:2011:Dewdrop,Ransford:2011:Mementos,Lucia:2015:Execution,Colin:2016:Debugger}.

\item \textbf{Voltage-frequency relations:} It might be interesting to explore the voltage-clock frequency relationship since1 tags are subject to varying voltage source when they are powered with only RF energy harvesting from the RFID reader. We anticipate that voltage level should be incorporated to the establishment of the software clock so that voltage dependent instability of clock frequency can be compensated. However, this is a ``chicken or egg'' problem since reading the voltage level also consumes energy.

\item \textbf{Synchronization in other IPD platforms:} Synchronization in other IPD platforms, e.g. that use ambient backscatter communication \cite{ambient_backscatter}, is worth exploring as well. 
\end{itemize}

% Bibliography
\bibliographystyle{IEEEtranTCOM}
\bibliography{IEEEabrv,references}

\pagebreak
%\appendix[Analysis of Algorithm~\ref{alg:sync}] 
\appendix

We provide the synchronization conditions of Algorithm~\ref{alg:sync} and show that the synchronization will be established eventually. We need to mention that we follow similar calculation steps to the ones presented in the studies~\cite{pi2015,Yildirim:Gradient:2016}. However, since we use a different clock drift model in this article, i.e. Brownian motion rather than uniform distribution, the following  mathematical analysis is moderately different and more realistic as compared to those in the previous studies.

Since $f_w$ and $f_r$ are time varying, by following 
\cite{Zhong:2011:On-demand,Huang:2013:PSR}, we model the 
\emph{evolution} 
of $f_w^r$ as 
integrated white noise, i.e. \emph{Brownian motion}, given as
\begin{eqnarray}
f_{w}^{r}(t_{0}+\delta) & \triangleq & 
f_{w}^{r}(t_{0})+\int_{t_{0}}^{t_{0}+\delta}\eta(t)dt.\label{eq:drift}
\end{eqnarray}
Here, $\delta$ denotes the amount of time after $t_0$ and $\eta(t)$ denotes 
the instantaneous change on the relative clock frequency, namely 
\emph{relative drift rate} at time $t$. We assume that 
$\eta(t)\sim\mathcal{N}(0,\sigma_{\eta}^{2})$ and for all 
time instants $t'$ and $t''$, $\eta(t')$ and $\eta(t'')$ are uncorrelated 
random variables. In order to ease the calculations in the following sections, 
we will consider the \emph{average relative clock frequency} in an interval of 
interest $[t_0,t_0+\delta]$ rather than instantaneous relative frequency 
in~\eqref{eq:drift}. We denote this value by $\bar{f}_{r}^{w}(t_0,t_0+\delta)$ 
and define it as 
\begin{eqnarray}
\bar{f}_{w}^{r}(t_0,t_0+\delta) & \triangleq & 
\frac{1}{\delta}\int_{t_0}^{t_0+\delta} 
f_w^r(t)dt \triangleq
f_{w}^{r}(t_{0})+\frac{1}{\delta}\int_{0}^{\delta}\int_{t_{0}}^{t_{0}+t}\eta(u)dudt.
\label{eq:drift_relation}
\end{eqnarray}

\subsection{Proof of Convergence}
Consider \eqref{eq:gamma_t_1} and without loss of 
generality we assume that 
$C_{w}(t_{1})-C_{w}(t_{0})=\tau(1+\bar{f}_{w}^{r}(t_0,t_1))+\varepsilon$
where $\tau$ denotes the event period and $\varepsilon$ accounts for the 
uncertainty of the event period (as 
indicated in Section \ref{subsec:event_period_uncertainty}). For ease of 
calculations we assume that $\varepsilon\sim\mathcal{N}(0,\sigma^2_{\tau})$. 
Define \emph{relative frequency estimation error} as
\begin{eqnarray}
\tilde{f}_{w}^{r}(t_0) & \triangleq & 
f_{w}^{r}(t_0)-\hat{f}_{w}^{r}(t_0).\label{eq:relative_drift_estimation_error}
\end{eqnarray}
Using \eqref{eq:gamma_t_1}, \eqref{eq:drift_relation} and 
\eqref{eq:relative_drift_estimation_error} we get:
\begin{eqnarray}
\gamma(t_1) & = & 
\tau(\bar{f}_{w}^{r}(t_0,t_1)-\hat{f}_{w}^{r}(t_0))+ \varepsilon = 
\tau\tilde{f}_{w}^{r}(t_0)+\frac{\tau}{t_1-t_0}\int_{0}^{t_1-t_0}\int_{t_{0}}^{t_0+t}\eta(u)dudt
+ \varepsilon.
\end{eqnarray}
For ease of representation, let us introduce the following 
notation:
\begin{eqnarray}
\Phi(h) &\triangleq& \int_{t_h}^{t_{h+1}}\eta(t)dt, \\
\Omega(h) &\triangleq& 
\frac{\tau}{t_{h+1}-t_h}\int_{0}^{t_{h+1}-t_h}\int_{t_{h}}^{t_h+t}\eta(u)dudt,
\\
\Delta(h) &\triangleq& \tilde{f}_{w}^{r}(t_{h}). \label{eq:delta}
\end{eqnarray}
Using this notation, we denote $\gamma(t_{h+1})$ by $\Theta(h+1)$ as
\begin{eqnarray}
\Theta(h+1) &\triangleq& \gamma(t_{h+1})=\tau\Delta(h) + \Omega(h)+\varepsilon. 
\label{eq:theta}
\end{eqnarray}
Moreover, since $\Delta(h)$, $\Omega(h)$ and $\varepsilon$ are 
independent random variables, we can write $\Delta(h+1)$ as
\begin{eqnarray}
\Delta(h+1) &\overset{\eqref{eq:delta}}{=}& 
\Delta(h)+\tilde{f}_{w}^{r}(t_{h+1})-\tilde{f}_{w}^{r}(t_{h}) \nonumber \\
&\overset{\eqref{eq:relative_drift_estimation_error}}{=}& \Delta(h)
+f_w^r(t_{h+1})-f_w^r(t_{h}) -\hat{f}_w^r(t_{h+1})+\hat{f}_w^r(t_{h})
\nonumber\\
&\overset{\eqref{eq:update_rule},\eqref{eq:drift}}{=}& 
\Delta(h)+\int_{t_h}^{t_{h+1}}\eta(t)dt-\beta\Theta(h+1) \nonumber \\
&\overset{\eqref{eq:theta}}{=}& 
(1-\beta\tau)\Delta(h)-\beta(\Omega(h)+\varepsilon) +\Phi(h). 
\label{eq:delta_recursive}
\end{eqnarray} 
Considering \eqref{eq:theta} and \eqref{eq:delta_recursive}, we can write the 
system evolution as
\begin{eqnarray}
\underset{X(h+1)}{
	\underbrace{
		\left[\begin{array}{c}
		\Theta(h+1)\\
		\Delta(h+1)
		\end{array}\right]
	}
} & = & 
\underset{A}{
	\underbrace{
		\left[\begin{array}{cc}
		0 & \tau\\
		0 & 1-\beta\tau
		\end{array}\right]
	}
}
\underset{X(h)}{
	\underbrace{
		\left[\begin{array}{c}
		\Theta(h)\\
		\Delta(h)
		\end{array}\right]}
}
+
\underset{Y(h)}{
	\underbrace{
		\left[\begin{array}{c}
		\Omega(h)+\varepsilon\\
		\Phi(h)-\beta(\Omega(h)+\varepsilon)
		\end{array}\right]}
}.\label{eq:state_space-pi}
\end{eqnarray}
Taking the expectation of both sides of \eqref{eq:state_space-pi} yields 
$\expected{X(h+1)} = A\expected{X(h)}$ since 
the means of the random variables in the matrix $Y(h)$ are all zero. According 
to stability of linear systems \cite{book-luenberger}, \emph{asymptotic 
	convergence} will be reached if and only if the magnitudes of the 
eigenvalues of the matrix $A$ are strictly smaller than one. The eigenvalues 
$\lambda_1$ and $\lambda_2$ of the matrix $A$ can be obtained 
by solving
\begin{eqnarray}
\left|\begin{array}{cc}
-\lambda & \tau\\
0 & 1-\beta\tau-\lambda
\end{array}\right|=0 &\implies& 
\begin{array}{c}
\lambda_1 = 0,\\
\lambda_2 = 1-\beta\tau.  
\end{array}\label{eq:eigens}
\end{eqnarray}
Therefore, the synchronization will be  established if 
and only if $|\lambda_1,\lambda_2|<1$, and in turn
\begin{align}
0 & <\beta<\frac{2}{\tau}\label{eq:step_size_bounds-pi}
\end{align}
should hold. The 
rate of convergence is governed by 
the largest eigenvalue $\lambda_2$, which means that bigger values of $\beta$ 
will lead to faster convergence. 
Consequently, selecting $\beta$ within the bound above will lead the system  
converge to the state  
\begin{eqnarray}
\underset{h\rightarrow\infty}{\lim}\expected{\Theta(h+1)} &=&
\expected{\Theta(h)}, \label{eq:asymp_theta} \\
\underset{h\rightarrow\infty}{\lim}\expected{\Delta(h+1)}&=&
\expected{\Delta(h)} \label{eq:asymp_delta}.
\end{eqnarray}
Hence, due to asymptotic convergence
\begin{eqnarray}
\expected{\Delta(\infty)} \overset{\eqref{eq:delta_recursive}}{=} 
\left(1-\beta 
\tau\right)\expected{\Delta(\infty)} &\implies& \expected{\Delta(\infty)}=0, 
\label{eq:delta_convergence}\\
\expected{\Theta(\infty)}  \overset{\eqref{eq:theta}}{=} 
\tau\expected{\Delta(\infty)}&\implies&\expected{\Theta(\infty)}=0. 
\label{eq:expected_error}
\end{eqnarray}
Consequently, the prediction error $\Theta(h)=\gamma(t_h)$ 
eventually converges to zero as $h$ goes to infinity. This means the WISP tag 
will estimate the occurrence times of the successive
events with zero error in expectation.

\subsection{Steady-State (Asymptotic) Synchronization Error}
\label{sec:app_b}
An analytical expression for the \emph{asyptotic 
	synchronization error} can be obtained by calculating 
$\underset{h\rightarrow \infty}{\lim}\mathrm{Var}(\Theta(h))$. This 
calculation reduces to find 
$\mathrm{Var}(\Theta(\infty))=\expected{\Theta(\infty)^2}$, since it holds 
from 
\eqref{eq:expected_error} that $\expected{\Theta(\infty)}=0$. Using 
\eqref{eq:theta}, we 
get
\begin{eqnarray}
\expected{\Theta(\infty)^2} &\overset{\eqref{eq:theta}}{=}& 
\tau^2\expected{\Delta(\infty)^2}  
+\expected{(\Omega(h)+\varepsilon)^2},\label{eq:expected_theta_infinity}
\end{eqnarray}
since $\expected{\Delta(\infty)}=0$ from \eqref{eq:delta_convergence} and 
$\Delta(h)$, $\Omega(h)$ and $\varepsilon$ are independent. 
In order to calculate \eqref{eq:expected_theta_infinity}, let us derive first 
$\expected{\Delta(\infty)^2}$ as
\begin{eqnarray}
\expected{\Delta(\infty)^2} 
&\overset{\eqref{eq:delta_recursive},\eqref{eq:asymp_delta}}{=}& 
(1-\beta\tau)^2	
\expected{\Delta(\infty)^2}  
+\expected{(\Phi(h)-\beta(\Omega(h)+\varepsilon)^2} 
\label{eq:expected_delta_infinity}.
\end{eqnarray}
In order to calculate \eqref{eq:expected_delta_infinity}, the first step is to 
obtain the following expectation:
\begin{eqnarray}
\expected{(\Phi(h)\!-\!\beta(\Omega(h)\!+\!\varepsilon)^2} 
&\!\!\!\!=\!\!\!\!&
\expected{(\Phi(h)^2}
\!-\!2\beta\expected{\Phi(h)\Omega(h)} 
\!+\!\beta^2\expected{\Omega(h)^2}\!+\!\beta^2\expected{\varepsilon^2}. 
\label{eq:stochastic_expectation}
\end{eqnarray}
The equality \eqref{eq:stochastic_expectation} holds due to the fact that 
$\expected{\Phi(h)\varepsilon}=\expected{\Phi(h)}\expected{\varepsilon}=0$ and 
$\expected{\varepsilon\Omega(h)}=\expected{\varepsilon}\expected{\Omega(h)}=0$ 
(since 
$\expected{\varepsilon}=\expected{\Phi(h)}=\expected{\Omega(h)}=0$ and 
$\varepsilon$ is independent from $\Phi(h)$ and $\Omega(h)$). The
following \emph{stochastic integrals} are required for 
\eqref{eq:stochastic_expectation}:
\begin{eqnarray}
\expected{\Phi(h)^2} 
& = & 
\int_{0}^{t_{h+1}-t_h}\sigma_{\eta}^{2}dt =  \sigma_{\eta}^{2}(t_{h+1}-t_h),\\
\expected{\Omega(h)^2} 
& = & 
\frac{\tau^2}{(t_{h+1}-t_h)^2}
\sigma_{\eta}^{2}\int_{0}^{t_{h+1}-t_h}t^2dt = 
\sigma_{\eta}^{2}\frac{\tau^2(t_{h+1}-t_h)}{3},\label{eq:stochastic_integral}\\
\expected{\Phi(h)\Omega(h)} 
& = &
\frac{\tau}{t_{h+1}-t_h}\sigma_{\eta}^{2}\int_{0}^{t_{h+1}-t_h}tdt = 
\sigma_{\eta}^{2}\frac{\tau(t_{h+1}-t_h)}{2}, \\
\expected{\varepsilon^2} 
& = & 
\sigma_{\tau}^{2}.
\end{eqnarray}	
The derivation of the expectations and stochastic integrals above relies on the 
assumptions we mentioned 
at the beginning of this Appendix: (i) $\eta(t')$ and $\eta(t'')$ are 
uncorrelated for all $t'\neq t''$, therefore $\expected{\eta(t')\eta(t'')}=0$; 
(ii) since $\expected{\eta(t')}=0$, therefore 
$\expected{(\eta(t'))^2}=\mathrm{Var}(\eta(t'))=\sigma_{\eta}^2$. Putting these 
expected 
values into \eqref{eq:stochastic_expectation}, then in 
turn into \eqref{eq:expected_delta_infinity} lead:
\begin{eqnarray}
\expected{\Delta(\infty)^2}
& = &
\frac{\sigma^2_{\eta}(t_{h+1}-t_h)\left(1-\beta\tau+\frac{\beta^2\tau^2}{3}\right)+\beta^2\sigma^2_{\tau}}{(2\beta\tau-\beta^2\tau^2)}.
\label{eq:delta_infinity}
\end{eqnarray}
Finally, putting \eqref{eq:delta_infinity} into 
\eqref{eq:expected_theta_infinity} 
leads to
\begin{eqnarray}
\expected{\Theta(\infty)^2} & = & 
\frac{\sigma^2_{\eta}(t_{h+1}-t_h)\left(\tau^2
	-\beta\tau^3-\frac{\beta^2\tau^4}{3}\right)
	+\beta^2\sigma^2_{\tau}\tau^2}{(2\beta\tau-\beta^2\tau^2)} + 
\sigma^2_{\eta}\frac{\tau^2(t_{h+1}-t_h)}{3}+\sigma^2_{\tau}.
\end{eqnarray}
Without loss of generality, assume that $t_{h+1}-t_h\approxeq\tau$. Therefore, 
the asymptotic estimation error of Algorithm \ref{alg:sync} is given by
\begin{eqnarray}
Var\left(\underset{h\rightarrow\infty}{\gamma(t_h)}\right)& = & 
\expected{\Theta(\infty)^2} = 
\frac{\sigma^2_{\eta}\left(\tau^2-\beta\tau^3
	+\frac{\beta^2\tau^4}{3}\right)
	+\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)}+\frac{\sigma^2_{\eta}\tau^3}{3}
+  \sigma^2_{\tau}
\end{eqnarray}
which results in \eqref{eq:sync_performance}.

\end{document}

% End of v2-acmsmall-sample.tex (March 2012) - Gerry Murray, ACM

### 2017-07-26 matlab/frequency_testbed.m

@@ -1,0 +1,150 @@
+function frequency_testbed
+    clear all;
+    close all;
+
+    % Open the comma seperated data file - [Timestamp Port]
+    fileID = fopen('testbed data/w1_DC.csv');
+    file = textscan(fileID,'%f %d', 'Delimiter', ',');
+    fclose(fileID);
+
+    wisp1_DC_t  = cell2mat(file(1,1));
+    
+    fileID = fopen('testbed data/w2_DC.csv');
+    file = textscan(fileID,'%f %d', 'Delimiter', ',');
+    fclose(fileID);
+    
+    wisp2_DC_t  = cell2mat(file(1,1));
+
+    fileID = fopen('testbed data/w1_RF.csv');
+    file = textscan(fileID,'%f %d', 'Delimiter', ',');
+    fclose(fileID);
+
+    wisp1_RF_t  = cell2mat(file(1,1));
+    
+    fileID = fopen('testbed data/w2_RF.csv');
+    file = textscan(fileID,'%f %d', 'Delimiter', ',');
+    fclose(fileID);
+
+    wisp2_RF_t  = cell2mat(file(1,1));
+    %------------------------------------------------------
+
+    drift_dc = get_drift(wisp1_DC_t);
+    drift_rf = get_drift(wisp1_RF_t);
+    
+    freq_dc = get_freq(wisp1_DC_t);
+    freq_rf = get_freq(wisp1_RF_t);
+    
+    freq_dc_2 = get_freq(wisp2_DC_t);
+    freq_rf_2 = get_freq(wisp2_RF_t);
+    
+    %------------------------------------------------------
+    hFig = figure('DefaultAxesFontSize',18);
+    subplot(1,2,2);
+    plot(1:size(drift_dc),drift_dc,'DisplayName','DC Voltage','LineWidth',1.5);
+    %hold on
+    %plot(1:size(drift_rf),drift_rf,'DisplayName','RF Power','LineWidth',1.5);
+    xlabel('Time (second)')
+    ylabel('Drift (microsec.)')
+    xlim([1 100])   
+    %set(gca,'YTick',[1:10]); 
+    grid on
+    %legend('show')
+    %set(hFig, 'Position', [0 0 600 300])
+    
+    subplot(1,2,1);
+    %hFig = figure('DefaultAxesFontSize',18);
+    plot(1:size(freq_dc),freq_dc,'DisplayName','DC Voltage','LineWidth',1.5);
[…]

### 2017-07-26 matlab/least_squares.m

@@ -1,0 +1,31 @@
+clear all;
+close all;
+
+[X Y] = textread('testbed data/least_squares.txt', '%d %d', 'delimiter', ' ')
+
+mean_delay = 59; % the mean transmission delay we measured
+
+Error = [];
+m = 9;
+while (m+8)<numel(X)
+    
+    A = X(m:m+7);
+    B = Y(m:m+7);
+    p = polyfit(A,B,1);
+    error = Y(m+8)-polyval(p,X(m+8))
+    m = m+8;
+    Error = [Error error];
+end
+Error(Error > 10) = [];
+Error(Error < -10) = [];
+
+hFig = figure(1)
+
+histogram(Error);
+
+xlabel('Synchronization error (clock ticks)')
+ylabel('Num. of occurence')
+set(gca,'FontSize',18)
+ylim([0 90]);
+set(hFig, 'Position', [0 0 600 300])
+grid on

### 2017-07-26 matlab/multiwisp.m

@@ -1,0 +1,128 @@
+function simulations
+    clear all;
+    close all;
+    
+    [x_all y_all x_2 y_2]= calculate_sync_error();
+    
+    hFig = figure('DefaultAxesFontSize',18);
+    plot(x_all,y_all,':','DisplayName','Max. \gamma (3 Wisps)','LineWidth',2);
+    hold on 
+    plot(x_2,y_2,'DisplayName','Max. \gamma (2 Wisps)','LineWidth',1.5);
+    hold on 
+    plot(x_all,repmat(mean(y_all),numel(x_all),1),'--','DisplayName','Avg. \gamma (3 Wisps)','LineWidth',1);
+    hold on
+    plot(x_2,repmat(mean(y_2),numel(x_2),1),'-.','DisplayName','Avg. \gamma (2 Wisps)','LineWidth',1);
+    xlabel('Time (second)')
+    ylabel('\gamma (ms)')
+    %set(gca,'YTick',[1:10]);
+    grid on
+    legend('show')
+    set(hFig, 'Position', [0 0 600 300])
+    ylim([0 10])
+    xlim([4 max(x_all)])
+    
+    print(hFig,'multiwisp-experiments','-depsc2')
+end
+
+function [x1 y1 x2 y2] = calculate_sync_error()
+    % Open the comma seperated data file - [Timestamp Ch1 Ch2 Ch3]
+    fileID = fopen('testbed data/3w.csv');
+    file = textscan(fileID,'%f %d %d %d', 'Delimiter', ',');
+    fclose(fileID);
+
+    timestamps = 1000.0*cell2mat(file(1,1)); % in ms
+    wisp1 = cell2mat(file(1,2));
+    wisp2 = cell2mat(file(1,3));
+    wisp3 = cell2mat(file(1,4));
+
+    % reset on and off times
+    starts = zeros(3,1);
+    ends =  zeros(3,1);
+
+    x1 = [];
+    y1 = [];
+    
+    x2 = [];
+    y2 = [];
+
+    for i = 1:size(timestamps)
+        
+        
+
+        % save on and off times of the ports
+        if(starts(1) == 0 && wisp1(i) == 1)
+            starts(1) = timestamps(i);
+        elseif (starts(1) > 0 && ends(1) == 0 && wisp1(i) == 0)
+            ends(1) = timestamps(i);
+        end
+
+        if(starts(2) == 0 && wisp2(i) == 1)
+            starts(2) = timestamps(i);
[…]

### 2017-07-26 matlab/sender_receiver_uncertainty.m

@@ -1,0 +1,31 @@
+clear all;
+close all;
+fileID = fopen('testbed data/sender_receiver_uncertainty.txt','r');
+[A] = fscanf(fileID, ['%d']);
+A = A/10000; % from 10MHZ sample rate to milliseconds
+
+hFig = figure(1);
+
+h = histogram(A,'Normalization','probability');
+h.NumBins = 50;
+xlim([1.80 2.4]);
+
+hold on 
+A(A > 2.1) = [];
+A(A < 1.8) = [];
+
+[muhat,sigmahat,muci,sigmaci] = normfit(A,0.01)
+pd = fitdist(A,'Normal');
+x_values = 1.4:0.001:2.4;
+y = pdf(pd,x_values);
+y = y ./h.NumBins;
+plot(x_values,y,'LineWidth',2)
+
+ylim([0 0.5]);
+
+xlabel('Delay (ms)')
+ylabel('Normalized num. of occurence')
+legend('Frequency of the transmission delay', 'Normal fit function')
+set(gca,'FontSize',18)
+set(hFig, 'Position', [0 0 600 300])
+grid on

### 2017-07-26 matlab/simulations.m

@@ -1,0 +1,170 @@
+function plot()
+
+    clear all;
+    close all;
+    
+
+    
+    %----------------------------------------------------
+    % theory simulations consistency
+    %----------------------------------------------------
+    eta =0.0002;
+    d = 0.00005;
+    [x2,y2,error2] = theory_simulations(eta,d);
+    eta = 0.0008;
+    d = 0.00005;
+    [x3,y3,error3] = theory_simulations(eta,d);
+    
+    hFig = figure(1);
+    histogram(error2,'Normalization','cdf','DisplayName','Simulations \sigma_\eta=0.0002');
+    hold on
+    plot(x2,y2,'LineWidth',2,'DisplayName','Theory \sigma_\eta=0.0002');
+    hold on 
+    histogram(error3,'Normalization','cdf','DisplayName','Simulations \sigma_\eta=0.0008');
+    hold on
+    plot(x3,y3,'LineWidth',2,'DisplayName','Theory \sigma_\eta=0.0008');
+    hold on 
+    grid on
+    set(gca,'FontSize',16);
+    xlabel('\gamma');
+    ylabel('CDF');
+    xlim([-3 2]);
+    legend('show')
+    set(hFig, 'Position', [0 0 600 300])    
+    %Save Figure
+    %print(hFig,'sim-theory','-depsc2')
+    
+    eta = 8e-04;
+    d = 41*(1e-5);
+    eta_ss = eta^2;
+    d_ss = d^2;
+    %----------------------------------------------------
+    % period vs variance
+    %----------------------------------------------------
+    hFig = figure(2);
+    subplot(1,2,1);
+    alpha = 1;
+    period=[1:1:25];
+    variance = get_variance(eta_ss,d_ss,period,alpha);
+    plot(period,variance,'LineWidth',2);
+    grid on
+    set(gca,'FontSize',16);
+    xlabel({'Communication period \tau (second)'});
+    ylabel({'\sigma^2_\gamma (second^2)'});
+    xlim([1 25]);
+    ylim([0 15e-5])
+    %set(gca,'XTick',[0:10:100]);
+    %set(hFig, 'Position', [0 0 600 300]);
+    
+    % Save Figure
+    print(hFig,'period-vs-variance','-depsc2')
+
+    %----------------------------------------------------
+    % alpha vs variance
+    %----------------------------------------------------
[…]

### 2017-07-26 source codes/edb/edb-measurements.c

@@ -1,0 +1,60 @@
+/*
+ Author: Amjad Yousef Majid
+ Date: 06/feb/2017
+*/
+#include <msp430.h>
+#include <stdbool.h>
+#include <stdint.h>
+#include <stdlib.h>
+
+#include <libmsp/clock.h>
+#include <libmsp/gpio.h>
+#include <libmsp/periph.h>
+#include <libmsp/watchdog.h>
+
+#ifdef CONFIG_EDB
+#include <libedb/edb.h>
+#endif
+
+#ifdef TEST_EIF_PRINTF
+#include <libio/printf.h>
+#endif // TEST_EIF_PRINTF
+
+#include "pins.h"
+
+#include <stdint.h>
+
+#define SECOND 31250
+#define beta 0.0001f
+#define tau 1.0f
+
+_attribute_((section(".infoA"))) float f_w = 0.0f;
+
+uint16_t t_f = 0;
+
+void wisp_sync(uint16_t c_w) {
+  float error = (float)c_w - ((float)t_f + tau * (1.0f + f_w));
+  f_w += beta * error;
+}
+
+static void init_hw() {
+  msp_watchdog_disable();
+  msp_gpio_unlock();
+  msp_clock_setup();
+}
+
+int main() {
+  init_hw();
+#ifdef CONFIG_EDB
+  edb_init();
+#endif
+
+  uint16_t i;
+  WATCHPOINT(0);
+  for (i = 0; i < 10; i++) {
+    wisp_sync(i);
+  }
+  WATCHPOINT(1);
+
+  return 0;
+}

### 2017-07-26 source codes/wisp-sync.c

@@ -1,0 +1,100 @@
+#include <in430.h>
+#include <inttypes.h>
+#include <msp430.h>
+#include <stdint.h>
+#include <stdlib.h>
+
+/* to compile
+  msp430-gcc -lm -mmcu=msp430fr5969 wisp-sync.c
+*/
+
+#define port_init(port, bit)                                                   \
+  P##port##DIR |= BIT##bit;                                                    \
+  P##port##OUT &= ~BIT##bit;
+
+#define port_on(port, bit) P##port##OUT |= BIT##bit;
+#define port_off(port, bit) P##port##OUT &= ~BIT##bit;
+#define port_toggle(port, bit) P##port##OUT ^= BIT##bit;
+
+#define LED1 BIT6
+#define LED1_OUT P4OUT
+#define LED1_DIR P4DIR
+
+#define LED2 BIT0
+#define LED2_OUT P1OUT
+#define LED2_DIR P1DIR
+
+/* Macros for easy access to the LEDs on the evaulation board. */
+#define led_init(led)                                                          \
+  led##_DIR |= led;                                                            \
+  led##_OUT &= ~led;
+
+#define led_on(led) led##_OUT |= led;
+#define led_off(led) led##_OUT &= ~led;
+#define led_toggle(led) led##_OUT ^= led;
+
+#define SECOND 31250
+#define beta 0.0001f
+#define tau 1.0f
+
+__attribute__((section(".infoA"))) float f_w = 0.0f;
+
+uint8_t even = 0;
+uint16_t t_f;
+
+void wisp_sync(uint16_t c_w) {
+  float error = (float)c_w - ((float)t_f + tau * (1.0f + f_w));
+  f_w += beta * error;
+}
+
+int main() {
+  WDTCTL = WDTPW | WDTHOLD; // Stop watchdog timer
+  PM5CTL0 &= ~LOCKLPM5; // Disable the GPIO power-on default high-impedance mode
+
+  // Disable FRAM wait cycles to allow clock operation over 8MHz
+  FRCTL0 = 0xA500 | ((1) << 4); // FRCTLPW | NWAITS_1;
+  __delay_cycles(3);
+
+  /* init FRAM */
+  FRCTL0_H |= (FWPW) >> 8;
+  __delay_cycles(3);
+
+  /* stop timer */
+  TA2CCTL0 = 0x00;
+  TA2CTL = 0;
[…]

### 2017-10-25 manuscript_tmc.tex
TMC submission version

--- 2017-07-26 manuscript.tex
@@
-\documentclass[journal,draftcls,onecolumn,12pt,twoside]{IEEEtranTCOM}
-
-\normalsize
-%\documentclass[english,journal]{IEEEtran}
-
-% Metadata Information
-%\acmVolume{9}
-%\acmNumber{4}
-%\acmArticle{39}
-%\acmYear{2016}
-%\acmMonth{3}
-
-% Copyright
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
-
-% DOI
-%\doi{0000001.0000001}
-
-%ISSN
-%\issn{1234-56789}
-
-%\usepackage[latin9]{inputenc}
-%\usepackage{hyphenat}
-%\usepackage{epstopdf}
-%\usepackage{flushend}
+\documentclass[10pt,journal,compsoc]{IEEEtran}
+
+\ifCLASSOPTIONcompsoc
+  % IEEE Computer Society needs nocompress option
+  % requires cite.sty v4.0 or later (November 2003)
+  \usepackage[nocompress]{cite}
+\else
+  % normal IEEE
+  \usepackage{cite}
+\fi
+
 \usepackage{algorithm,setspace}
@@
 
-\newcommand{\expected}[1]{E\left\llbracket #1 \right\rrbracket}
+\newcommand{\expected}[1]{E\!\left\llbracket #1 \right\rrbracket}
 \newcommand{\revised}[1]{\textcolor{blue}{#1}}
@@
 % text revision commands
-%\newcommand{\add}[1]{\textcolor{blue}{#1}}
-%\newcommand{\remove}[1]{\textcolor{red}{\st{#1}}}
-%\newcommand{\replace}[2]{\remove{#1} \add{#2}}
-
-%\makeatother
-
-\setcounter{secnumdepth}{6}
-
-% Document starts
+\newcommand{\add}[1]{\textcolor{blue}{#1}}
+\newcommand{\remove}[1]{\textcolor{red}{\st{#1}}}
+\newcommand{\replace}[2]{\remove{#1} \add{#2}}
+
 \begin{document}
 
-% Title portion
-\title{On the Synchronization of Intermittently Powered Wireless Embedded Systems}
-
-\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, Henko~Aantjes, Przemys{\l}a~Pawe{\l}czak and Amjad~Yousef~Majid
-
-\thanks{A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}.}
-\thanks{The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).}}
-
-\markboth{IEEE Transactions on Communications}%
-{Submitted paper}
-
-\maketitle
-
-\vspace{-1cm}
-
+\title{ \color{blue} On the Synchronization of Computational RFIDs}
+
+\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, 
+	Henko~Aantjes, 
+	Przemys{\l}a~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
+	and~Amjad~Yousef~Majid
+	%
+	\IEEEcompsocitemizethanks{
+		\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}. \protect\\
+		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).
+	}
+%	\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}
+}
+
+\IEEEtitleabstractindextext{%
 \begin{abstract}
@@
 
-\begin{IEEEkeywords}Time Synchronization, Computational RFIDs, Wireless Identification and Sensing Platform (WISP).
-\end{IEEEkeywords}
+% Note that keywords are not normally used for peerreview papers.
+\begin{IEEEkeywords}
+Time Synchronization, Computational RFIDs, Wireless Identification and Sensing Platform (WISP).
+\end{IEEEkeywords}}
+
+\maketitle
+
+\IEEEdisplaynontitleabstractindextext
 
@@
 
-\section{Introduction}
-\label{sec:Intro}
-
-\bstctlcite{BSTcontrol}\IEEEPARstart{L}{ow-power} wireless embedded systems consist of tiny, low-cost and spatially separated computers that communicate with each other by exchanging radio packets.
+\IEEEraisesectionheading{\section{Introduction}\label{sec:Intro}}
+
+\IEEEPARstart{L}{ow-power} wireless embedded systems consist of tiny, low-cost and spatially separated computers that communicate with each other by exchanging radio packets.
 Powering these small-scale embedded systems, e.g.
@@
 
+{\color{blue}{\subsection{The Need for CRFID Network Synchronization}}}
+
 As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}.
@@
 We anticipate that these multi-hop networks will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.
-In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences.
-
-%\add{\textbf{Distributed Sensing.} As an example, consider applications of monitoring inaccessible areas or general surveillance in which multiple battery-less cameras, such as WISPCam~\cite{wispcam_2015}, capture images from different locations simultaneously and forward the collected data to the RFID reader.
-In such applications, the WISPCams can coordinate their image capture actions by means of the almost simultaneously received periodic ``\texttt{perfrom image snapshot}'' commands from the RFID reader.
-However, the reception and processing of the commands, and in turn the coordination, can be interrupted since WISPCams might exhibit different \emph{charge/discharge} patterns.
-In principle, WISPCams operate in a power cycling mode, in which their storage capacitor is fully-charged in the charging phase and then discharged until a lower threshold (no operation) voltage is attained.
-Due to their different distances to the RFID reader leading to different harvested energy levels, it can happen that a WISPCam would have to wait until it is fully-charged to capture the image.
-It has been observed that ``the charging time for voltage to reach 1.8\,V to power a WISP equipped with a 100\,$\mu$F capacitor can be as large as 155 seconds, when the RFID reader is 10.0 meters away''~\cite{fu2013minimizing}.
-This extra delay might result in the loss of coordination and the capture of images at different time instants.
-To prevent this situation, the reader can send ``\texttt{perform image snapshot $x$ seconds later}' command and continue charging the WISPCams.
-Then, WISPCams can activate themselves and capture the image simultaneously, just after $x$ seconds after the reception of the reader command.
-Since each WISPCam has its own clock hardware whose oscillator generate pulses at slightly different speeds, a \emph{synchronization} service that allows WISPCams to wake-up simultaneously and facilitates such collaborative and coordinated actions becomes a fundamental requirement for these systems.}
-
-%that the charging time for voltage to reach 1.8V to power a WISP equipped with a 100uF capacitor can be as large as 155 seconds, when the RFID reader is 10.0 meters away
-
-%It should be noted that there can also be a degree of synchronization on the 
-%reader side by timestamping the received 
-%information. However, the WISPCams that experience a power 
-%loss might resume transmission of the captured image to the reader after a 
-%relatively long delay due to harvesting sufficient energy. In this case, the 
-%timestamps do not make sense since they do not reflect the actual capture time 
-%of the images. Therefore, the establishment of a ``common time'' notion
-
+In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences \add{at the CRFID tag side}.
+\add{Therefore, establishing a common time notion among the tags, rather than synchronizing RFID readers only, will be a crucial service.}
+
+{\color{blue}{\subsection{Differences From Battery-Powered WSNs }}}
 \label{sec:Motivation}
 
-The characteristics of CRFID systems expose fundamentally different challenges than sensor networks to implement the synchronization service. The reason is mainly twofold:
+\add{However, }the characteristics of CRFID systems expose fundamentally different challenges than \add{battery-powered} sensor networks to implement the synchronization service.
+The reason is mainly twofold:
 
 \begin{itemize}
-\item \textbf{Challenge I:} CRFID systems should perform computations in an energy efficient manner despite intermittent RF power that leads to frequent loss of computational state, e.g.
+	\item \textbf{Challenge I:} CRFID systems should perform computations \replace{in an energy efficient manner despite}{efficiently using marginal and} intermittent RF power that leads to frequent loss of computational state, e.g.
 when the RFID reader moves away from the CRFID tag \cite{ransford:2008}.
+\add{Therefore, computations pertaining to time synchronization have an extremely short time budget, e.g.
+should be completed with the other computation tasks within 10--100\,ms in WISP~\cite{Ransford:2011:Mementos}.
+On the contrary, the synchronization approaches in WSNs, e.g.
+the de facto protocol \cite{Maroti2004}, keep communication infrequent by pushing computation instead, so that the battery power is used efficiently in order to extend the life-time of the network up to several years.}
 
@@
 varying oscillator frequencies in short-term affects the stability of the clocks and degrades the accuracy sensing.
+\add{Therefore, computations pertaining to time synchronization should be reactive against very-frequent voltage effects; e.g.
+during super-capacitor discharge.
+On the contrary, in WSNs the input voltage is constant and the temperature is the main factor affecting the clock frequencies~\cite{schmid2010interaction}.
+Temperature compensation techniques, e.g.
+\cite{Schmid:2009}, are not applicable to compensate for the voltage instabilities in CRFIDs due to their computational complexity.}
+
 \end{itemize}
 
-The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered wireless embedded systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices.
+{\color{blue}{\subsection{Contributions}}}
+
+The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered \replace{wireless embedded}{CRFID} systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices.
 In particular, the main contributions and findings of
@@
 	\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems.
-Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides explicit synchronization of individual tags in the communication range of an RFID reader.
-\item By addressing the aforementioned challenges, we design and implement\emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015}.
+Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides synchronization of individual tags in the communication range of an RFID reader.
+	\item By addressing the aforementioned challenges, we design and implement \emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015,yildirim2017adaptive}.
 WISP-Sync (i)
@@
 
-\section{Synchronizing Battery-Free Wireless Embedded Systems}
+\section{Synchronizing Battery-Free {\color{blue}CRFIDs}}
 \label{sec:Related-Work}
 
-In this section, we present a brief  state of the art on synchronization issues in conventional WSNs. Then, we emphasize the 
-main challenges of synchronizing wireless battery-free embedded systems by focusing on a particular IPD platform: computational RFIDs.
-
-\subsection{Synchronization in Conventional Wireless Sensor Networks}
+In this section, we present a brief  state of the art on synchronization issues in conventional WSNs.
+Then, we emphasize the main challenges of synchronizing wireless battery-free embedded systems by focusing on a particular IPD platform: computational RFIDs.
+
+\subsection{Synchronization in Conventional WSNs}
 \label{sec:WSN-Sync}
@@
 
-\subsubsection{Computation Methods}
+\subsubsection{{\color{blue}State-of-the-art Synchronization Techniques}}
 
@@
 On the other hand \emph{iterative} computation methods, e.g.
-in ~\cite{Yildirim:AVTS:2014,pi2015} and \cite{Yildirim:Gradient:2016}, perform identical computation steps at each re-synchronization round.
+in ~\cite{Yildirim:AVTS:2014,pi2015,yildirim2017adaptive,Yildirim:Gradient:2016}, perform identical computation steps at each re-synchronization round.
 With iterative methods, sensor nodes establish synchronization after a finite number of rounds in a lightweight manner in terms of computation and memory requirements.
@@
 In these approaches, nodes employ computation methods based on \emph{distributed consensus}~\cite{Sommer2009Gradient,SchenatoFiorentin:2011,Yildirim:External:2014}.
-
-\subsection{Fundamental Challenges of Synchronizing IPDs: The CRFID Case}
+\add{Since the prominent factor affecting the clock frequencies in WSNs is the temperature~\cite{schmid2010case}, there are also temperature compensated time synchronization techniques, e.g.
+\cite{Schmid:2009}.
+However, such techniques require a calibration phase in order to build a temperature--clock frequency table.}
+
+\subsection{\color{blue}Fundamental Challenges of Synchronizing CRFIDs}
 \label{sec:CRFIDs}
 
-We now delve into the synchronization characteristics of RF-powered wireless embedded systems.
-To this end, we will consider the WISP, the de facto CRFID platform.
+We now delve into the synchronization characteristics of \remove{RF-powered wireless embedded systems.
+To this end, we will consider the} WISP, the de facto CRFID platform.
 The main aspects pertaining to synchronization in the WISP platform can be stated as follows:
 
-\paragraph{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication.
+\subsubsection{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication.
 Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}.
@@
 	\centering
-	\includegraphics[scale=0.35]{figures/clock_frequency_dc_rf.eps}
+	\includegraphics[width=0.85\columnwidth]{figures/clock_frequency_dc_rf.eps}
 	\caption{\label{fig:frequency_dc_rf} In our testbed, see Section~\ref{sec:testbed} for the details, we programmed a single WISP tag so that it toggles one of its output port whenever its 32~KHz built-in timer progresses 32000 clock ticks.
@@
 We plotted our measurements when the tag is powered by the Flash Emulation Tool (FET) with a stable DC power and when the tag is powered by the RFID Reader at a distance of 5~cm.
-Since harvested RF energy was different and unstable than the constant DC case, the frequency of the oscillator and in turn the toggle event was quite different.
-We measured the mean and standard deviation of the frequency in the DC case as 1.1653 and 0.00028553, and these values for the RF case were 1.1793 and 0.00026049, respectively.
+\remove{Since harvested RF energy was different and unstable than the constant DC case, the frequency of the oscillator and in turn the toggle event was quite different.
+}We measured the mean and standard deviation of the frequency in the DC case as 1.1653 and 0.00028553, and these values for the RF case were 1.1793 and 0.00026049, respectively.
 }
@@
 
-\paragraph{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term.
+\subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term.
 On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift.
-Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level rather than the temperature of the WISP tags, see Fig.~\ref{fig:frequency_dc_rf}.
-
-\paragraph{Frequent loss of synchronization state} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g.
+Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level\remove{ rather than the temperature of the WISP tags}, see Fig.~\ref{fig:frequency_dc_rf}.
+
+\subsubsection{Frequent loss of synchronization state} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g.
 the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up.
@@
 
-\paragraph{Computation and memory overhead sensitivity} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p.
+\subsubsection{Computation and memory overhead sensitivity} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p.
 44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}.
@@
 
-\paragraph{Limitations of the EPC Gen 2 standard} The WISP firmware implements the EPC Gen 2 standard~\cite{epc_gen2} which increases the compatibility with the existing RFID systems.
+\subsubsection{Limitations of the EPC Gen 2 standard} The WISP firmware implements the EPC Gen 2 standard~\cite{epc_gen2} which increases the compatibility with the existing RFID systems.
 However, the standard introduces limitations, e.g.
@@
 
-\subsection{Low Power Operation Modes and WISP Clock System}
+\subsection{Low Power Operation Modes and The Clock System}
 \label{subsec:clock_system}
@@
 	\node[anchor=south west, inner sep=0] at 
-	(0,0){\includegraphics[height=.35\columnwidth]{figures/testbed_resized.jpg}};
-
-	\node [fill= white,circle] at (9,2) { };
-	\node [color=black ] at (9,2) { 1 };
-
-	\node [fill= white,circle] at (9,4) {  };
-	\node [color=black ] at (9,4) { 2 };
+	(0,0){\includegraphics[width=\columnwidth]{figures/testbed_resized.jpg}};
+
+	\node [fill= white,circle] at (7,2) { };
+	\node [color=black ] at (7,2) { 1 };
+
+	\node [fill= white,circle] at (8,4) {  };
+	\node [color=black ] at (8,4) { 2 };
 
@@
 
-	\node [fill= white,circle] at (5.5,2) {  };
-	\node [color=black ] at (5.5,2) { 4 };
-
-	\node [fill= white,circle] at (8,5) {  };
-	\node [color=black ] at (8,5) { 5 };
+	\node [fill= white,circle] at (5.1,2) {  };
+	\node [color=black ] at (5.1,2) { 4 };
+
+	\node [fill= white,circle] at (6.5,4.5) {  };
+	\node [color=black ] at (6.5,4.5) { 5 };
 
@@
 	\centering
-\includegraphics[scale=0.35] {figures/relative_clock_frequency.eps}
+	\includegraphics[width=1.4\columnwidth] {figures/relative_clock_frequency.eps}
 	\caption{\label{fig:frequency}The frequencies of the toggle events (obtained using the same steps presented in Fig.\ref{fig:frequency_dc_rf}) of two RF-powered WISP tags, placed at 5 cm distance from the RFID reader.
@@
 	\centering
-\includegraphics[scale=0.8]{figures/Sender-Receiver.pdf}
+	\includegraphics[width=\columnwidth]{figures/Sender-Receiver.pdf}
 	\caption{\label{fig:sender_receiver} The message exchange among the host computer, the RFID reader and the tag for sender-receiver synchronization.
@@
 
-\paragraph{Transmission Delay and Jitter Measurements}
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figures/sender_receiver_uncertainty.eps}
+\subsection{Transmission Delay and Jitter Measurements}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\columnwidth]{figures/sender_receiver_uncertainty.eps}
 	\caption{\label{fig:transmission_delay} The normalized number of occurrences of the transmission delays measured by sniffing the communication between the RFID reader and the WISP tag.
@@
 
-\paragraph{Collection of Clock Values for Offline Processing}
+\subsection{Collection of Clock Values for Off-line Processing}
 
@@
 
-\subsection{Synchronization Approach 1: Software Clock Computation Using  Least-Squares Regression}
+\subsection{Software Clock Computation Using  Least-Squares}
 
@@
 	\centering
-\includegraphics[scale=0.35]{figures/least_squares.eps}
+	\includegraphics[width=0.9\columnwidth]{figures/least_squares.eps}
 	\caption{\label{fig:least-squares}Synchronization error by employing least-squares regression on the collected timestamps.
-We observed a maximum synchronization error of 10 clock ticks between the RFID reader and the WISP tag, leading to 3.2\,ms synchronization accuracy.}
+We observed a maximum synchronization error of 10 clock ticks between the RFID reader and the WISP tag, leading to \replace{3.2}{0.32}\,ms synchronization accuracy.}
 \end{figure}
@@
 
-\paragraph{Limitations of the Approach 1}
+\subsection{Limitations of the Approach}
 
@@
 
-\emph{Host computer computation:} We are unaware of any EPC Gen2 command that will allow to send the \emph{FistSeenTimestamp} to the tag.
+\subsubsection{Host computer computation} We are unaware of any EPC Gen2 command that will allow to send the \emph{FistSeenTimestamp} to the tag.
 Hence, even though we were able to collect $(C_w(t),C_r(t))$ pairs for offline processing, it is not possible for the tag to collect $C_r(t)$ and synchronize itself to the reader.
@@
 
-\emph{Lack of broadcast primitive:} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader.
+\subsubsection{The Lack of broadcast primitive} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader.
 Unfortunately, in the current EPC Gen2 and LLRP standard, we were unable to find any mechanism to send a global timestamp that is received by all tags inside the range of the RFID reader to establish synchronization in one step.
 
-\emph{Big and reader dependent transmission delays:} As depicted previously, we observed that transmission delays in our testbed were on the order of milliseconds, see Fig.~\ref{fig:least-squares}, whereas they are reported as a few microseconds in conventional WSN platforms.
+\subsubsection{Big and reader dependent transmission delays} As depicted previously, we observed that transmission delays in our testbed were on the order of milliseconds, see Fig.~\ref{fig:least-squares}, whereas they are reported as a few microseconds in conventional WSN platforms.
 This is one of the crucial limitations since the smaller the transmission delays are, the better the synchronization.
@@
 
-In this section, we introduce a simple but novel \emph{event-based} reader-tag synchronization inspired by simple PI controllers and provide its advantages over Approach 1.
+In this section, we introduce a simple but novel \emph{event-based} reader-tag synchronization inspired by simple PI controllers and provide its advantages over \replace{Approach 1}{the previous approach}.
 In event-based synchronization, a common event which is observable by all receiver devices simultaneously is generated by the reference device.
@@
 	\centering
-\includegraphics[scale=0.8]{figures/Event-Based.pdf}
+	\includegraphics[width=\columnwidth]{figures/Event-Based.pdf}
 	\caption{\label{fig:event_based}Event-Based synchronization steps: The tag timestamps successive \emph{BlockWrite} events and adjusts its software clock.}
@@
 
-\paragraph{Event Period Measurements}
+\subsection{Event Period Measurements}
 \label{subsec:event_period_uncertainty}
@@
 	\centering
-\includegraphics[scale=0.35]{figures/blockwrite_delay.eps}
+	\includegraphics[width = 0.9\columnwidth]{figures/blockwrite_delay.eps}
 	\caption{\label{fig:block_write_delay}The delay between the first and the last \emph{BlockWrite} event, i.e.
@@
 
-\subsection{Synchronization Approach 2: Event-Based Synchronization Inspired By PI Controllers }
+\subsection{Event-Based Synchronization Inspired By PI Controllers}
 
@@
 
-Considering these facts, we designed a lightweight and adaptive clock synchronization approach inspired by the PI-controller based solution introduced in~\cite{pi2015}.
-Let $t_0$ denote the first event reception time at the tag.
-Using the pre-measured event period $\tau$ of the RFID reader, the tag can estimate the next event reception time at time $t_1$ as
+Considering these facts, we designed a lightweight and adaptive clock synchronization approach inspired by the PI-controller based solution introduced in~\cite{pi2015,yildirim2017adaptive}: \add{(i) we apply PI-controller based clock computation technique to the WISP platform due to its computational efficiency and fast adaptation; (ii) we introduce power-interrupt recovery mechanism so that the WISP tags do not loose their synchronization-related information.
+}
+
+Let $t_0$ denote the first event reception time at the tag. Using the pre-measured event period $\tau$ of the RFID reader, the tag can estimate the next event reception time at time $t_1$ as
 \begin{eqnarray}
@@
 \begin{eqnarray}
-\gamma(t_1) & = & 
-C_{w}(t_{1})-\hat{C}_{w}(t_{1}) = 
-C_{w}(t_{1})-C_{w}(t_{0})-\tau(1+\hat{f}_{w}^{r}(t_0)). \label{eq:gamma_t_1}
+\gamma(t_1) & \!\!\!\!\!=\!\!\!\!\!\! & 
+C_{w}(t_{1}\!)\!-\!\hat{C}_{w}(t_{1}\!) \!=\! 
+C_{w}(t_{1}\!)\!-\!C_{w}(t_{0}\!)\!-\!\tau(1\!\!+\!\!\hat{f}_{w}^{r}(t_0\!)). \label{eq:gamma_t_1}
 \end{eqnarray}
@@
 \begin{eqnarray}
-\gamma & \sim & \mathcal{N} \left(0,\sigma^2_\gamma=\frac{\sigma^2_{\eta}\left(\tau^2-\beta\tau^3+\frac{\beta^2\tau^4}{3}\right)+\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)}+\frac{\sigma^2_{\eta}\tau^3}{3}+  \sigma^2_{\tau} \right),
+\gamma & \!\!\!\!\!\sim\!\!\!\!\!
+& \mathcal{N} \!\!\left(\!\!0,\sigma^2_\gamma\!\!=\!\!\frac{\sigma^2_{\eta}\!\left(\!\tau^2\!-\!\beta\tau^3\!+\!\frac{\beta^2\tau^4}{3}\!\right)\!\!+\!\!\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)}\!\!+\!\!\frac{\sigma^2_{\eta}\tau^3}{3}\!+\!\sigma^2_{\tau} \!\!\right),
 \label{eq:sync_performance}
@@
 
-\begin{figure*}
-	\centering
-	\includegraphics[scale=0.35]{figures/sim-theory.eps}
+\begin{figure}
+	\centering
+	\includegraphics[width=\columnwidth]{figures/sim-theory.eps}
 	\caption{\label{fig:theory-vs-simulations} The plot of the cumulative distribution function (CDF) of $\gamma$ in \eqref{eq:sync_performance} and the CDF of the steady-state $\gamma$ obtained during numerical simulations of \eqref{eq:update_rule} (by implementing \eqref{eq:theta} and \eqref{eq:delta_recursive} in Appendix).
 We used $\tau=1$, $\beta= 0.0001$, $\sigma_{\tau}=0.00005$ and our plots depict the theoretical and simulation results with respect to $\sigma_\eta = 0.0002$ and $\sigma_\eta = 0.0008.$ It can be observed that \eqref{eq:sync_performance} is consistent with the simulations and bigger uncertainty leads to bigger synchronization error.}
-\end{figure*}
+\end{figure}
 In Fig.~\ref{fig:theory-vs-simulations}, we emphasize that our theoretical derivation in \eqref{eq:sync_performance} is consistent with the simulations of \eqref{eq:update_rule}.
@@
 \begin{algorithm}[t]
-			\setstretch{1.35}
+%	\setstretch{1.35}
 	\caption{\label{alg:sync} \emph{WISP-Sync}: An integral controller-based 
@@
 	\begin{algorithmic}[1]	
-		\scriptsize
+		\small
 
@@
 			frequency}	
-		\Statex $\tau\leftarrow$measured value of a particular RFID reader from 
-		testbed 
-		\Comment{mean \emph{event period}}	
-		\Statex $\beta\leftarrow$ choose a value satisfying 
+		\Statex $\tau\leftarrow$measured value \Comment{mean \emph{event period}}	
+		\Statex $\beta\leftarrow$ a value satisfying 
 		$0<\beta<\frac{2}{\tau}$ \Comment{integral gain}	
-		\Statex $V_{\min}, V_{\text{op}}, V_{\text{cur}}$ \Comment{minimum voltage threshold, operation voltage threshold ($V_{\text{op}}>V_{\min}$) and current voltage}
+		\Statex $V_{\min}, V_{\text{op}}$ \Comment{voltage thresholds ($V_{\text{op}}>V_{\min}$)}
+		\Statex $V_{\text{cur}}$ \Comment{current voltage}
 
@@
 		\State \quad $\hat{f}_w^r = \hat{f}_w^r +\beta\gamma$ \Comment{apply 
-			integral control 
-			to update relative clock 
-			frequency}
+			integral control}
 		\State $\square$ \textbf{Upon} $V_{\text{cur}}=V_{\min}$
-		\State \quad store $\hat{f}_w^r$ into FRAM \Comment{store the relative 
-			clock 
-			frequency in non-volatile memory}
+		\State \quad store $\hat{f}_w^r$ into FRAM \Comment{store in non-volatile memory}
 		\State $\square$ \textbf{Upon} $V_{\text{cur}} < V_{\text{op}}$
-		\State \quad restore $\hat{f}_w^r$ from FRAM \Comment{restore the 
-			relative 
-			clock frequency}
+		\State \quad restore $\hat{f}_w^r$ from FRAM \Comment{restore the clock frequency}
 	\end{algorithmic}	
@@
 	\centering
-\includegraphics[scale=0.35]{figures/alpha-period-vs-variance.eps}
-\caption{\label{fig:simulations} Numerical results showing the relations between communication period, integral gain and the steady-state variance.
+	\includegraphics[width=1.4\columnwidth]{figures/alpha-period-vs-variance.eps}
+	\caption{\label{fig:simulations} Numerical results showing the relations between \replace{communication}{event} period, integral gain and the steady-state variance.
 First, we set $\sigma_{\tau}=0.4097$, $\sigma_\eta = 0.0008$, $\beta=1$ and plotted $\sigma^2_{\gamma}$ using different $\tau$ (left hand side).
@@
 
-\paragraph{Selection of the Integral Gain $\beta$}
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figures/integral_gain.eps}
+\subsection{Selection of the Integral Gain $\beta$}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.85\columnwidth]{figures/integral_gain.eps}
 	\caption{\label{fig:beta} The synchronization error tends to get smaller with smaller integral gains, but after some point decreasing it has no significant effect due to the precision of 32\,kHz clock.
@@
 
-\paragraph{Synchronization Performance}
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figures/synchronization_accuracy.eps}
+\subsection{Synchronization Performance}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\columnwidth]{figures/synchronization_accuracy.eps}
 	\caption{\label{fig:sync_error}Event-based synchronization performance when the tag is powered through a constant voltage source and through RF power.
@@
 
-\paragraph{Power Loss and Recovery}
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.35]{figures/power_loss.eps}
+\subsection{Power Loss and Recovery}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.85\columnwidth]{figures/power_loss.eps}
 	\caption{\label{fig:power_loss}Impact of the state recovery on the synchronization performance.
@@
 %
-\paragraph{Energy Overhead}
+\subsection{Energy Overhead}
 %
@@
 
-\paragraph{Limitations of WISP-Sync}
+\subsection{Limitations of WISP-Sync}
 
@@
 	\node[anchor=south west, inner sep=0] at 
-	(0,0){\includegraphics[height=.3\columnwidth]{figures/3-wisps_resized.jpg}};
+	(0,0){\includegraphics[width=0.8\columnwidth]{figures/3-wisps_resized.jpg}};
 
@@
 
-	\node [fill= white,circle] at (1,4.5) {  };
-	\node [color=black ] at (1,4.5) { 2 };
+	\node [fill= white,circle] at (1,5.1) {  };
+	\node [color=black ] at (1,5.1) { 2 };
 
@@
 	\centering
-\includegraphics[scale=0.35]{figures/multiwisp-experiments.eps}
+	\includegraphics[width=0.9\columnwidth]{figures/multiwisp-experiments.eps}
 	\caption{\label{fig:multi-sync-results}The synchronization error calculated during the experiments in our testbed.
@@
 
-% Bibliography
-\bibliographystyle{IEEEtranTCOM}
-\bibliography{IEEEabrv,references}
-
-\pagebreak
-%\appendix[Analysis of Algorithm~\ref{alg:sync}] 
-\appendix
+\appendices
+\section{Analysis of Algorithm~\ref{alg:sync}}
 
@@
 \begin{eqnarray}
-\bar{f}_{w}^{r}(t_0,t_0+\delta) & \triangleq & 
-\frac{1}{\delta}\int_{t_0}^{t_0+\delta} 
+\bar{f}_{w}^{r}(t_0,t_0\!+\!\delta) &\!\!\!\!\! \triangleq\!\!\!\!\! & 
+\frac{1}{\delta}\!\!\int_{t_0}^{t_0+\delta}\!\!\!\!\!\!\!\!\!\!\!\! 
 f_w^r(t)dt \triangleq
-f_{w}^{r}(t_{0})+\frac{1}{\delta}\int_{0}^{\delta}\int_{t_{0}}^{t_{0}+t}\eta(u)dudt.
+f_{w}^{r}(t_{0})\!+\!\frac{1}{\delta}\!\!\int_{0}^{\delta}\!\!\int_{t_{0}}^{t_{0}+t}\!\!\!\!\!\!\!\!\!\!\!\eta(u)dudt.
 \label{eq:drift_relation}
@@
 \gamma(t_1) & = & 
-\tau(\bar{f}_{w}^{r}(t_0,t_1)-\hat{f}_{w}^{r}(t_0))+ \varepsilon = 
-\tau\tilde{f}_{w}^{r}(t_0)+\frac{\tau}{t_1-t_0}\int_{0}^{t_1-t_0}\int_{t_{0}}^{t_0+t}\eta(u)dudt
+\tau(\bar{f}_{w}^{r}(t_0,t_1)-\hat{f}_{w}^{r}(t_0))+ \varepsilon \nonumber \\
+ &=&  
+\tau\tilde{f}_{w}^{r}(t_0)+\frac{\tau}{t_1-t_0}\int_{0}^{t_1-t_0}\!\!\!\int_{t_{0}}^{t_0+t}\!\!\!\!\!\!\!\!\!\!\!\eta(u)dudt
 + \varepsilon.
@@
 &\overset{\eqref{eq:relative_drift_estimation_error}}{=}& \Delta(h)
-+f_w^r(t_{h+1})-f_w^r(t_{h}) -\hat{f}_w^r(t_{h+1})+\hat{f}_w^r(t_{h})
+\!+\!f_w^r(\!t_{h+1}\!)\!-\!\!f_w^r(\!t_{h}\!) \!-\!\!\hat{f}_w^r(\!t_{h+1}\!)\!+\!\!\hat{f}_w^r(\!t_{h}\!)
 \nonumber\\
@@
 	\underbrace{
-		\left[\begin{array}{c}
-		\Theta(h+1)\\
-		\Delta(h+1)
-		\end{array}\right]
+		\left[\!\!\!\begin{array}{c}
+		\Theta(h\!+\!1)\\
+		\Delta(h\!+\!1)
+		\end{array}\!\!\!\right]
 	}
-} & = & 
+} &\!\!\!\!\!\!\! =\!\!\!\!\!\!\! & 
 \underset{A}{
 	\underbrace{
-		\left[\begin{array}{cc}
+		\left[\!\!\!\begin{array}{cc}
 		0 & \tau\\
-		0 & 1-\beta\tau
-		\end{array}\right]
+		0 & 1\!-\!\beta\tau
+		\end{array}\!\!\!\right]
 	}
@@
 	\underbrace{
-		\left[\begin{array}{c}
-		\Theta(h)\\
-		\Delta(h)
-		\end{array}\right]}
+		\left[\!\!\!\begin{array}{c}
+		\Theta(\!h\!)\\
+		\Delta(\!h\!)
+		\end{array}\!\!\!\right]}
 }
-+
+\!\!+\!\!
 \underset{Y(h)}{
 	\underbrace{
-		\left[\begin{array}{c}
+		\left[\!\!\!\begin{array}{c}
 		\Omega(h)+\varepsilon\\
-		\Phi(h)-\beta(\Omega(h)+\varepsilon)
-		\end{array}\right]}
+		\Phi(\!h\!)\!-\!\beta(\Omega(\!h\!)\!+\!\varepsilon)
+		\end{array}\!\!\!\right]}
 }.\label{eq:state_space-pi}
@@
 \left(1-\beta 
-\tau\right)\expected{\Delta(\infty)} &\implies& \expected{\Delta(\infty)}=0, 
+\tau\right)\expected{\Delta(\infty)} &\!\!\!\!\!\!\!\implies\!\!\!\!\!\!\!& \expected{\Delta(\infty)}\!=\!0, 
 \label{eq:delta_convergence}\\
 \expected{\Theta(\infty)}  \overset{\eqref{eq:theta}}{=} 
-\tau\expected{\Delta(\infty)}&\implies&\expected{\Theta(\infty)}=0. 
+\tau\expected{\Delta(\infty)}&\!\!\!\!\!\!\!\implies\!\!\!\!\!\!\!&\expected{\Theta(\infty)}\!=\!0. 
 \label{eq:expected_error}
@@
 \begin{eqnarray}
-\expected{\Theta(\infty)^2} &\overset{\eqref{eq:theta}}{=}& 
+\expected{\Theta(\infty)^2} &\!\!\!\!\overset{\eqref{eq:theta}}{=}\!\!\!\!& 
 \tau^2\expected{\Delta(\infty)^2}  
@@
 \begin{eqnarray}
-\expected{\Delta(\infty)^2} 
-&\overset{\eqref{eq:delta_recursive},\eqref{eq:asymp_delta}}{=}& 
-(1-\beta\tau)^2	
-\expected{\Delta(\infty)^2}  
-+\expected{(\Phi(h)-\beta(\Omega(h)+\varepsilon)^2} 
-\label{eq:expected_delta_infinity}.
+\!\!\!\!\!\expected{\!\Delta(\!\infty\!)^2\!} 
+&\!\!\!\!\!\!\!\overset{\eqref{eq:delta_recursive},\eqref{eq:asymp_delta}}{=}\!\!\!\!\!\!& 
+(\!1\!\!-\!\!\beta\tau\!)^2\!	
+\expected{\!\Delta(\!\infty\!)^2}  
+\!\!+\!\!\expected{\!(\!\Phi(\!h\!)\!\!-\!\!\beta\!(\!\Omega(\!h\!)\!+\!\varepsilon\!)^2}\!. 
+\label{eq:expected_delta_infinity}
 \end{eqnarray}
@@
 \!-\!2\beta\expected{\Phi(h)\Omega(h)} 
-\!+\!\beta^2\expected{\Omega(h)^2}\!+\!\beta^2\expected{\varepsilon^2}. 
+\! \nonumber \\ 
+& & +\!\beta^2\expected{\Omega(h)^2}\!+\!\beta^2\expected{\varepsilon^2}. 
 \label{eq:stochastic_expectation}
@@
 \begin{eqnarray}
-\expected{\Phi(h)^2} 
-& = & 
-\int_{0}^{t_{h+1}-t_h}\sigma_{\eta}^{2}dt =  \sigma_{\eta}^{2}(t_{h+1}-t_h),\\
-\expected{\Omega(h)^2} 
-& = & 
-\frac{\tau^2}{(t_{h+1}-t_h)^2}
-\sigma_{\eta}^{2}\int_{0}^{t_{h+1}-t_h}t^2dt = 
-\sigma_{\eta}^{2}\frac{\tau^2(t_{h+1}-t_h)}{3},\label{eq:stochastic_integral}\\
-\expected{\Phi(h)\Omega(h)} 
-& = &
-\frac{\tau}{t_{h+1}-t_h}\sigma_{\eta}^{2}\int_{0}^{t_{h+1}-t_h}tdt = 
-\sigma_{\eta}^{2}\frac{\tau(t_{h+1}-t_h)}{2}, \\
-\expected{\varepsilon^2} 
-& = & 
+\expected{\Phi(h)^2}\!\!\!\!\!\!\! 
+& = & \!\!\!\!\!\!
+\int_{0}^{t_{h+1}-t_h}\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\sigma_{\eta}^{2}dt =  \sigma_{\eta}^{2}(t_{h+1}-t_h),\\
+\expected{\Omega(h)^2}\!\!\!\!\!\!\! 
+& = & \!\!\!\!\!\!
+\frac{\tau^2}{(t_{h+1}\!\!-\!\!t_h)^2}
+\sigma_{\eta}^{2}\!\!\!\int_{0}^{t_{h+1}-t_h}\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!t^2dt\! = \!
+\sigma_{\eta}^{2}\!\frac{\tau^2(\!t_{h+1}\!\!-\!\!t_h\!)}{3},\label{eq:stochastic_integral}\\
+\expected{\Phi(h)\Omega(h)}\!\!\!\!\!\!\! 
+& = & \!\!\!\!\!\!
+\frac{\tau}{t_{h+1}-t_h}\sigma_{\eta}^{2}\int_{0}^{t_{h+1}-t_h}\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!tdt = 
+\sigma_{\eta}^{2}\frac{\tau(\!t_{h+1}\!\!-\!\!t_h\!)}{2}, \\
+\expected{\varepsilon^2}\!\!\!\!\!\!\! 
+& = & \!\!\!\!\!\! 
 \sigma_{\tau}^{2}.
@@
 \begin{eqnarray}
-\expected{\Delta(\infty)^2}
-& = &
-\frac{\sigma^2_{\eta}(t_{h+1}-t_h)\left(1-\beta\tau+\frac{\beta^2\tau^2}{3}\right)+\beta^2\sigma^2_{\tau}}{(2\beta\tau-\beta^2\tau^2)}.
+\expected{\Delta(\infty)^2}\!\!\!\!\!\!
+& = &\!\!\!\!\!\!
+\frac{\sigma^2_{\eta}(t_{h+1}-t_h)\left(1-\beta\tau+\frac{\beta^2\tau^2}{3}\right)\!\!+\!\!\beta^2\sigma^2_{\tau}}{(2\beta\tau-\beta^2\tau^2)}.
 \label{eq:delta_infinity}
@@
 \begin{eqnarray}
-\expected{\Theta(\infty)^2} & = & 
+\expected{\Theta(\infty)^2}\!\!\!\!\!\!\! & = & \!\!\!\!\!\!\!
 \frac{\sigma^2_{\eta}(t_{h+1}-t_h)\left(\tau^2
 	-\beta\tau^3-\frac{\beta^2\tau^4}{3}\right)
-	+\beta^2\sigma^2_{\tau}\tau^2}{(2\beta\tau-\beta^2\tau^2)} + 
+	+\beta^2\sigma^2_{\tau}\tau^2}{(2\beta\tau-\beta^2\tau^2)} \nonumber \\ 
+\!\!\!\!\!\!\! &  & \!\!\!\!\!\!\!+ 
 \sigma^2_{\eta}\frac{\tau^2(t_{h+1}-t_h)}{3}+\sigma^2_{\tau}.
@@
 \begin{eqnarray}
-Var\left(\underset{h\rightarrow\infty}{\gamma(t_h)}\right)& = & 
-\expected{\Theta(\infty)^2} = 
+Var\left(\underset{h\rightarrow\infty}{\gamma(t_h)}\right)\!\!\!\!\!\!\!& = &\!\!\!\!\!\! 
+\expected{\Theta(\infty)^2}\!\! = \!\!
 \frac{\sigma^2_{\eta}\left(\tau^2-\beta\tau^3
 	+\frac{\beta^2\tau^4}{3}\right)
-	+\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)}+\frac{\sigma^2_{\eta}\tau^3}{3}
+	+\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)} \nonumber \\
+\!\!\!\!\!\!\!& = &\!\!\!\!\!\!  +\frac{\sigma^2_{\eta}\tau^3}{3}
 +  \sigma^2_{\tau}
@@
 
+\ifCLASSOPTIONcaptionsoff
+  \newpage
+\fi
+
+\bibliographystyle{IEEEtran}
+\bibliography{IEEEabrv,references}
+
+% biography section
+% 
+% If you have an EPS/PDF photo (graphicx package needed) extra braces are
+% needed around the contents of the optional argument to biography to prevent
+% the LaTeX parser from getting confused when it sees the complicated
+% \includegraphics command within an optional argument. (You could create
+% your own custom macro containing the \includegraphics command to make things
+% simpler here.)
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
+% or if you just want to reserve a space for a photo:
+
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/ksy.jpeg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
+%received BSc(Eng), M.Sc. and Ph.D. degrees from Ege University, İzmir, 
+%Turkey in 2003, 2006 and 2012 respectively.
+He is  currently with Embedded Software group of TU Delft.
+His research interests are embedded systems, distributed algorithms,  wireless sensor networks, energy harvesting systems and computational RFIDs.
+%\end{IEEEbiography}
+%
+%% if you will not have a photo at all:
+%\begin{IEEEbiographynophoto}{Henko Aantjes}
+%Biography text here.
+%\end{IEEEbiographynophoto}
+%
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.jpg}}]{}
+%Biography text here.	
+%\end{IEEEbiography}
+%
+%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
+%Biography text here.
+%\end{IEEEbiography}
+%
+% insert where needed to balance the two columns on the last page with
+% biographies
+%\newpage
+
+% You can push biographies down or up by placing
+% a \vfill before or after them. The appropriate
+% use of \vfill depends on what kind of text is
+% on the last page and whether or not the columns
+% are being equalized.
+
+%\vfill
+
+% Can be used to pull up biographies so that the bottom of the last one
+% is flush with the other column.
+%\enlargethispage{-5in}
+
+% that's all folks
 \end{document}
-
-% End of v2-acmsmall-sample.tex (March 2012) - Gerry Murray, ACM

### 2017-10-25 cover letter/cover.tex
Cover TMC

@@
 
-	With this letter we are submitting an article titled \emph{On the 
-		Synchronization 
-		of Intermittently Powered Wireless Embedded Systems}.
+	With this letter we are submitting an article titled \emph{On the Synchronization of Computational RFIDs}.
 	The objective of this article is to provide initial observations pertaining to 
@@
 
-	We believe that \emph{IEEE Transactions on Communications} is 
+	We believe that \emph{IEEE Transactions on Mobile Computing} is 
 	the best 

### 2017-11-01 manuscript_tmc.tex
TMC final

@@
 
-\title{ \color{blue} On the Synchronization of Computational RFIDs}
+\title{On the Synchronization of Computational RFIDs}
 
@@
 
-{\color{blue}{\subsection{The Need for CRFID Network Synchronization}}}
+\subsection{The Need for CRFID Network Synchronization}
 
@@
 We anticipate that these multi-hop networks will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.
-In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences \add{at the CRFID tag side}.
-\add{Therefore, establishing a common time notion among the tags, rather than synchronizing RFID readers only, will be a crucial service.}
-
-{\color{blue}{\subsection{Differences From Battery-Powered WSNs }}}
+In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences at the CRFID tag side.
+Therefore, establishing a common time notion among the tags, rather than synchronizing RFID readers only, will be a crucial service.
+
+\subsection{Differences From Battery-Powered WSNs }
 \label{sec:Motivation}
 
-\add{However, }the characteristics of CRFID systems expose fundamentally different challenges than \add{battery-powered} sensor networks to implement the synchronization service.
-The reason is mainly twofold:
+However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service. The reason is mainly twofold:
 
 \begin{itemize}
-	\item \textbf{Challenge I:} CRFID systems should perform computations \replace{in an energy efficient manner despite}{efficiently using marginal and} intermittent RF power that leads to frequent loss of computational state, e.g.
+	\item \textbf{Challenge I:} CRFID systems should perform computations efficiently using \emph{marginal} and intermittent RF power that leads to frequent loss of computational state, e.g.
 when the RFID reader moves away from the CRFID tag \cite{ransford:2008}.
-\add{Therefore, computations pertaining to time synchronization have an extremely short time budget, e.g.
+Therefore, computations pertaining to time synchronization have an extremely short time budget, e.g.
 should be completed with the other computation tasks within 10--100\,ms in WISP~\cite{Ransford:2011:Mementos}.
 On the contrary, the synchronization approaches in WSNs, e.g.
-the de facto protocol \cite{Maroti2004}, keep communication infrequent by pushing computation instead, so that the battery power is used efficiently in order to extend the life-time of the network up to several years.}
+the de facto protocol \cite{Maroti2004}, keep communication infrequent by pushing computation instead, so that the battery power is used efficiently in order to extend the life-time of the network up to several years.
 
@@
 varying oscillator frequencies in short-term affects the stability of the clocks and degrades the accuracy sensing.
-\add{Therefore, computations pertaining to time synchronization should be reactive against very-frequent voltage effects; e.g.
+Therefore, computations pertaining to time synchronization should be reactive against very-frequent voltage effects; e.g.
 during super-capacitor discharge.
@@
 Temperature compensation techniques, e.g.
-\cite{Schmid:2009}, are not applicable to compensate for the voltage instabilities in CRFIDs due to their computational complexity.}
+\cite{Schmid:2009}, are not applicable to compensate for the voltage instabilities in CRFIDs due to their computational complexity.
 
@@
 
-{\color{blue}{\subsection{Contributions}}}
-
-The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered \replace{wireless embedded}{CRFID} systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices.
+\subsection{Contributions}
+
+The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered CRFID systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices.
 In particular, the main contributions and findings of
@@
 
-\section{Synchronizing Battery-Free {\color{blue}CRFIDs}}
+\section{Synchronizing Battery-Free CRFIDs}
 \label{sec:Related-Work}
@@
 
-\subsubsection{{\color{blue}State-of-the-art Synchronization Techniques}}
+\subsubsection{State-of-the-art Synchronization Techniques}
 
@@
 In these approaches, nodes employ computation methods based on \emph{distributed consensus}~\cite{Sommer2009Gradient,SchenatoFiorentin:2011,Yildirim:External:2014}.
-\add{Since the prominent factor affecting the clock frequencies in WSNs is the temperature~\cite{schmid2010case}, there are also temperature compensated time synchronization techniques, e.g.
+Since the prominent factor affecting the clock frequencies in WSNs is the temperature~\cite{schmid2010case}, there are also temperature compensated time synchronization techniques, e.g.
 \cite{Schmid:2009}.
-However, such techniques require a calibration phase in order to build a temperature--clock frequency table.}
-
-\subsection{\color{blue}Fundamental Challenges of Synchronizing CRFIDs}
+However, such techniques require a calibration phase in order to build a temperature--clock frequency table.
+
+\subsection{Fundamental Challenges of Synchronizing CRFIDs}
 \label{sec:CRFIDs}
 
-We now delve into the synchronization characteristics of \remove{RF-powered wireless embedded systems.
-To this end, we will consider the} WISP, the de facto CRFID platform.
-The main aspects pertaining to synchronization in the WISP platform can be stated as follows:
+We now delve into the synchronization characteristics of WISP, the de facto CRFID platform. The main aspects pertaining to synchronization in the WISP platform can be stated as follows:
 
@@
 We plotted our measurements when the tag is powered by the Flash Emulation Tool (FET) with a stable DC power and when the tag is powered by the RFID Reader at a distance of 5~cm.
-\remove{Since harvested RF energy was different and unstable than the constant DC case, the frequency of the oscillator and in turn the toggle event was quite different.
-}We measured the mean and standard deviation of the frequency in the DC case as 1.1653 and 0.00028553, and these values for the RF case were 1.1793 and 0.00026049, respectively.
+We measured the mean and standard deviation of the frequency in the DC case as 1.1653 and 0.00028553, and these values for the RF case were 1.1793 and 0.00026049, respectively.
 }
@@
 On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift.
-Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level\remove{ rather than the temperature of the WISP tags}, see Fig.~\ref{fig:frequency_dc_rf}.
+Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.
 
@@
 Unfortunately, these issues lead to less accurate synchronization as compared to existing WSN solutions, as shown by our measurements presented in the following sections.
+Consequently, WSN synchronization protocols that can guarantee $\mu s$-level synchronization accuracy, e.g.~\cite{Maroti2004}, are hard to achieve in CRFID domain.
 
@@
 	\caption{\label{fig:least-squares}Synchronization error by employing least-squares regression on the collected timestamps.
-We observed a maximum synchronization error of 10 clock ticks between the RFID reader and the WISP tag, leading to \replace{3.2}{0.32}\,ms synchronization accuracy.}
+We observed a maximum synchronization error of 10 clock ticks between the RFID reader and the WISP tag, leading to 0.32\,ms synchronization accuracy.}
 \end{figure}
@@
 
-In this section, we introduce a simple but novel \emph{event-based} reader-tag synchronization inspired by simple PI controllers and provide its advantages over \replace{Approach 1}{the previous approach}.
+In this section, we introduce a simple but novel \emph{event-based} reader-tag synchronization inspired by simple PI controllers and provide its advantages over the previous approach.
 In event-based synchronization, a common event which is observable by all receiver devices simultaneously is generated by the reference device.
@@
 
-Considering these facts, we designed a lightweight and adaptive clock synchronization approach inspired by the PI-controller based solution introduced in~\cite{pi2015,yildirim2017adaptive}: \add{(i) we apply PI-controller based clock computation technique to the WISP platform due to its computational efficiency and fast adaptation; (ii) we introduce power-interrupt recovery mechanism so that the WISP tags do not loose their synchronization-related information.
-}
+Considering these facts, we designed a lightweight and adaptive clock synchronization approach inspired by the PI-controller based solution introduced in~\cite{pi2015,yildirim2017adaptive}: (i) we apply PI-controller based clock computation technique to the WISP platform due to its computational efficiency and fast adaptation; (ii) we introduce power-interrupt recovery mechanism so that the WISP tags do not loose their synchronization-related information.
+We modified the algorithm in our previous work so that PI-controllers use the periodic event occurrences rather than explicit clock values of a master node.
+We also provide a moderately different but more realistic theoretical analysis by using a more realistic clock drift model, i.e.
+Brownian motion rather than uniform distribution---see Appendix.
 
@@
 	\includegraphics[width=1.4\columnwidth]{figures/alpha-period-vs-variance.eps}
-	\caption{\label{fig:simulations} Numerical results showing the relations between \replace{communication}{event} period, integral gain and the steady-state variance.
+	\caption{\label{fig:simulations} Numerical results showing the relations between event period, integral gain and the steady-state variance.
 First, we set $\sigma_{\tau}=0.4097$, $\sigma_\eta = 0.0008$, $\beta=1$ and plotted $\sigma^2_{\gamma}$ using different $\tau$ (left hand side).

### 2017-11-01 appendix_tmc.tex
TMC submission

--- 2017-11-01 manuscript_tmc.tex
@@
-\documentclass[10pt,journal,compsoc]{IEEEtran}
-
-\ifCLASSOPTIONcompsoc
-  % IEEE Computer Society needs nocompress option
-  % requires cite.sty v4.0 or later (November 2003)
-  \usepackage[nocompress]{cite}
-\else
-  % normal IEEE
-  \usepackage{cite}
-\fi
-
-\usepackage{algorithm,setspace}
-\usepackage[noend]{algpseudocode} % needed for: algorithm pseudocode
-\usepackage{color}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{hyperref}
-\usepackage{stmaryrd}
-\usepackage{soul}
-\usepackage{cancel}
-\usepackage{listings}
-\usepackage{tikz} % needed for: tikz pics
-\usetikzlibrary{arrows}
-\usetikzlibrary{decorations.pathreplacing}
-
-\definecolor{aliceblue}{rgb}{0.94, 0.97, 1.0}
-\definecolor{applegreen}{rgb}{0.55, 0.71, 0.0}
-\definecolor{amber}{rgb}{1.0, 0.75, 0.0}
-
-\lstdefinestyle{mystyle}{
-	tabsize          = 1,     % tab space width
-	numbers          = left,  % display line numbers on the left
-	%stepnumber		 = 2,
-	numberfirstline=false,
-	%framesep         = 1pt,   % expand outward
-	framerule        = 0.2pt, % expand outward 
-	commentstyle     = \color{applegreen},      % comment color
-	keywordstyle     = \color{blue},       % keyword color
-	stringstyle      = \color{amber},    % string color
-	backgroundcolor  = \color{aliceblue}, % backgroundcolor color
-	showstringspaces = false,              % do not mark spaces in strings
-	otherkeywords	 ={signal,inos_task,uint16_t},
-	language = C,
-	%numbers = none, 
-	escapechar = !,
-	basicstyle = \ttfamily\bfseries\scriptsize, 
-	xleftmargin=0.5ex,
-	linewidth = .75\linewidth,
-	basicstyle=\tiny,
-}
-
-\newcommand{\expected}[1]{E\!\left\llbracket #1 \right\rrbracket}
-\newcommand{\revised}[1]{\textcolor{blue}{#1}}
-\newcommand{\TODO}[1]{\textcolor{blue}{\textbf{TODO: }}\textcolor{red}{#1}}
-\newtheorem{fact}{Fact}
-
-% text revision commands
-\newcommand{\add}[1]{\textcolor{blue}{#1}}
-\newcommand{\remove}[1]{\textcolor{red}{\st{#1}}}
-\newcommand{\replace}[2]{\remove{#1} \add{#2}}
-
-\begin{document}
-
-\title{On the Synchronization of Computational RFIDs}
-
-\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, 
-	Henko~Aantjes, 
-	Przemys{\l}a~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
-	and~Amjad~Yousef~Majid
-	%
-	\IEEEcompsocitemizethanks{
-		\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}. \protect\\
-		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).
-	}
-%	\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}
-}
-
-\IEEEtitleabstractindextext{%
-\begin{abstract}
-Battery-free computational RFID platforms, such as WISP (Wireless Identification and Sensing Platform), are intermittently-powered devices designed for replacing existing sensor networks.
-Accordingly, synchronization appears as one of the  crucial building blocks for collaborative and coordinated actions in these platforms.
-However, intermittent power leads to frequent loss of computational state and short-term clock frequency instability that makes synchronization challenging.
-In this article, we introduce WISP-Sync protocol that provides synchronization among WISP tags in the communication range of an RFID reader.
-WISP-Sync overcomes the aforementioned challenges by employing a Proportional-Integral (PI) controller-inspired algorithm which (i) is adaptive---reactive to short-term clock instabilities; (ii) requires only a few computation steps---suitable for limited harvested energy; and (iii) keeps a few variables to hold the synchronization state---minimum overhead to recover from power interrupts.	Evaluations in our testbed showed that WISP-Sync ensured an average synchronization error of approximately 1\,ms among the tags with an average energy overhead of 1.85\,$\mu$J per synchronization round.
-\end{abstract}
-
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Time Synchronization, Computational RFIDs, Wireless Identification and Sensing Platform (WISP).
-\end{IEEEkeywords}}
-
-\maketitle
-
-\IEEEdisplaynontitleabstractindextext
-
-\IEEEpeerreviewmaketitle
-
-\IEEEraisesectionheading{\section{Introduction}\label{sec:Intro}}
-
-\IEEEPARstart{L}{ow-power} wireless embedded systems consist of tiny, low-cost and spatially separated computers that communicate with each other by exchanging radio packets.
-Powering these small-scale embedded systems, e.g.
-wireless sensor networks (WSNs), is still a crucial problem~\cite{emergence}.
-Replacing or recharging their batteries is impractical and inhibits long-term operation.
-Moreover, batteries increase the size and cost of the hardware.
-Fortunately, the energy efficiency of these systems has improved considerably such that their power requirements are in the order of a few $\mu$W \cite{booksmith2013}.
-Furthermore, recent advancements in microelectronics technology enabled harvesting power from radio frequency (RF) sources that is sufficient to operate low-power embedded systems in practice~\cite{60ghz,emergence,bi:commag:2015,xie_wcom_2013,garnica_procieee_2013}.
-Nowadays, the growth of the RF-powered computing paradigm is bringing new research opportunities and challenges~\cite{emergence}, leading to a promising class of low-power embedded systems, the so-called Intermittently Powered Devices (IPDs).
-
-By taking existing RFID (Radio Frequency IDentification) technology as a foundation, computational RFIDs (CRFIDs) are emerging IPDs that allow sensing, computation and communication without batteries---replacing existing battery-powered sensor networks~\cite{booksmith2013}.
-CRFIDs are equipped with a backscatter radio~\cite[Chapter 4]{rfid_handbook} composed of a simple circuitry that modulates the carrier wave generated by a reader to transmit information.
-This allows communication to come almost for free.
-This is a fundamental difference from sensor networks where the transceiver circuit is the most energy-hungry component.
-In the CRFID domain, the bottleneck in terms of power consumption has shifted from communication to computation and sensing~\cite{ekhonet}.
-A typical example of CRFIDs is the WISP (Wireless Identification and Sensing Platform)~\cite{wisp}.
-Commercial RFID readers implementing the EPC Gen2 standard~\cite{epc_gen2} are used to power WISP tags.
-Apart from low data rate sensing~\cite{neuralWISP}, these maintenance-free devices are evolving to support high data rate and more complex sensing applications such as continuous sensor-data streaming, e.g.
-capture and transfer of images via battery-free cameras, i.e.
-WISPCam~\cite{wispcam_2015,cameralocalize_2015}.
-
-\subsection{The Need for CRFID Network Synchronization}
-
-As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}.
-However, recent advancements in the wireless communication enable tag-to-tag communication, e.g.~\cite{liu2013ambient}.
-We anticipate that these multi-hop networks will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.
-In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences at the CRFID tag side.
-Therefore, establishing a common time notion among the tags, rather than synchronizing RFID readers only, will be a crucial service.
-
-\subsection{Differences From Battery-Powered WSNs }
-\label{sec:Motivation}
-
-However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service. The reason is mainly twofold:
-
-\begin{itemize}
-	\item \textbf{Challenge I:} CRFID systems should perform computations efficiently using \emph{marginal} and intermittent RF power that leads to frequent loss of computational state, e.g.
-when the RFID reader moves away from the CRFID tag \cite{ransford:2008}.
-Therefore, computations pertaining to time synchronization have an extremely short time budget, e.g.
-should be completed with the other computation tasks within 10--100\,ms in WISP~\cite{Ransford:2011:Mementos}.
-On the contrary, the synchronization approaches in WSNs, e.g.
-the de facto protocol \cite{Maroti2004}, keep communication infrequent by pushing computation instead, so that the battery power is used efficiently in order to extend the life-time of the network up to several years.
-
-	\item \textbf{Challenge II:} The continuously varying voltage supply introduces severe hardware instability, e.g.
-varying oscillator frequencies in short-term affects the stability of the clocks and degrades the accuracy sensing.
-Therefore, computations pertaining to time synchronization should be reactive against very-frequent voltage effects; e.g.
-during super-capacitor discharge.
-On the contrary, in WSNs the input voltage is constant and the temperature is the main factor affecting the clock frequencies~\cite{schmid2010interaction}.
-Temperature compensation techniques, e.g.
-\cite{Schmid:2009}, are not applicable to compensate for the voltage instabilities in CRFIDs due to their computational complexity.
-
-\end{itemize}
-
-\subsection{Contributions}
-
-The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered CRFID systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices.
-In particular, the main contributions and findings of
-this article are:
-%
-\begin{itemize}
-	\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems.
-Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides synchronization of individual tags in the communication range of an RFID reader.
-	\item By addressing the aforementioned challenges, we design and implement \emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015,yildirim2017adaptive}.
-WISP-Sync (i)
-	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
-	%
-	\item We provide theoretical analysis to prove that this primitive establishes synchronization and to reveal the factors affecting its synchronization performance.
-	%
-	\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
-\end{itemize}
-%
-We note that WISP-side implementations throughout the article are done using C and assembly language.
-All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\href{https://goo.gl/RuWIZ2}{https://goo.gl/RuWIZ2}.
-
-The remainder of this article is organized as follows.
-In Section~\ref{sec:Related-Work} we present the fundamental challenges of synchronizing IPDs.
-Section~\ref{sec:Clock_Hardware} presents the hardware, in particular clock system of the WISP platform briefly, while Section~\ref{sec:clock_model} provides the mathematical notation to be used throughout the article.
-In
-Section~\ref{sec:sender_receiver} we propose a sender-receiver synchronization approach between an RFID reader and a single tag.
-Sections~\ref{sec:event_based} and~\ref{sec:eval-wispsync} present the WISP-Sync protocol and its evaluation in our testbed, respectively.
-We also present the synchronization of multiple tags in Section~\ref{sec:multi-wisp}.
-Finally, we provide our conclusions and future research directions in Section~\ref{sec:Conclusion}.
-
-\section{Synchronizing Battery-Free CRFIDs}
-\label{sec:Related-Work}
-
-In this section, we present a brief  state of the art on synchronization issues in conventional WSNs.
-Then, we emphasize the main challenges of synchronizing wireless battery-free embedded systems by focusing on a particular IPD platform: computational RFIDs.
-
-\subsection{Synchronization in Conventional WSNs}
-\label{sec:WSN-Sync}
-
-The instability of the clock hardware, delays during communication among sensor nodes, and software methods to establish synchronization are the main factors affecting the synchronization in conventional WSNs.
-
-\subsubsection{Clock Hardware}
-
-In WSNs, each sensor node is equipped with a built-in clock that is implemented as a counter register clocked by a low-cost external crystal oscillator.
-At each oscillator pulse, i.e.
-\emph{tick} of the clock, the counter register is incremented.
-The duration between two consecutive ticks is the \emph{rate} of the built-in clock.
-Environmental factors such as temperature, supply voltage and aging of the crystal prevent built-in clocks to generate ticks at the exact speed of real-time, leading to bounded \emph{clock drift}.
-The prominent environmental factor affecting the frequency of the built-in clocks is the temperature~\cite{Schmid:2009,Schmid:2010}.
-Moreover, \emph{quantization errors} occur with low-frequency built-in clocks, which prevents precise timing measurements.
-
-\subsubsection{Wireless Communication}
-
-Sensor nodes exchange their clock information periodically to compute a \emph{software clock} that represents the synchronized notion of time.
-A software clock is composed of an offset and a frequency that hold the value and speed difference between the corresponding built-in clock and the reference time, respectively.
-The difference between the reference time and the software clock is the \emph{synchronization error}.
-In WSNs, the synchronization error is affected by several sources of errors.
-The \emph{transmission delay}, defined as the time that passes between the start of the broadcast and the receipt by the receiver node, is the major error source and composed of deterministic and non-deterministic components~\cite{Maroti2004}.
-Assigning timestamps at the MAC layer removes the deterministic delay components and improves synchronization accuracy.
-This obligates the use of transceivers that allow assignment of time information to a radio packet just before transmission and reception, e.g.
-Chipcon CC2420~\cite{chipcon:cc2420}.
-Since the transceiver is the most power-hungry circuit, nodes in WSNs should reduce re-synchronization frequency to decrease communication overhead and save power.
-
-\subsubsection{State-of-the-art Synchronization Techniques}
-
-The most common synchronization mechanism is to propagate the time information of a particular \emph{reference node} to let receiver nodes synchronize themselves to the received reference time information.
-Commonly, \emph{least-squares regression} is employed by several practical synchronization protocols to adjust the offset and the frequency of the software clocks ~\cite{Maroti2004,Lenzen:2015,TATS:2016}.
-On the other hand \emph{iterative} computation methods, e.g.
-in ~\cite{Yildirim:AVTS:2014,pi2015,yildirim2017adaptive,Yildirim:Gradient:2016}, perform identical computation steps at each re-synchronization round.
-With iterative methods, sensor nodes establish synchronization after a finite number of rounds in a lightweight manner in terms of computation and memory requirements.
-There are also \emph{fully distributed} approaches in which sensor nodes interact only with and synchronize to their direct neighbors in a peer-to-peer fashion without the requirement of a dedicated reference node.
-In these approaches, nodes employ computation methods based on \emph{distributed consensus}~\cite{Sommer2009Gradient,SchenatoFiorentin:2011,Yildirim:External:2014}.
-Since the prominent factor affecting the clock frequencies in WSNs is the temperature~\cite{schmid2010case}, there are also temperature compensated time synchronization techniques, e.g.
-\cite{Schmid:2009}.
-However, such techniques require a calibration phase in order to build a temperature--clock frequency table.
-
-\subsection{Fundamental Challenges of Synchronizing CRFIDs}
-\label{sec:CRFIDs}
-
-We now delve into the synchronization characteristics of WISP, the de facto CRFID platform. The main aspects pertaining to synchronization in the WISP platform can be stated as follows:
-
-\subsubsection{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication.
-Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}.
-Since WISP tags are unable to communicate with their neighboring nodes directly using backscatter communication as of now, \emph{tag-tag synchronization} is impossible.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.85\columnwidth]{figures/clock_frequency_dc_rf.eps}
-	\caption{\label{fig:frequency_dc_rf} In our testbed, see Section~\ref{sec:testbed} for the details, we programmed a single WISP tag so that it toggles one of its output port whenever its 32~KHz built-in timer progresses 32000 clock ticks.
-In the ideal case where the crystal oscillator of the tag does not drift, the tag would toggle its output port at each 1 second, i.e.
-with a frequency of 1\,Hz.
-During our experiment under a stable environmental temperature, we connected the output port of the tag to a logic analyzer and measured the frequency of the toggle event.
-We plotted our measurements when the tag is powered by the Flash Emulation Tool (FET) with a stable DC power and when the tag is powered by the RFID Reader at a distance of 5~cm.
-We measured the mean and standard deviation of the frequency in the DC case as 1.1653 and 0.00028553, and these values for the RF case were 1.1793 and 0.00026049, respectively.
-}
-\end{figure}
-
-\subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term.
-On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift.
-Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.
-
-\subsubsection{Frequent loss of synchronization state} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g.
-the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up.
-However, saving computational state is also an energy consuming task~\cite{blisp}.
-
-\subsubsection{Computation and memory overhead sensitivity} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p.
-44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}.
-Due to the intermittent power, lightweight methods in terms of computation and memory are required for the synchronization.
-Since methods like least-squares regression are computationally heavy and require considerable amount of memory~\cite{pi2015} they should be avoided.
-
-\subsubsection{Limitations of the EPC Gen 2 standard} The WISP firmware implements the EPC Gen 2 standard~\cite{epc_gen2} which increases the compatibility with the existing RFID systems.
-However, the standard introduces limitations, e.g.
-currently it does not assign timestamps to the radio packets, which is a fundamental requirement to establish synchronization.
-Moreover, communication delays between the reader and tag are quite dependent on the implementation of this standard by RFID readers.
-Unfortunately, these issues lead to less accurate synchronization as compared to existing WSN solutions, as shown by our measurements presented in the following sections.
-Consequently, WSN synchronization protocols that can guarantee $\mu s$-level synchronization accuracy, e.g.~\cite{Maroti2004}, are hard to achieve in CRFID domain.
-
-\section{WISP Hardware and CRFID Networks: A Brief Look}
-\label{sec:Clock_Hardware}
-
-In this section we present brief information about the WISP hardware, in particular its clock system. For the sake of clarity, we 
-also give details about a typical CRFID network, which we used to implement and evaluate our synchronization methods in this article.
-
-\subsection{Low Power Operation Modes and The Clock System}
-\label{subsec:clock_system}
-
-The WISP 5.0 platform comes with the MSP430FR5969 \cite{msp430fr5969_data_sheet} microcontroller (later called as MSP430 in this article) with FRAM non-volatile memory.
-The MSP430 clock system includes (i) a 32\,kHz external crystal oscillator, (ii) an internal very-low-power low-frequency oscillator, and (iii) an integrated internal digitally controlled oscillator (DCO).
-The five built-in 16-bit timers (TA0--TA3, TB0) in this system can be clocked with the auxiliary clock (ACLK) signal that is sourced from the relatively stable external 32\,kHz oscillator.
-The MSP430 has one active mode and seven software selectable low-power operation modes.
-Low Power operation Mode 3 (LPM3) is the standby mode where CPU, FRAM and high-frequency peripherals are off; the external 32\,kHz oscillator and the ACKL signal are active.
-In our target implementations, we configured TB0 to be clocked  with ACKL so that it has 16-bit 32\,kHz precision and runs continuously in LPM3 mode.
-
-\subsection{\label{sec:testbed}An Experimental CRFID System}
-
-\begin{figure}
-	\centering
-	\begin{tikzpicture}
-	\node[anchor=south west, inner sep=0] at 
-	(0,0){\includegraphics[width=\columnwidth]{figures/testbed_resized.jpg}};
-
-	\node [fill= white,circle] at (7,2) { };
-	\node [color=black ] at (7,2) { 1 };
-
-	\node [fill= white,circle] at (8,4) {  };
-	\node [color=black ] at (8,4) { 2 };
-
-	\node [fill= white,circle] at (4,2) {  };
-	\node [color=black ] at (4,2) { 3 };
-
-	\node [fill= white,circle] at (5.1,2) {  };
-	\node [color=black ] at (5.1,2) { 4 };
-
-	\node [fill= white,circle] at (6.5,4.5) {  };
-	\node [color=black ] at (6.5,4.5) { 5 };
-
-	\end{tikzpicture}
-	\caption{\label{fig:crfid_testbed} A single-hop CRFID testbed: (1) an Impinj Speedway R1000 RFID reader, (2) a single WISP 5.1 tag, (3) a host computer, (4) FET programmer and (5) a Laird S9028PCR antenna.
-The same testbed setup has been used previously in~\cite{wisent:2016} by us to evaluate downstream data transmission in CRFIDs.}
-\end{figure}
-
-As presented in Fig.~\ref{fig:crfid_testbed}, a typical CRFID system is composed of an RFID reader, a single WISP tag placed inside the communication range of this reader and a host computer to control this reader.
-In our testbed, we used a 915\,MHz Impinj Speedway R1000 RFID reader with firmware version 3.2.4 connected to a Laird S9028PCR 8.5\,dBic gain antenna.
-We placed the WISP tag at the line-of-sight from the reader antenna.
-For the host-reader control operations, we used sllurp~\cite{sllrp_github}, a LLRP (Low-Level Reader Protocol)~\cite{llrp}  control library written in Python.
-To program the WISP tag, we used MSP430 Flash Emulation Tool (FET), in combination with TI Code Composer Studio (CCS), attached to the host.
-All of our experimental evaluation presented in the next sections were conducted using this setup at a university office with human presence.
-
-\subsection{\label{sec:sniffer}A Sniffer for the CRFID System}
-
-Communication delays are one of the main sources of errors in synchronization~\cite{Maroti2004} and they should be kept as small as possible.
-In order to characterize the communication delays between the RFID reader and the WISP tags during backscatter communication and understand the effect of these delays on the synchronization error, we used USRP 210 software-defined radio, another Laird antenna placed at 50\,cm from the tag and the GNU Radio toolkit~\cite{gnuradio} with the custom software to sniff the radio packets.
-The sniffer was presented first in~\cite{stork:2017} and we refer the reader to this study for more details.
-Implementing sniffing mechanism allowed us to store the timings of the transmission and reception of the EPC Gen 2 packets, which we later processed offline to measure the actual transmission delays.
-We will present these measurements in the next sections.
-\section{System Model, Notation and Preliminaries}
-\label{sec:clock_model}
-
-In this section, we provide a mathematical framework that models the built-in clocks and in particular the clock drifts of the WISP tags and the RFID reader.
-We will use these models to prove the correctness of our synchronization algorithms and to derive mathematical expressions that represent their synchronization performance.
-\begin{figure*}
-	\centering
-	\includegraphics[width=1.4\columnwidth] {figures/relative_clock_frequency.eps}
-	\caption{\label{fig:frequency}The frequencies of the toggle events (obtained using the same steps presented in Fig.\ref{fig:frequency_dc_rf}) of two RF-powered WISP tags, placed at 5 cm distance from the RFID reader.
-On the left the individual event frequencies ($f_1$ and $f_2$) are presented, and their relative clock frequency ($\frac{f_1}{f_2}-1$) is presented on the right.
-We measured the mean and standard deviation of $f_1$ as 1.1793 and 0.00026049 and those of $f_2$ as 1.1705 and  0.00068239, respectively.}
-\end{figure*}
-In this framework, we denote a single WISP tag by $w$ and the RFID reader by $r$.
-The Timer B0 mentioned in Section~\ref{subsec:clock_system} can be considered as the built-in clock, i.e.
-\emph{local clock}, of the tag $w$,
-which we denote by $C_{w}$. The value of $C_{w}$ at time $t$ can be modeled as
-%
-\begin{eqnarray}
-C_{w}(t) & \triangleq & C_w(t_0)+ \int_{t_0}^{t}f_{w}(u)du, 
-\label{eq:local_clock}
-\end{eqnarray}
-%
-where $t_0$ represents the time at which tag $w$ is powered on and  $f_{w}(u)$ represents the \emph{instantaneous oscillator frequency} of $C_{w}$ at time $u$.
-Since the frequency of the oscillators are affected by environmental factors such as temperature and voltage level, they are not stable~\cite{Lenzen2009Optimal} and $f_{w}$ is not a constant but it is \emph{time varying}, see Fig.~\ref{fig:frequency}.
-From \eqref{eq:local_clock}, it can be noticed that $C_w$ is \emph{unitless}.
-Indeed, it can be seen as a counter of oscillation events that are produced at a frequency of $f_w$.
-
-The RFID reader $r$ can be considered as a natural time reference for the tag $w$.
-We denote the local clock of the RFID reader at time $t$ by $C_{r}(t)$ and its frequency by $f_r(t)$.
-By collecting (time) information from the RFID reader, the tag $w$ runs a \emph{clock synchronization algorithm} to calculate a \emph{software clock}, denoted by $S_{w}(t)$, whose value represents the synchronized notion of time.
-The objective of the synchronization algorithm is minimizing the \emph{synchronization error} with respect to the reader $r$ at any time $t$, which is defined as
-%
-\begin{eqnarray}
-\gamma(t)&\triangleq&S_w(t)-C_r(t). \label{eq:gamma_error}
-\end{eqnarray}
-%
-Considering the individual clock frequencies $f_w$ and $f_r$, we denote the \emph{instantaneous relative clock 
-	frequency} of the WISP tag 
-with respect to the RFID reader at time $t$ as
-%
-\begin{eqnarray}
-f_w^r(t)&\triangleq&\frac{f_w(t)}{f_r(t)}-1. \label{eq:relative_frequency}	
-\end{eqnarray}
-\section{Sender-Receiver Based Reader-Tag Synchronization in CRFID Systems}
-\label{sec:sender_receiver}
-
-In this section we provide initial observations, design and implementation of a \emph{sender-receiver} based synchronization approach that synchronizes a single WISP tag to the RFID reader and we reveal its limitations in current CRFID networks.
-In sender-receiver based synchronization mechanisms, receiver devices synchronize to the clock of a reference sender device.
-In order to synchronize itself to the RFID reader with such a mechanism, the WISP tag should obtain several $(C_w(t),C_r(t))$ synchronization points to establish a \emph{relationship} between its local clock $C_w$ and the reader clock $C_r $, represented by its software clock $S_w$.
-The value $S_w(t)$ will provide an estimate of the reference clock $C_r(t)$ at any time instant $t$.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/Sender-Receiver.pdf}
-	\caption{\label{fig:sender_receiver} The message exchange among the host computer, the RFID reader and the tag for sender-receiver synchronization.
-The host machine, i.e.
-a PC,  sends the high level commands to the reader via LLRP \emph{AccessSpec} message and receives the results through an \emph{ASReport}.
-The RFID reader follows the steps defined in the EPC Gen2 standard: (i) performs a \emph{Handshake} to initialize the communication with the active tag; (ii) performs a \emph{Command Initialization} (Cmd Init) by requesting a random number from the tag (Req\_n) and receiving the random number (newRN); (iii)  performs a \emph{Read} command by sending its request and receiving the timer value.
-The reader assigns the \emph{FirstSeenTimestamp} to the tag at time $t_1$ and the tag timestamps the command reception event at time $t_2$ with its local clock reading.}
-\end{figure}
-
-We explored the EPC Gen2 standard and the LLRP protocol and found out that LLRP assigns a \emph{FirstSeenTimestamp} in UTC (Coordinated Universal Time), which is defined as \emph{``The Reader SHALL set it to the time of the first observation amongst the tag reports that get accumulated in the TagReportData''} \cite[p.
-87]{llrp}.
-From this definition, we assume that this timestamp is assigned by the reader when it receives the EPC during the handshake operation with the corresponding tag, shown in Fig.
-\ref{fig:sender_receiver} as the timestamp assigned at time $t_1$.
-Therefore, the \emph{FirstSeenTimestamp} can be considered as $C_r(t_1)$.
-In order to obtain the corresponding local time $C_w(t_1)$, one strategy is to force the reader to send a special ``synchronization'' command after the handshake so that the tag timestamps the command reception event using its local clock, shown in Fig.\,\ref{fig:sender_receiver} as time $t_2$.
-As we emphasized in Section~\ref{sec:WSN-Sync}, transmission delays are the main source of errors in time synchronization and they should be kept as deterministic as possible.
-We denote the \emph{transmission delay} in this case by $\Delta t= t_2-t_1$ and it is desirable to keep the \emph{variation} of $\Delta t$, i.e.
-\emph{jitter}, as small as possible~\cite{Maroti2004}.
-
-\subsection{Transmission Delay and Jitter Measurements}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.9\columnwidth]{figures/sender_receiver_uncertainty.eps}
-	\caption{\label{fig:transmission_delay} The normalized number of occurrences of the transmission delays measured by sniffing the communication between the RFID reader and the WISP tag.
-We calculated the mean transmission delay and its standard deviation as 1.89\,ms and 0.0164\,ms with a 99\% confidence interval of [1.8874,1.8925] and [0.0148,0.0184], respectively.}
-\end{figure}
-
-In order to characterize the transmission delay $\Delta t$, we placed the WISP tag 20 cm away from the antenna using the setup presented in Fig.~\ref{fig:crfid_testbed} and we sniffed the communication between the RFID reader and the WISP tag during the communication scenario of Fig.~\ref{fig:sender_receiver}.
-Fig.~\ref{fig:transmission_delay} presents a summary of our measurements (with 300 samples gathered).
-We observed that the transmission delay is distributed with a mean of 1.89\,ms and standard deviation of 0.0164\,ms.
-Moreover, we observed some outliers, presented at the right hand side of the figure, which we attribute to the EPC Gen2 implementation of the Impinj reader.
-
-\subsection{Collection of Clock Values for Off-line Processing}
-
-After characterizing the variation of the transmission delay, the next step was the collection of the clock values from the RFID Reader and the WISP tag in order to try synchronization mechanisms offline, which allowed to evaluate the performance of sender-receiver synchronization mechanism in CRFID networks.
-We collected $(C_w(t_2),C_r(t_1))$ pairs for offline processing by controlling the RFID reader to send a \emph{Read} command to the tag and by programming the WISP tag so that it backscatters $C_w(t_2)$ upon receiving this command.
-The received pairs $(C_w(t_2),C_r(t_1))$ were logged by the host computer.
-
-\subsection{Software Clock Computation Using  Least-Squares}
-
-Following studies~\cite{Maroti2004,Schmid:2010,Lenzen:2015,TATS:2016} that use a linear clock model, we assume a \emph{linear} relationship between $C_r$ and $C_w$ and we model the software clock of the WISP tag as $S_w(C_w(t)) = a + b C_w(t)$ where $S_w$ represents an estimator of the reader's clock $C_r$, $a$ is the \emph{offset} and $b$ is the \emph{relative speed} with respect to the local clock $C_w$.
-To establish such a relationship, we performed \emph{least-squares regression} offline in MATLAB using the collected pairs in the previous subsection.
-As in \cite{Maroti2004}, since the WISP tag has limited memory, computation capability and energy, at each step only the most recent $N$ pairs are used to estimate the regression line.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.9\columnwidth]{figures/least_squares.eps}
-	\caption{\label{fig:least-squares}Synchronization error by employing least-squares regression on the collected timestamps.
-We observed a maximum synchronization error of 10 clock ticks between the RFID reader and the WISP tag, leading to 0.32\,ms synchronization accuracy.}
-\end{figure}
-
-Formally, let $[C_w(t_k),C_r(t_k)]$ denote the $k$th pair in the log file where $t_k$ denotes the real-time at which 
-\emph{FirstSeenTimestamp} has been assigned during the collection of this pair.
-At each $k$th step, the pairs $\{[C_w(t_k),C_r(t_k)],$\ldots$,[(C_w(t_{k+N-1}),C_r(t_{k+N-1})]\}$ are used to calculate the parameters $a$ and $b$, i.e.
-the \emph{intercept} and the \emph{slope} of the estimated regression line \cite{book-SheldonRoss}, using \textsf{polyfit} function of MATLAB.
-We calculated the synchronization error as
-$\gamma(t_{k+N})\triangleq S_w(C_w(t_{k+N}))-C_r(t_{k+N})$
-%
-that represents the difference between the predicted reference time and the received reference time.
-In our implementation we used $N=8$ as in~\cite{Maroti2004} and Fig.~\ref{fig:least-squares} presents the synchronization error at each step.
-We observed a maximum synchronization error of 0.32\,ms in this one-hop network, which is more than one order of magnitude larger than the synchronization performance of the de facto WSN solution \cite{Maroti2004}, which was reported as approximately 10\,$\mu$s.
-
-\subsection{Limitations of the Approach}
-
-In addition to the challenges listed in Section \ref{sec:CRFIDs}, we observed two crucial limitations for the WISP platform, which prevents to build up a sender-receiver synchronization building block:
-
-\subsubsection{Host computer computation} We are unaware of any EPC Gen2 command that will allow to send the \emph{FistSeenTimestamp} to the tag.
-Hence, even though we were able to collect $(C_w(t),C_r(t))$ pairs for offline processing, it is not possible for the tag to collect $C_r(t)$ and synchronize itself to the reader.
-Therefore, with this limitation, only a host computer can collect and log the timestamps, calculate the relationship between the clock of the tag and the clock of the reader and send the data that represents this relationship to the tag for synchronization, making this method only of theoretical nature as of now.
-
-\subsubsection{The Lack of broadcast primitive} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader.
-Unfortunately, in the current EPC Gen2 and LLRP standard, we were unable to find any mechanism to send a global timestamp that is received by all tags inside the range of the RFID reader to establish synchronization in one step.
-
-\subsubsection{Big and reader dependent transmission delays} As depicted previously, we observed that transmission delays in our testbed were on the order of milliseconds, see Fig.~\ref{fig:least-squares}, whereas they are reported as a few microseconds in conventional WSN platforms.
-This is one of the crucial limitations since the smaller the transmission delays are, the better the synchronization.
-
-\section{Event-Based Reader-Tag Synchronization In CRFID Systems}\label{sec:event_based}
-
-In this section, we introduce a simple but novel \emph{event-based} reader-tag synchronization inspired by simple PI controllers and provide its advantages over the previous approach.
-In event-based synchronization, a common event which is observable by all receiver devices simultaneously is generated by the reference device.
-Upon receiving events generated at regular intervals, receiver devices can predict the occurrence time of future events.
-In order to synchronize the tag with such a mechanism, the RFID reader does not send explicit timestamp values as in sender-receiver based synchronization but instead generates events at regular intervals.
-Upon observing these events, the receiver tag adjusts the rate of its software clock so that it predicts the occurrence of the next event precisely.
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/Event-Based.pdf}
-	\caption{\label{fig:event_based}Event-Based synchronization steps: The tag timestamps successive \emph{BlockWrite} events and adjusts its software clock.}
-\end{figure}
-We explored the EPC Gen2 standard to see how to generate events at regular intervals and realized that the \emph{BlockWrite} operation allows this feature.\footnote{It is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0, i.e.
-Miller 1, modulation) in order to get the timing measurements of the regular patterns in the signal.
-However, this requires using a high-precision clock to obtain microsecond-precision timing measurements.
-In this study, we used the higher-level properties of the EPC Gen2 standard, rather than exploiting its low-level signal properties.
-By means of this, we could  use the external 32\,KHz clock that runs even in low-power standby mode (see Section \ref{sec:Clock_Hardware}) to obtain timing measurements.} Fig.~\ref{fig:event_based} presents the steps of event-based synchronization.
-During the command phase, the EPC Gen2 standard allows a maximum of eight successive \emph{BlockWrite} operations.
-It is desirable to use the first and the last \emph{BlockWrite} events for synchronization since it is better to compensate frequency differences observed in longer time intervals to adjust the software clock.
-
-\subsection{Event Period Measurements}
-\label{subsec:event_period_uncertainty}
-
-\begin{figure}
-	\centering
-	\includegraphics[width = 0.9\columnwidth]{figures/blockwrite_delay.eps}
-	\caption{\label{fig:block_write_delay}The delay between the first and the last \emph{BlockWrite} event, i.e.
-\emph{event period}, by considering 20 samples during the communication scenario in Fig.~\ref{fig:event_based}.
-We measured its mean and standard deviation as 226.7667\,ms and 0.4097\,ms with a 99\% confidence interval of [226.4961,227.0372] and [0.2852,0.6945], respectively.}
-\end{figure}
-
-The real-time length between the first and the last \emph{BlockWrite} operation is the \emph{event period}, we denote by $\tau$, and its variation is the main error source.
-Therefore, it is important to explore its characteristics.
-We sniffed the communication protocol between the RFID reader and a tag presented in Fig.
-\ref{fig:event_based}.
-We took 20 sample measurements about the \emph{event period}, which is summarized in Fig.~\ref{fig:block_write_delay}.
-According to our measurements, the \emph{event period} was distributed with a mean of 226.76\,ms and standard deviation of 0.41\,ms; respectively.
-We would like to mention that these values are dependent on the RFID Reader implementation of the EPC Gen2 protocol.
-
-\subsection{Event-Based Synchronization Inspired By PI Controllers}
-
-For the computation of the software clock, \emph{adaptive} and \emph{lightweight} solutions are crucial due to the following important requirements:
-
-\begin{itemize}
-	\item Since the voltage level is not stable and the harvested power is limited, the computations should demand little amount of energy to calculate the software clock, i.e.
-marginal number of steps---\emph{Addressing Challenge I in Section~\ref{sec:Motivation}}.
-
-	\item Since power is intermittent, the number of variables pertaining to the software clock should be marginal so that saving the state of the synchronization to non-volatile memory will demand little time and less energy.
-Therefore, lightweight computation methods in terms of main memory overhead are required---\emph{Addressing Challenge I in Section~\ref{sec:Motivation}}.
-
-	\item Unstable voltage level leads to frequently varying clock frequencies.
-Therefore, the computation methods should adapt such dynamic conditions in a very fast manner to keep synchronization accuracy stable---\emph{Addressing Challenge II in Section~\ref{sec:Motivation}}
-\end{itemize}
-
-Considering these facts, we designed a lightweight and adaptive clock synchronization approach inspired by the PI-controller based solution introduced in~\cite{pi2015,yildirim2017adaptive}: (i) we apply PI-controller based clock computation technique to the WISP platform due to its computational efficiency and fast adaptation; (ii) we introduce power-interrupt recovery mechanism so that the WISP tags do not loose their synchronization-related information.
-We modified the algorithm in our previous work so that PI-controllers use the periodic event occurrences rather than explicit clock values of a master node.
-We also provide a moderately different but more realistic theoretical analysis by using a more realistic clock drift model, i.e.
-Brownian motion rather than uniform distribution---see Appendix.
-
-Let $t_0$ denote the first event reception time at the tag. Using the pre-measured event period $\tau$ of the RFID reader, the tag can estimate the next event reception time at time $t_1$ as
-\begin{eqnarray}
-\hat{C}_w(t_1) &=& C_w(t_0)+\tau(1+\hat{f}_w^r(t_0)), \label{eq:core}
-\end{eqnarray}
-where $\hat{f}_w^r(t_0)$ denotes the \emph{estimated relative frequency} at time $t_0$, which is an estimate for \eqref{eq:relative_frequency}.
-The intuition behind \eqref{eq:core} is that $1+\hat{f}_w^r(t_0)$ gives the estimated value of $f_r(t_0)/f_w(t_0)$, $\tau f_r(t_0)$ gives the estimated local time passed on the reader side and $\tau f_r(t_0)/f_w(t_0)$ gives the estimated number of clock ticks on the tag side that will pass until the next packet reception.
-Redefining \eqref{eq:gamma_error}, the \emph{prediction error} at time $t_1$ can be written as
-\begin{eqnarray}
-\gamma(t_1) & \!\!\!\!\!=\!\!\!\!\!\! & 
-C_{w}(t_{1}\!)\!-\!\hat{C}_{w}(t_{1}\!) \!=\! 
-C_{w}(t_{1}\!)\!-\!C_{w}(t_{0}\!)\!-\!\tau(1\!\!+\!\!\hat{f}_{w}^{r}(t_0\!)). \label{eq:gamma_t_1}
-\end{eqnarray}
-Apparently, the objective is to minimize the prediction error so that the event reception time is estimated correctly.
-As can be observed from \eqref{eq:gamma_t_1}, the estimation error on the relative clock frequency at time $t_0$, i.e.
-$\hat{f}_w^r(t_0)$, depends on the mean event period $\tau$, and its variance (since they affect $C_w(t_1)-C_w(t_0)$).
-Inspired by the integral controller update rule in \cite[Section III, (5)]{pi2015}, we propose to update the estimated relative frequency at time $t_1$ as
-\begin{eqnarray}
-\hat{f}_{w}^{r}(t_1) &=& \hat{f}_{w}^{r}(t_0)+\beta\gamma(t_1), 
-\label{eq:update_rule}
-\end{eqnarray}
-where $\beta$ is the \emph{integral gain}.
-Assume that the uncertainty of the event period is $\varepsilon\sim\mathcal{N}(0,\sigma^2_{\tau})$ and the instantaneous change on the relative clock frequency is $\frac{df_{w}^{r}(t)}{dt}\sim\mathcal{N}(0,\sigma_{\eta}^{2})$.
-Under these assumptions, it can be proven that if $0<\beta<\frac{2}{\tau}$ is satisfied the synchronization will hold eventually; i.e.
-$\hat{f}_{w}^{r}=f_{w}^{r}$.
-Moreover, the mean and the variance ($\sigma^2_\gamma$) of the steady-state (asymptotic) estimation error of WISP-Sync can be given by
-\begin{eqnarray}
-\gamma & \!\!\!\!\!\sim\!\!\!\!\!
-& \mathcal{N} \!\!\left(\!\!0,\sigma^2_\gamma\!\!=\!\!\frac{\sigma^2_{\eta}\!\left(\!\tau^2\!-\!\beta\tau^3\!+\!\frac{\beta^2\tau^4}{3}\!\right)\!\!+\!\!\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)}\!\!+\!\!\frac{\sigma^2_{\eta}\tau^3}{3}\!+\!\sigma^2_{\tau} \!\!\right),
-\label{eq:sync_performance}
-\end{eqnarray}
-which is derived in Section~\ref{sec:app_b} of the Appendix.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=\columnwidth]{figures/sim-theory.eps}
-	\caption{\label{fig:theory-vs-simulations} The plot of the cumulative distribution function (CDF) of $\gamma$ in \eqref{eq:sync_performance} and the CDF of the steady-state $\gamma$ obtained during numerical simulations of \eqref{eq:update_rule} (by implementing \eqref{eq:theta} and \eqref{eq:delta_recursive} in Appendix).
-We used $\tau=1$, $\beta= 0.0001$, $\sigma_{\tau}=0.00005$ and our plots depict the theoretical and simulation results with respect to $\sigma_\eta = 0.0002$ and $\sigma_\eta = 0.0008.$ It can be observed that \eqref{eq:sync_performance} is consistent with the simulations and bigger uncertainty leads to bigger synchronization error.}
-\end{figure}
-In Fig.~\ref{fig:theory-vs-simulations}, we emphasize that our theoretical derivation in \eqref{eq:sync_performance} is consistent with the simulations of \eqref{eq:update_rule}.
-We refer again the reader to the Appendix for the detailed proofs of the convergence and the asymptotic estimation error.
-Moreover, in Section~\ref{sec:adaptibility}, we discuss the impact of the parameters $\tau$ and $\beta$ on the synchronization accuracy.
-
-\subsubsection{The Synchronization Algorithm}
-
-\begin{algorithm}[t]
-%	\setstretch{1.35}
-	\caption{\label{alg:sync} \emph{WISP-Sync}: An integral controller-based 
-		algorithm at WISP tag side.}
-	\begin{algorithmic}[1]	
-		\small
-
-		\Statex\textbf{Definitions:}
-		\Statex$t_f$ \Comment{local time of the first \emph{BlockWrite}} 
-		\Statex $C_w$ \Comment{bult-in clock of the WISP tag}
-		\Statex $\hat{f}_w^r\leftarrow0$ \Comment{estimated relative clock 
-			frequency}	
-		\Statex $\tau\leftarrow$measured value \Comment{mean \emph{event period}}	
-		\Statex $\beta\leftarrow$ a value satisfying 
-		$0<\beta<\frac{2}{\tau}$ \Comment{integral gain}	
-		\Statex $V_{\min}, V_{\text{op}}$ \Comment{voltage thresholds ($V_{\text{op}}>V_{\min}$)}
-		\Statex $V_{\text{cur}}$ \Comment{current voltage}
-
-		\State $\square$ \textbf{Upon receiving the first \emph{BlockWrite}} 
-		\State \quad $ t_f = C_w$ \Comment{store the local time in $t_f$}
-		\State $\square$ \textbf{Upon receiving the last \emph{BlockWrite}} 
-		\State \quad $\gamma = C_w- (t_f+\tau(1+\hat{f}_w^r))$ 
-		\Comment{calculate estimation error $\gamma$} 
-		\State \quad $\hat{f}_w^r = \hat{f}_w^r +\beta\gamma$ \Comment{apply 
-			integral control}
-		\State $\square$ \textbf{Upon} $V_{\text{cur}}=V_{\min}$
-		\State \quad store $\hat{f}_w^r$ into FRAM \Comment{store in non-volatile memory}
-		\State $\square$ \textbf{Upon} $V_{\text{cur}} < V_{\text{op}}$
-		\State \quad restore $\hat{f}_w^r$ from FRAM \Comment{restore the clock frequency}
-	\end{algorithmic}	
-\end{algorithm}
-
-Based on this model we introduce Algorithm \ref{alg:sync}, namely \emph{WISP-Sync}, that synchronizes the speed of the software clock to the reference clock's speed using the update rule defined in \eqref{eq:update_rule}.
-The steps of this algorithm can be explained as follows.
-Initially, the tag lets its software clock run at the same speed of its local clock by setting its relative clock rate to zero.
-Upon receiving the first \emph{BlockWrite} event (Line 1), the tag stores its local clock value at the variable $t_f$ (Line 2).
-After receiving the last \emph{BlockWrite} event (Line 3), first it calculates the estimation error (Line 4).
-Then, it applies the correction on its estimated relative rate $\hat{f}_w^r$ by multiplying the estimation error with the integral gain $\beta$ and adding it to $\hat{f}_w^r$ (Line 5).
-After receiving successive events, $\hat{f}_w^r$ will converge to its desired value eventually, as proven in the Appendix.
-
-\subsubsection{State Recovery}
-
-Due to intermittent RF power, the tag is subject to frequent power losses, which will lead to loss of the synchronization state.
-In voltage-aware systems~\cite{Buettner:2011:Dewdrop,Ransford:2011:Mementos,Lucia:2015:Execution,Colin:2016:Debugger}, the general idea is to save the system state when the input voltage is below a pre-defined threshold $V_{\min}$ and then recover it back when the input voltage is above a pre-defined threshold $V_{\text{op}}$.
-As shown in Algorithm~\ref{alg:sync}, the only part of the system state pertaining to synchronization is the variable $\hat{f}_w^r$.
-Therefore, if this variable is also saved to FRAM (when there is a power loss and recovered when there is sufficient voltage) the synchronization will resume from the point of power loss, that will prevent loss of synchronization and lead to improved synchronization accuracy.
-These steps are summarized in Lines 6--9 of Algorithm~\ref{alg:sync}.
-It should be noted that these steps require hardware support so that the voltage level is monitored continuously and the corresponding voltage level indicator events are generated.
-Another approach can be keeping $\hat{f}_w^r$ in non-volatile memory as a \emph{persistent variable} so that the corresponding FRAM memory location of $\hat{f}_w^r$ is updated immediately when this variable is updated---eliminating the need for Lines 6--9 of Algorithm~\ref{alg:sync}, and in turn extra voltage measurement hardware, but introducing more FRAM access overhead.
-
-\subsubsection{Computation Complexity\label{subsec:complexity}}
-
-It can be observed that WISP-Sync employs only very simple arithmetic operations at each step, e.g.
-Lines 3--5 are composed of only two subtraction, two addition and two multiplication operations.
-Moreover, the \emph{state of synchronization}, i.e.
-the parameters that are required to be saved in the non-volatile memory, is represented by only the variable $\hat{f}_w^r$.
-As a consequence, WISP-Sync addresses \emph{Challenge I} in Section~\ref{sec:Motivation} in terms of computation and memory overhead.
-
-\subsubsection{Adaptability} 
-\label{sec:adaptibility}
-
-\begin{figure*}
-	\centering
-	\includegraphics[width=1.4\columnwidth]{figures/alpha-period-vs-variance.eps}
-	\caption{\label{fig:simulations} Numerical results showing the relations between event period, integral gain and the steady-state variance.
-First, we set $\sigma_{\tau}=0.4097$, $\sigma_\eta = 0.0008$, $\beta=1$ and plotted $\sigma^2_{\gamma}$ using different $\tau$ (left hand side).
-Observe that increasing $\tau$ increases $\sigma^2_{\gamma}$.
-Then, we set $\sigma_{\tau}=0.4097$,	$\tau=1$ and $\beta \in (0,2)$ due to \eqref{eq:step_size_bounds-pi}.
-Using different values for $\sigma_\eta$, we plotted how $\sigma^2_{\gamma}$ changes with respect to the $\beta$ (right hand side).
-Observe that for each value of $\sigma_\eta$, there is a particular $\beta$ that minimizes $\sigma^2_{\gamma}$.}
-\end{figure*}
-
-In Section \ref{subsec:event_period_uncertainty} we presented our measurements pertaining to the event period and showed 
-that $\sigma_{\tau}=0.4097$.
-Moreover, from Fig.~\ref{fig:frequency}, we calculated $\sigma_{\eta} \approxeq 0.0008$ using the \textsf{normfit} function of MATLAB.
-Substituting these values in \eqref{eq:sync_performance}, we performed MATLAB simulations to calculate $\sigma^2_\gamma$, which are depicted in Fig.~\ref{fig:simulations}.
-These simulations led us to reveal the following points:
-
-\begin{itemize}
-	\item Considering our previous measurements, it is apparent that $\sigma^2_{\tau}$ is the dominant factor effecting the synchronization performance.
-Therefore, it is desirable to keep transmission delay variations and also $\tau$, as small as possible.
-Apparently, the bigger the resynchronization period $\tau$, the worse the synchronization error is (see left hand side of Fig.~\ref{fig:simulations}).
-
-	\item Parameter $\beta$ has significant impact on the performance of the WISP-Sync.
-From 	\eqref{eq:sync_performance}, it can be concluded that the synchronization error can be minimized using a proper value of $\beta$.
-However, the $\beta$ value that minimizes 	\eqref{eq:sync_performance} depends on the parameters $\sigma_{\eta}$ and $\sigma_{\tau}$ which represent randomness of the system.
-In Fig.~\ref{fig:simulations} (right hand side), we fixed $\tau$ and $\sigma_{\tau}$, we used different values of $\sigma_{\eta}$ and we plotted $\sigma^2_{\gamma}$ with respect to the values of the $\beta$ that will lead to convergence.
-Observe that a different $\beta$ value minimizes $\sigma^2_{\gamma}$ for each different $\sigma_{\eta}$.
-
-	\item Due to varying RF power levels, $\sigma_{\eta}$ is also a time-varying value.
-Therefore, $\beta$ should be selected to compensate for also \emph{dynamic} $\sigma_{\eta}$ in a \emph{fast manner} to keep synchronization error within desired bounds---which addresses \emph{Challenge II} in Section~\ref{sec:Motivation}.
-However, the $\beta$ that minimizes $\sigma^2_{\gamma}$ may lead to longer convergence time (see Appendix).
-Therefore, $\beta$ also defines how \emph{reactive} WISP-Sync is against the dynamic environment, i.e.
-the \emph{adaptivity}.
-\end{itemize}
-
-\section{Evaluation of WISP-Sync}
-\label{sec:eval-wispsync}
-
-In order to evaluate WISP-Sync, we considered the value of $\gamma$ in Algorithm~\ref{alg:sync} (the estimation error) as an evaluation metric.
-We collected the local clock readings of the tag at the first and last \emph{BlockWrite} events using our testbed setup and sniffer, and used MATLAB to implement WISP-Sync algorithm which processes the collected clock values to output the software clock.
-This gave us the flexibility to try different approaches without reprogramming the tag.
-
-\subsection{Selection of the Integral Gain $\beta$}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.85\columnwidth]{figures/integral_gain.eps}
-	\caption{\label{fig:beta} The synchronization error tends to get smaller with smaller integral gains, but after some point decreasing it has no significant effect due to the precision of 32\,kHz clock.
-Hence, we selected a fixed value of $\beta=0.0001$ for the next steps.
-}
-\end{figure}
-
-Since we measured the event period as 226.76\,ms on average (see Fig.~\ref{fig:block_write_delay}), we set $\tau=7086$ 
-clock ticks since the local clock of the tag is operating at 32\,kHz and each clock tick occurs every 32 microseconds.
-First, we explored how $\beta$ affects the performance of the algorithm.
-By substituting $\tau$ into \eqref{eq:step_size_bounds-pi}, the convergence condition becomes $0<\beta<0.000284$ in our case.
-We present the synchronization error with different $\beta$ values in Fig.~\ref{fig:beta}.
-Observe that the synchronization error tends to get smaller with smaller $\beta$ values but the convergence time increases negligibly.
-Moreover, due to the low-precision 32\,kHz clock, decreasing $\beta$ has no significant effect after some point.
-We mention that  in order to make WISP-Sync adapt faster, WISP-Sync can start with a big value of $\beta$ to converge fast and then decrease $\beta$ gradually to decrease the synchronization error, i.e.
-the adaptive strategy proposed in~\cite{pi2015}.
-However, from  Fig.~\ref{fig:beta}, we consider a fixed value of $\beta=0.0001$ for the next evaluation steps.
-
-\subsection{Synchronization Performance}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.9\columnwidth]{figures/synchronization_accuracy.eps}
-	\caption{\label{fig:sync_error}Event-based synchronization performance when the tag is powered through a constant voltage source and through RF power.
-Mean synchronization errors without and with WISP-Sync were almost 127 and 16 clock ticks under stable voltage, and 175 and 22 clock ticks with RF power harvesting (represented by the straight lines in this figure), respectively.}
-\end{figure}
-
-Fig.~\ref{fig:sync_error} presents the synchronization error measurements when the tag is powered by using a constant voltage source, and powered only by an RFID reader.
-Measurements under constant and stable voltage allowed us to observe synchronization under stable clock frequency.
-We obtained a significant synchronization performance with WISP-Sync---almost a factor of eight less synchronization error as compared to the case where we did not perform any synchronization.
-It should be noted that even in this case, we observed quite fluctuating synchronization errors due to the varying transmission delays, that led a peak error to appear between samples 50 and 60.
-Apart from experiments with constant voltage input, measurements under highly varying RF power led us to observe the behavior of synchronization under highly varying clock frequencies.
-In this case, there is a considerable amount of increase on the error, i.e.
-approximately twice as much, as compared to the stable voltage case.
-However, we also obtained considerable improvements with our approach with unstable clock frequencies.
-WISP-Sync was quite reactive to the frequently varying clock frequencies and provided almost a factor of eight better accuracy.
-
-\subsection{Power Loss and Recovery}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.85\columnwidth]{figures/power_loss.eps}
-	\caption{\label{fig:power_loss}Impact of the state recovery on the synchronization performance.
-During the power loss periods, the tag is assumed to be turned off.
-With state recovery, presented by the solid line, the relative clock frequency is assumed to be saved in the FRAM and recovered back when sufficient power is supplied.
-On the other hand, the dashed line represents the case of starting synchronization from scratch.
-It can be observed that with state recovery, the tag adapts itself quickly and synchronization performance is considerably better, i.e.
-more than 100 clock ticks less synchronization error on power loss.}
-\end{figure}
-
-In order to evaluate the performance of WISP-Sync under power losses, we simulated power loss in MATLAB by omitting the collected clock readings during predefined time intervals.
-As presented in Fig.~\ref{fig:power_loss}, after samples 10 and 30 WISP-Sync is disabled and before the beginning of the sample times 20 and 40 the algorithm is enabled.
-The solid line in this figure indicates enabling algorithm with state recovery, where the algorithm starts to execute using the value of $\hat{f}_w^r$  previously saved in the FRAM before power loss.
-It is apparent that with state recovery, the tag quickly adapts itself and re-synchronization is quickly established.
-On the other hand, starting the algorithm from scratch, as presented by the dashed lines, is problematic since it takes several steps to re-synchronize the tag.
-%
-%\begin{figure}
-%		\begin{lstlisting}[style = mystyle,escapechar=|] 
-%		_attribute_((section(".infoA"))) float f_w = 0.0f;
-%		uint16_t t_f = 0;
-%		
-%		void wisp_sync(uint16_t c_w) {
-%			float error = (float)c_w - ((float)t_f + TAU * (1.0f + f_w));
-%			f_w += BETA * error;
-%		}
-%		\end{lstlisting}
-%\caption{The actual implementation of  lines 3--5 of Algorithm~\ref{alg:sync} in C language for TI MSP430FR5969 platform.
-Constants \texttt{BETA} and \texttt{TAU} are defined outside of this function in the remaining part of the C program implementing Algorithm~\ref{alg:sync}.}
-%		\label{fig:wisp_implementation}
-%\end{figure}
-%
-\subsection{Energy Overhead}
-%
-We now present the real energy measurements of the actual WISP tag-side C language implementation of lines 3--5 of Algorithm~\ref{alg:sync} which are the main computation steps to calculate the new values of the variables pertaining to the synchronization.
-We performed our measurements using the Energy-interference-free Debugger (EDB)~\cite{colin2016energy} to measure the output voltage without affecting the energy state of the tag.
-EDB charges the capacitor of a device to a certain level and then disconnects itself.
-Using this feature of EDB we charge the capacitor of WISP to 2.45\,V, i.e.
-just above the operational level of WISP.
-While the program is being executed, the value of the capacitor is measured at specific code locations---in this case before Line 3 and after Line 5 of Algorithm~\ref{alg:sync}.
-The difference between these two results represents the energy overhead of WISP-Sync.
-We repeated this experiment ten times to eliminate quantization effects during measurements.
-We observed an average energy of $\approx1.85 \mu$J with a standard deviation of 0.038\,$\mu$J is spent by the tag in order to perform the aforementioned computations.
-
-The main source of energy consumption in our implementation is the \emph{floating point arithmetic}: WISP-Sync requires two multiplications, three additions and one subtraction on float variables.
-Since MSP430 microcontroller does not have any floating point coprocessor, the float operations are handled using a software library, that makes them relatively slow.
-We used the standard gcc compiler (specifically, msp430-gcc version 6.2.1.16) math library for the sake of the portability of float operations.
-However, we expect that a more efficient architecture-specific implementation of the math library in software or an architecture with a math coprocessor will lead to less computation time and in turn relatively less energy overhead for WISP-Sync~\cite{mathlib_website}.
-Another point we would like to mention is that non-volatile memory access consumes more power as compared to SRAM access~\cite[Section 3.3]{ti_fram}.
-The only non-volatile variable in WISP-Sync is $\hat{f}_w^r$ in Algorithm~\ref{alg:sync}.
-This variable is read twice but modified once in our implementation.
-On the other hand, WISP-Sync maintains only one variable in SRAM, i.e.
-$t_f$, which is read only once.
-We conclude that WISP-Sync performs three FRAM and one SRAM access operations---low energy overhead in terms of memory access.
-
-\subsection{Limitations of WISP-Sync}
-
-In conclusion, the experimental results so far indicated that a maximum synchronization error of approximately 1.5 ms can be ensured between an RFID reader and a single WISP tag when employing event-based synchronization.
-The main factors that limited the WISP-Sync synchronization accuracy were the clock precision of the WISP platform and the non-deterministic transmission delays during reader-tag communication.
-As we indicated, we used the external 32\,kHz clock of the platform since it is also running when the tag is in very low-power operation mode.
-Therefore, in the best case when there are no transmission delays, the quantization effect due to the timing measurements using the low precision clock will not allow a better synchronization accuracy than 32\,$\mu$s.
-On the other hand, to obtain this synchronization accuracy with WISP-Sync, the reader and the tags shoud support a high-precision and accurate packet timestamping as in sensor networks~\cite{Maroti2004}.
-As indicated in Section \ref{sec:CRFIDs}, the effect of the non-deterministic delays should be reduced for a better synchronization accuracy.
-
-Even though the event-based approach is relatively lightweight and adaptive as compared to the sender-receiver synchronization, it has some limitations.
-First, the tag is unable to obtain an \emph{explicit reference clock value}.
-The only variable it tunes is the $\hat{f}_w^r$ that represents the relative frequency of the software clock with respect to the clock of the reader.
-Therefore, we require additional steps that will allow tags to obtain the actual reference clock value.
-However, to the best of our knowledge, there is no such command in the EPC Gen 2 standard that allows to send the explicit RFID reader time to the tag.
-Second, WISP-Sync requires knowledge about $\tau$, which can be RFID reader dependent and requires a sniffer setup described in Section~\ref{sec:sniffer} to measure.
-This limitation can be addressed by employing an iterative estimation as an initial step to the WISP-Sync: the tag can start with an initial estimated $\tau$ value, and improve its estimation as it receives successive packets from the reader.
-After some iterations, the tag can have a rough estimate $\hat{\tau}$ that can be used on behalf of $\tau$.
-However, estimation errors might also perturb the synchronization accuracy.
-Last, the steps in Fig.~\ref{fig:event_based} present the communication scenario between one WISP and the reader.
-To allow other tags sniff this communication and synchronize, a broadcast primitive is crucial---in the next section we show how to overcome this limitation.
-
-\section{The Synchronization of Multiple WISP Tags Using the WISP-Sync}
-\label{sec:multi-wisp}
-
-In this section, we turn our attention to the synchronization of  multiple WISP tags using the WISP-Sync.
-To this end, we have done the \emph{real implementation} of the WISP-Sync using the C programming language and TI's CCS~\cite{ti_css}, instead of using MATLAB to simulate our protocol using the collected the clock values from our testbed.
-Moreover, we extended our experimental testbed in Section~\ref{sec:testbed} by introducing two additional tags, as presented in Fig.~\ref{fig:testbed-new}.
-
-\begin{figure}
-	\centering
-	\begin{tikzpicture}
-	\node[anchor=south west, inner sep=0] at 
-	(0,0){\includegraphics[width=0.8\columnwidth]{figures/3-wisps_resized.jpg}};
-
-	\node [fill= white,circle] at (0.8,2.5) { };
-	\node [color=black ] at (0.8,2.5) { 1 };
-
-	\node [fill= white,circle] at (1,5.1) {  };
-	\node [color=black ] at (1,5.1) { 2 };
-
-	\node [fill= white,circle] at (2.3,2.1) {  };
-	\node [color=black ] at (2.3,2.1) { 3 };
-
-	\node [fill= white,circle] at (5.5,1) {  };
-	\node [color=black ] at (5.5,1) { 4 };
-
-	\node [fill= white,circle] at (5.5,2.8) {  };
-	\node [color=black ] at (5.5,2.8) { 5 };
-
-	\end{tikzpicture}
-	\caption{\label{fig:testbed-new} A testbed of multiple tags: (1,2,3) tags, (4) logic analyzer, (5) breadboard.
-Tags 1 and 3 were at 8\,cm while tag 2 was at 20\,cm distance to the reader.
-The breadboard was used to make connections between the logic analyzer and the tags.}
-\end{figure}
-
-We programmed the RFID Reader to send synchronization points via \emph{BlockWrite} command with a period of one second and we programmed two tags using our WISP-Sync implementation.
-On the other hand, we disabled clock synchronization for the last tag---which was a baseline for our evaluation.
-In order to allow tags to simultaneously receive the information sent by the RFID Reader, we exploited an overhearing mechanism presented first in ~\cite[Section III-B]{stork:2017}.
-Therefore, during the experiments in our testbed (i) all of the tags received the periodic synchronization points sent by the RFID Reader; (ii) only two tags process this information and run Algorithm~\ref{alg:sync} to update their software clocks; (iii) the last tag did not participate in synchronization.
-
-In order to measure the synchronization accuracies of the tags, we configured one of the ports of the MSP430 microcontroller as an output port and let tags (i) to set this port when they receive the first \emph{BlockWrite} command in Algorithm~\ref{alg:sync} (Line 1); (ii) to set a timer that will fire at their predicted time of the reception of the last \emph{BlockWrite} command from the RFID Reader; (iii) to clear this port when their timer fires.
-We connected the output ports to logic analyzer~\cite{saleae} and measured the time difference between the times at which the output ports are cleared.
-This allowed us to measure the prediction errors among the tags.
-We observed that all the tags set their output ports when they overhear the first \emph{BlockWrite} command, however their clear times are quite different and depend on their prediction error---the port clear times of the synchronizing two tags were closer to each other while they were quite different than the unsynchronized tag.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.9\columnwidth]{figures/multiwisp-experiments.eps}
-	\caption{\label{fig:multi-sync-results}The synchronization error calculated during the experiments in our testbed.
-The blue dotted line denotes the maximum synchronization error among all of the tags (including the one which did not participate in synchronization) whereas the red line denotes the maximum error just between the synchronized tags.}
-\end{figure}
-
-In Fig.~\ref{fig:multi-sync-results}, we plotted two different results: (i) the blue line depicts the maximum synchronization error (i.e.
-the maximum pairwise difference between the port clear times) when we considered all of the tags; (ii) the red line depicts when only synchronized tags are considered.
-From this figure it can be concluded that synchronization led to a factor of approximately four times better prediction on average among the tags: we observed an average synchronization error of 1\,ms while it was 4\,ms when one of the tags was unsynchronized.
-
-\section{Conclusions and Future Research Directions}
-\label{sec:Conclusion}
-
-In this article we explored synchronization scenarios between an RFID reader and WISP tags.
-We studied sender-receiver and event-based synchronization mechanisms in this setting and provided initial designs that will guide future explicit synchronization mechanisms among individual WISPs that reside inside the communication range of a common RFID reader.
-We provided implementation and evaluation of these designs in our testbed and identified their limitations and drawbacks.
-Our main finding is that with lightweight mechanisms, as of now, an average synchronization error of approximately 1\,ms can be ensured among tags.
-We provide the following issues for future studies in this domain:
-
-\begin{itemize}
-	\item \textbf{Network-wide synchronization:} We studied event-based synchronization among many tags.
-However, synchronization of these tags where they can access the explicit clock of the RFID reader and also synchronization of the whole CRFID network composed of several RFID readers and tags is still an issue, due to the single-hop nature of backscatter communication.
-Since tags can only communicate with the reader, not with their neighboring tags, it is interesting to explore \emph{WISP to WISP} synchronization.
-
-	\item \textbf{Power loss and recovery:} It is crucial to save the synchronization status to non-volatile memory just before the power loss, but writing to non-volatile memory also consumes energy.
-Hence, when to save the synchronization status without extra hardware support is worth exploring~\cite{Buettner:2011:Dewdrop,Ransford:2011:Mementos,Lucia:2015:Execution,Colin:2016:Debugger}.
-
-	\item \textbf{Voltage-frequency relations:} It might be interesting to explore the voltage-clock frequency relationship since1 tags are subject to varying voltage source when they are powered with only RF energy harvesting from the RFID reader.
-We anticipate that voltage level should be incorporated to the establishment of the software clock so that voltage dependent instability of clock frequency can be compensated.
-However, this is a ``chicken or egg'' problem since reading the voltage level also consumes energy.
-
-	\item \textbf{Synchronization in other IPD platforms:} Synchronization in other IPD platforms, e.g. that use ambient backscatter communication \cite{ambient_backscatter}, is worth exploring as well. 
-\end{itemize}
-
+\newpage\null\thispagestyle{empty}\newpage
 \appendices
@@
 which results in \eqref{eq:sync_performance}.
-
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-\bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,references}
-
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/ksy.jpeg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
-%received BSc(Eng), M.Sc. and Ph.D. degrees from Ege University, İzmir, 
-%Turkey in 2003, 2006 and 2012 respectively.
-He is  currently with Embedded Software group of TU Delft.
-His research interests are embedded systems, distributed algorithms,  wireless sensor networks, energy harvesting systems and computational RFIDs.
-%\end{IEEEbiography}
-%
-%% if you will not have a photo at all:
-%\begin{IEEEbiographynophoto}{Henko Aantjes}
-%Biography text here.
-%\end{IEEEbiographynophoto}
-%
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.jpg}}]{}
-%Biography text here.	
-%\end{IEEEbiography}
-%
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
-%Biography text here.
-%\end{IEEEbiography}
-%
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
-\end{document}

### 2017-11-01 cover letter/cover.tex
TMC submission

@@
 	\begin{itemize}
-		\item a more detailed related work in Section II;
+		\item a more detailed related work in Section 2;
 		\item extended descriptions of the hardware platform and testbed in Section 
-		III;
-		\item a mathametical framework in Section IV to analyze the proposed 
+		3;
+		\item a mathametical framework in Section 4 to analyze the proposed 
 		synchronization algorithms;
 		\item extended descriptions of the proposed synchronization algorithms in 
-		Section V and VI;
+		Section 5 and 6;
 		\item a rigorous theoretical analysis of the proof of convergence and 
@@
 		performance of the proposed solutions under the loss of computational 
-		state in Section VII-C;	
-		\item an experimental work to evaluate the energy overhead of the proposed solutions in Section VII-D;	
+		state in Section 7-3;	
+		\item an experimental work to evaluate the energy overhead of the proposed solutions in Section 7-4;	
 		\item implementation and evaluation of the synchronization of 
-		many WISP tags in Section VIII.
+		many WISP tags in Section 8.
 	\end{itemize}

### 2017-11-01 manuscript_tmc.tex
TMC submission

@@
 
-\appendices
-\section{Analysis of Algorithm~\ref{alg:sync}}
-
-We provide the synchronization conditions of Algorithm~\ref{alg:sync} and show that the synchronization will be established eventually.
-We need to mention that we follow similar calculation steps to the ones presented in the studies~\cite{pi2015,Yildirim:Gradient:2016}.
-However, since we use a different clock drift model in this article, i.e.
-Brownian motion rather than uniform distribution, the following  mathematical analysis is moderately different and more realistic as compared to those in the previous studies.
-
-Since $f_w$ and $f_r$ are time varying, by following 
-\cite{Zhong:2011:On-demand,Huang:2013:PSR}, we model the 
-\emph{evolution} 
-of $f_w^r$ as 
-integrated white noise, i.e. \emph{Brownian motion}, given as
-\begin{eqnarray}
-f_{w}^{r}(t_{0}+\delta) & \triangleq & 
-f_{w}^{r}(t_{0})+\int_{t_{0}}^{t_{0}+\delta}\eta(t)dt.\label{eq:drift}
-\end{eqnarray}
-Here, $\delta$ denotes the amount of time after $t_0$ and $\eta(t)$ denotes 
-the instantaneous change on the relative clock frequency, namely 
-\emph{relative drift rate} at time $t$. We assume that 
-$\eta(t)\sim\mathcal{N}(0,\sigma_{\eta}^{2})$ and for all 
-time instants $t'$ and $t''$, $\eta(t')$ and $\eta(t'')$ are uncorrelated 
-random variables. In order to ease the calculations in the following sections, 
-we will consider the \emph{average relative clock frequency} in an interval of 
-interest $[t_0,t_0+\delta]$ rather than instantaneous relative frequency 
-in~\eqref{eq:drift}. We denote this value by $\bar{f}_{r}^{w}(t_0,t_0+\delta)$ 
-and define it as 
-\begin{eqnarray}
-\bar{f}_{w}^{r}(t_0,t_0\!+\!\delta) &\!\!\!\!\! \triangleq\!\!\!\!\! & 
-\frac{1}{\delta}\!\!\int_{t_0}^{t_0+\delta}\!\!\!\!\!\!\!\!\!\!\!\! 
-f_w^r(t)dt \triangleq
-f_{w}^{r}(t_{0})\!+\!\frac{1}{\delta}\!\!\int_{0}^{\delta}\!\!\int_{t_{0}}^{t_{0}+t}\!\!\!\!\!\!\!\!\!\!\!\eta(u)dudt.
-\label{eq:drift_relation}
-\end{eqnarray}
-
-\subsection{Proof of Convergence}
-Consider \eqref{eq:gamma_t_1} and without loss of 
-generality we assume that 
-$C_{w}(t_{1})-C_{w}(t_{0})=\tau(1+\bar{f}_{w}^{r}(t_0,t_1))+\varepsilon$
-where $\tau$ denotes the event period and $\varepsilon$ accounts for the 
-uncertainty of the event period (as 
-indicated in Section \ref{subsec:event_period_uncertainty}). For ease of 
-calculations we assume that $\varepsilon\sim\mathcal{N}(0,\sigma^2_{\tau})$. 
-Define \emph{relative frequency estimation error} as
-\begin{eqnarray}
-\tilde{f}_{w}^{r}(t_0) & \triangleq & 
-f_{w}^{r}(t_0)-\hat{f}_{w}^{r}(t_0).\label{eq:relative_drift_estimation_error}
-\end{eqnarray}
-Using \eqref{eq:gamma_t_1}, \eqref{eq:drift_relation} and 
-\eqref{eq:relative_drift_estimation_error} we get:
-\begin{eqnarray}
-\gamma(t_1) & = & 
-\tau(\bar{f}_{w}^{r}(t_0,t_1)-\hat{f}_{w}^{r}(t_0))+ \varepsilon \nonumber \\
- &=&  
-\tau\tilde{f}_{w}^{r}(t_0)+\frac{\tau}{t_1-t_0}\int_{0}^{t_1-t_0}\!\!\!\int_{t_{0}}^{t_0+t}\!\!\!\!\!\!\!\!\!\!\!\eta(u)dudt
-+ \varepsilon.
-\end{eqnarray}
-For ease of representation, let us introduce the following 
-notation:
-\begin{eqnarray}
-\Phi(h) &\triangleq& \int_{t_h}^{t_{h+1}}\eta(t)dt, \\
-\Omega(h) &\triangleq& 
-\frac{\tau}{t_{h+1}-t_h}\int_{0}^{t_{h+1}-t_h}\int_{t_{h}}^{t_h+t}\eta(u)dudt,
-\\
-\Delta(h) &\triangleq& \tilde{f}_{w}^{r}(t_{h}). \label{eq:delta}
-\end{eqnarray}
-Using this notation, we denote $\gamma(t_{h+1})$ by $\Theta(h+1)$ as
-\begin{eqnarray}
-\Theta(h+1) &\triangleq& \gamma(t_{h+1})=\tau\Delta(h) + \Omega(h)+\varepsilon. 
-\label{eq:theta}
-\end{eqnarray}
-Moreover, since $\Delta(h)$, $\Omega(h)$ and $\varepsilon$ are 
-independent random variables, we can write $\Delta(h+1)$ as
-\begin{eqnarray}
-\Delta(h+1) &\overset{\eqref{eq:delta}}{=}& 
-\Delta(h)+\tilde{f}_{w}^{r}(t_{h+1})-\tilde{f}_{w}^{r}(t_{h}) \nonumber \\
-&\overset{\eqref{eq:relative_drift_estimation_error}}{=}& \Delta(h)
-\!+\!f_w^r(\!t_{h+1}\!)\!-\!\!f_w^r(\!t_{h}\!) \!-\!\!\hat{f}_w^r(\!t_{h+1}\!)\!+\!\!\hat{f}_w^r(\!t_{h}\!)
-\nonumber\\
-&\overset{\eqref{eq:update_rule},\eqref{eq:drift}}{=}& 
-\Delta(h)+\int_{t_h}^{t_{h+1}}\eta(t)dt-\beta\Theta(h+1) \nonumber \\
-&\overset{\eqref{eq:theta}}{=}& 
-(1-\beta\tau)\Delta(h)-\beta(\Omega(h)+\varepsilon) +\Phi(h). 
-\label{eq:delta_recursive}
-\end{eqnarray} 
-Considering \eqref{eq:theta} and \eqref{eq:delta_recursive}, we can write the 
-system evolution as
-\begin{eqnarray}
-\underset{X(h+1)}{
-	\underbrace{
-		\left[\!\!\!\begin{array}{c}
-		\Theta(h\!+\!1)\\
-		\Delta(h\!+\!1)
-		\end{array}\!\!\!\right]
-	}
-} &\!\!\!\!\!\!\! =\!\!\!\!\!\!\! & 
-\underset{A}{
-	\underbrace{
-		\left[\!\!\!\begin{array}{cc}
-		0 & \tau\\
-		0 & 1\!-\!\beta\tau
-		\end{array}\!\!\!\right]
-	}
-}
-\underset{X(h)}{
-	\underbrace{
-		\left[\!\!\!\begin{array}{c}
-		\Theta(\!h\!)\\
-		\Delta(\!h\!)
-		\end{array}\!\!\!\right]}
-}
-\!\!+\!\!
-\underset{Y(h)}{
-	\underbrace{
-		\left[\!\!\!\begin{array}{c}
-		\Omega(h)+\varepsilon\\
-		\Phi(\!h\!)\!-\!\beta(\Omega(\!h\!)\!+\!\varepsilon)
-		\end{array}\!\!\!\right]}
-}.\label{eq:state_space-pi}
-\end{eqnarray}
-Taking the expectation of both sides of \eqref{eq:state_space-pi} yields 
-$\expected{X(h+1)} = A\expected{X(h)}$ since 
-the means of the random variables in the matrix $Y(h)$ are all zero. According 
-to stability of linear systems \cite{book-luenberger}, \emph{asymptotic 
-	convergence} will be reached if and only if the magnitudes of the 
-eigenvalues of the matrix $A$ are strictly smaller than one. The eigenvalues 
-$\lambda_1$ and $\lambda_2$ of the matrix $A$ can be obtained 
-by solving
-\begin{eqnarray}
-\left|\begin{array}{cc}
--\lambda & \tau\\
-0 & 1-\beta\tau-\lambda
-\end{array}\right|=0 &\implies& 
-\begin{array}{c}
-\lambda_1 = 0,\\
-\lambda_2 = 1-\beta\tau.  
-\end{array}\label{eq:eigens}
-\end{eqnarray}
-Therefore, the synchronization will be  established if 
-and only if $|\lambda_1,\lambda_2|<1$, and in turn
-\begin{align}
-0 & <\beta<\frac{2}{\tau}\label{eq:step_size_bounds-pi}
-\end{align}
-should hold. The 
-rate of convergence is governed by 
-the largest eigenvalue $\lambda_2$, which means that bigger values of $\beta$ 
-will lead to faster convergence. 
-Consequently, selecting $\beta$ within the bound above will lead the system  
-converge to the state  
-\begin{eqnarray}
-\underset{h\rightarrow\infty}{\lim}\expected{\Theta(h+1)} &=&
-\expected{\Theta(h)}, \label{eq:asymp_theta} \\
-\underset{h\rightarrow\infty}{\lim}\expected{\Delta(h+1)}&=&
-\expected{\Delta(h)} \label{eq:asymp_delta}.
-\end{eqnarray}
-Hence, due to asymptotic convergence
-\begin{eqnarray}
-\expected{\Delta(\infty)} \overset{\eqref{eq:delta_recursive}}{=} 
-\left(1-\beta 
-\tau\right)\expected{\Delta(\infty)} &\!\!\!\!\!\!\!\implies\!\!\!\!\!\!\!& \expected{\Delta(\infty)}\!=\!0, 
-\label{eq:delta_convergence}\\
-\expected{\Theta(\infty)}  \overset{\eqref{eq:theta}}{=} 
-\tau\expected{\Delta(\infty)}&\!\!\!\!\!\!\!\implies\!\!\!\!\!\!\!&\expected{\Theta(\infty)}\!=\!0. 
-\label{eq:expected_error}
-\end{eqnarray}
-Consequently, the prediction error $\Theta(h)=\gamma(t_h)$ 
-eventually converges to zero as $h$ goes to infinity. This means the WISP tag 
-will estimate the occurrence times of the successive
-events with zero error in expectation.
-
-\subsection{Steady-State (Asymptotic) Synchronization Error}
-\label{sec:app_b}
-An analytical expression for the \emph{asyptotic 
-	synchronization error} can be obtained by calculating 
-$\underset{h\rightarrow \infty}{\lim}\mathrm{Var}(\Theta(h))$. This 
-calculation reduces to find 
-$\mathrm{Var}(\Theta(\infty))=\expected{\Theta(\infty)^2}$, since it holds 
-from 
-\eqref{eq:expected_error} that $\expected{\Theta(\infty)}=0$. Using 
-\eqref{eq:theta}, we 
-get
-\begin{eqnarray}
-\expected{\Theta(\infty)^2} &\!\!\!\!\overset{\eqref{eq:theta}}{=}\!\!\!\!& 
-\tau^2\expected{\Delta(\infty)^2}  
-+\expected{(\Omega(h)+\varepsilon)^2},\label{eq:expected_theta_infinity}
-\end{eqnarray}
-since $\expected{\Delta(\infty)}=0$ from \eqref{eq:delta_convergence} and 
-$\Delta(h)$, $\Omega(h)$ and $\varepsilon$ are independent. 
-In order to calculate \eqref{eq:expected_theta_infinity}, let us derive first 
-$\expected{\Delta(\infty)^2}$ as
-\begin{eqnarray}
-\!\!\!\!\!\expected{\!\Delta(\!\infty\!)^2\!} 
-&\!\!\!\!\!\!\!\overset{\eqref{eq:delta_recursive},\eqref{eq:asymp_delta}}{=}\!\!\!\!\!\!& 
-(\!1\!\!-\!\!\beta\tau\!)^2\!	
-\expected{\!\Delta(\!\infty\!)^2}  
-\!\!+\!\!\expected{\!(\!\Phi(\!h\!)\!\!-\!\!\beta\!(\!\Omega(\!h\!)\!+\!\varepsilon\!)^2}\!. 
-\label{eq:expected_delta_infinity}
-\end{eqnarray}
-In order to calculate \eqref{eq:expected_delta_infinity}, the first step is to 
-obtain the following expectation:
-\begin{eqnarray}
-\expected{(\Phi(h)\!-\!\beta(\Omega(h)\!+\!\varepsilon)^2} 
-&\!\!\!\!=\!\!\!\!&
-\expected{(\Phi(h)^2}
-\!-\!2\beta\expected{\Phi(h)\Omega(h)} 
-\! \nonumber \\ 
-& & +\!\beta^2\expected{\Omega(h)^2}\!+\!\beta^2\expected{\varepsilon^2}. 
-\label{eq:stochastic_expectation}
-\end{eqnarray}
-The equality \eqref{eq:stochastic_expectation} holds due to the fact that 
-$\expected{\Phi(h)\varepsilon}=\expected{\Phi(h)}\expected{\varepsilon}=0$ and 
-$\expected{\varepsilon\Omega(h)}=\expected{\varepsilon}\expected{\Omega(h)}=0$ 
-(since 
-$\expected{\varepsilon}=\expected{\Phi(h)}=\expected{\Omega(h)}=0$ and 
-$\varepsilon$ is independent from $\Phi(h)$ and $\Omega(h)$). The
-following \emph{stochastic integrals} are required for 
-\eqref{eq:stochastic_expectation}:
-\begin{eqnarray}
-\expected{\Phi(h)^2}\!\!\!\!\!\!\! 
-& = & \!\!\!\!\!\!
-\int_{0}^{t_{h+1}-t_h}\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\sigma_{\eta}^{2}dt =  \sigma_{\eta}^{2}(t_{h+1}-t_h),\\
-\expected{\Omega(h)^2}\!\!\!\!\!\!\! 
-& = & \!\!\!\!\!\!
-\frac{\tau^2}{(t_{h+1}\!\!-\!\!t_h)^2}
-\sigma_{\eta}^{2}\!\!\!\int_{0}^{t_{h+1}-t_h}\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!t^2dt\! = \!
-\sigma_{\eta}^{2}\!\frac{\tau^2(\!t_{h+1}\!\!-\!\!t_h\!)}{3},\label{eq:stochastic_integral}\\
-\expected{\Phi(h)\Omega(h)}\!\!\!\!\!\!\! 
-& = & \!\!\!\!\!\!
-\frac{\tau}{t_{h+1}-t_h}\sigma_{\eta}^{2}\int_{0}^{t_{h+1}-t_h}\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!tdt = 
-\sigma_{\eta}^{2}\frac{\tau(\!t_{h+1}\!\!-\!\!t_h\!)}{2}, \\
-\expected{\varepsilon^2}\!\!\!\!\!\!\! 
-& = & \!\!\!\!\!\! 
-\sigma_{\tau}^{2}.
-\end{eqnarray}	
-The derivation of the expectations and stochastic integrals above relies on the 
-assumptions we mentioned 
-at the beginning of this Appendix: (i) $\eta(t')$ and $\eta(t'')$ are 
-uncorrelated for all $t'\neq t''$, therefore $\expected{\eta(t')\eta(t'')}=0$; 
-(ii) since $\expected{\eta(t')}=0$, therefore 
-$\expected{(\eta(t'))^2}=\mathrm{Var}(\eta(t'))=\sigma_{\eta}^2$. Putting these 
-expected 
-values into \eqref{eq:stochastic_expectation}, then in 
-turn into \eqref{eq:expected_delta_infinity} lead:
-\begin{eqnarray}
-\expected{\Delta(\infty)^2}\!\!\!\!\!\!
-& = &\!\!\!\!\!\!
-\frac{\sigma^2_{\eta}(t_{h+1}-t_h)\left(1-\beta\tau+\frac{\beta^2\tau^2}{3}\right)\!\!+\!\!\beta^2\sigma^2_{\tau}}{(2\beta\tau-\beta^2\tau^2)}.
-\label{eq:delta_infinity}
-\end{eqnarray}
-Finally, putting \eqref{eq:delta_infinity} into 
-\eqref{eq:expected_theta_infinity} 
-leads to
-\begin{eqnarray}
-\expected{\Theta(\infty)^2}\!\!\!\!\!\!\! & = & \!\!\!\!\!\!\!
-\frac{\sigma^2_{\eta}(t_{h+1}-t_h)\left(\tau^2
-	-\beta\tau^3-\frac{\beta^2\tau^4}{3}\right)
-	+\beta^2\sigma^2_{\tau}\tau^2}{(2\beta\tau-\beta^2\tau^2)} \nonumber \\ 
-\!\!\!\!\!\!\! &  & \!\!\!\!\!\!\!+ 
-\sigma^2_{\eta}\frac{\tau^2(t_{h+1}-t_h)}{3}+\sigma^2_{\tau}.
-\end{eqnarray}
-Without loss of generality, assume that $t_{h+1}-t_h\approxeq\tau$. Therefore, 
-the asymptotic estimation error of Algorithm \ref{alg:sync} is given by
-\begin{eqnarray}
-Var\left(\underset{h\rightarrow\infty}{\gamma(t_h)}\right)\!\!\!\!\!\!\!& = &\!\!\!\!\!\! 
-\expected{\Theta(\infty)^2}\!\! = \!\!
-\frac{\sigma^2_{\eta}\left(\tau^2-\beta\tau^3
-	+\frac{\beta^2\tau^4}{3}\right)
-	+\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)} \nonumber \\
-\!\!\!\!\!\!\!& = &\!\!\!\!\!\!  +\frac{\sigma^2_{\eta}\tau^3}{3}
-+  \sigma^2_{\tau}
-\end{eqnarray}
-which results in \eqref{eq:sync_performance}.
-
 \ifCLASSOPTIONcaptionsoff
@@
 \bibliography{IEEEabrv,references}
+
+%\input{appendix_tmc}
 

### 2018-04-25 response/response_letter.tex
TMC major revision

\documentclass[10pt]{article}
\usepackage{amsmath,amsfonts,amssymb,graphicx,color,psfrag}
\usepackage[utf8]{inputenc}

\usepackage{vmargin}
\setmarginsrb{2.25cm}{2.25cm}{2.25cm}{2.25cm}
         {12pt}{20pt}{12pt}{36pt}

\usepackage{soul}

\newcommand{\referee}[1]{{\item \color{OliveGreen} \emph{{#1}}}}

\definecolor{OliveGreen}{rgb}{0.1,0.2,0.5}

\newcommand{\response}[1]{{\color{blue} #1}}

\newenvironment{editor}{%
\textbf{\large Comments by the Editor}
\begin{enumerate}%
\renewcommand{\labelenumi}{\textbf{[E:\,\arabic{enumi}]}} %
}{\pagebreak\end{enumerate}}

\newcounter{refereeCounter}

\newenvironment{responses}{%
\refstepcounter{refereeCounter}%
\textbf{\large Comments by Referee \therefereeCounter}
\begin{enumerate}%
\renewcommand{\labelenumi}{\textbf{[R\therefereeCounter :\,\arabic{enumi}]}} %
}{\end{enumerate}}

\setlength{\parindent}{0pt}

\begin{document}

\pagestyle{myheadings}
\thispagestyle{empty}

\markright{\footnotesize COMMAG-16-00166: {\sl Safe and Secure Wireless Power Transfer Networks ($\ldots$)}}

\headsep 0.5cm

\bigskip\bigskip

\noindent{Dr. Tom Alexander\\
Associate Editor\\
IEEE Communications Magazine}

\bigskip\bigskip

\begin{flushright}
Dr. Przemys{\l}aw Pawe{\l}czak \\
Embedded Software Group \\ 
Delft University of Technology \\ 
Mekelweg 4, 2628 CD, Delft, The Netherlands \\
E-mail: p.pawelczak@tudelft.nl\\
Phone: +31 15 278 7491\\
\end{flushright}

\vspace*{2cm}

\today
\medskip

\textbf{COMMAG-16-00166:} {\sl Safe and Secure Wireless Power Transfer Networks: Challenges and Opportunities in RF-Based Systems}

\bigskip

Dear Dr. Alexander,\\

We would like to thank for handling our paper and providing a rapid review results. We are also grateful to the Reviewers for their substantial effort spent for evaluating our work. \\

With this letter we provide our response to all comments of the Reviewers and the revised version of our paper. We look forward to hearing from you at the earliest and we would be happy to provide further clarifications and revisions if necessary. \\

With best regards, \\

in the name of all Authors---Dr. Przemys{\l}aw Pawe{\l}czak

\pagebreak

\begin{editor}
\referee{The authors need to revise the paper accordingly and prepare for the second review. The reviewers mainly concerns in the following aspects. \\
1. Clearly explain the motivation, describe clearly about the problem definition, and describe what new challenges address
in this paper. \\
2. Address the technical details mentioned by Reviewers 1 and 2. \\
3. More results to present mentioned by Reviewer 3.
}

\response{
	Fill
}

\end{editor}

\begin{responses}

\referee{This paper studies the problem of synchronization of computational RFID devices, which is a practically important issue. And the authors propose a lightweight synchronization method and implement it on the WISP tags. I still have the following concerns that require the authors to seriously consider. \\
\\	
In the Introduction part, the authors should clarify what type of synchronization they care. There are two types of synchronization of RFID communication: syn. between reader and tag; syn. among multiple tags. The COTS RFID devices exploit the EPC C1G2 standard, which uses framed slotted Aloha protocol to query tags. We found that the tag reading performance is very well. Hence, I think the synchronization among tags is not a big problem that hinder RFID communication. Hence, this study is not well motivated a least currently. }

\response{
	Fill
}

\referee{Also, there are another view to classify the synchronization problems: slot level synchronization (time frame is divided into slots) or bit level synchronization. Slot level synchronization has been studied for decades and is well addressed. The bit-level synchronization is not well investigated yet and attracts much attention from academic communities, e.g., the following three references. The authors should discuss and clarify whether they study the bit-level synchronization mentioned in these references. \\
**Top-k Queries for Categorized RFID Systems, IEEE/ACM Transactions on Networking (TON), 2017. \\
**Multi-category RFID Estimation, IEEE/ACM Transactions on Networking (TON), 2017. \\
**Completely Pinpointing the Missing RFID Tags in a Time-efficient Way, IEEE Transactions on Computers (TC), 2015.}

\response{
	Fill
}

\referee{What if WISP tags are simultaneously covered by multiple readers. In this case, the readers are not synchronized themselves and may have different time-stamps. Then, how does the proposed method work?}

\response{
	Fill
}

\referee{The novelty of this paper is relatively limited because the proposed event-based reader-tag synchronization method is originated from previous literature. Moreover, the main part of the theoretical analytics is also from previous work [24,31]. Although the authors claimed that they use a different clock drift model in this article, i.e. Brownian motion rather than uniform distribution, I still this significantly reduces the theoretical depth of this paper. The authors should show what NEW challenges beyond previous work are addressed by this paper. }

\response{
	Fill
}

\end{responses}

\begin{responses}

\referee{In this paper, the authors study the time synchronization problem in low-power backscatter systems, i.e., WISP. They point out the unique challenges in designing time synchronization algorithms for these energy-harvesting tags, and propose two synchronization protocols in response to these challenges. The proposed protocol is compatible with the default reader-to-tag communication protocol and consumes an ignorable amount of energy. The authors evaluate their proposed protocols through real-world experiments. \\
\\
+: an interesting problem which has not been explored before. \\
+: the proposed protocols are simple yet look effective. \\
+: evaluating the proposed protocols with real-world experiments; \\
\\		
-: the motivation behind is not well explained; \\
-: not sure the proposed protocol will work in large-scale deployment. \\
\\
My major concern about this paper is its motivation: I do agree with the authors that the time synchronization is important to the multi-hop backscatter networks, but I don’t think it is necessary for the single-hop network like the WISP networks. The ALOHA-type protocol has already synchronized the tags in nature. 
}

\response{
	Fill
}

\referee{I’m not sure synchronizing tags through the command initiated by the 
excitation generator (e.g., RFID reader) is feasible in real-world applications. This is because these energy-harvesting sensors can be charged within a very short range. Hence if the sensors are deployed sparsely, multiple excitation generators are needed, and it is thus important to synchronize these excitation generators as well. However, the authors did not consider this point.
}

\response{
	Fill
}

\referee{Some questions about the technical parts: \\
1): the authors claim that the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant clock drift in WISP. However, as far as I know, WISP uses a regulator to supply a constant 1.8V DC to power the WISP.}

\response{
	Fill
}

\referee{2): section 5.1, the authors observe a 1.89ms transmission delay. What does this delay account for? the in-air time should not be so long.
}

\response{
	Fill
}

\end{responses}

\begin{responses}

\referee{The authors have investigated the synchronization problem of RFID networks which consists of RFID readers and WISP tags. Several challenges of existing hardware platform and systems are identified and discussed. An event-based synchronization approach is proposed for solving the problem. A real demo system is implemented and used to evaluate the effectiveness of the proposed approach. The results show that the error of synchronization can be about 1ms and the energy overhead is about 1.85uJ. \\
\\
Generally, this paper is well written. It is a novel idea to improve the synchronization ability of RFID systems. Considering the emerging of internet of things, this work  is attractive.  The original version of this paper is published in a conference and this version includes sufficient additional contents for the journal. Therefore, I recommend this paper to be accepted after some revisions as follows:  \\
\\
1) It is better to simplify the descriptions of the challenges and existing works, especially some contents that are not directly related to the synchronization.}

\response{
	Fill
}

\referee{2) It is better to report some results or discuss the performance of the proposed approach in the multi-hop networks, since the authors highlights the advantages of  multi-hop networks several times in the introduction.}

\response{
	Fill
}

\referee{3) The authors may consider to discuss a specific scenario that can use the proposed approach.}

\response{
	Fill
}

\end{responses}

\end{document}

### 2018-05-03 response/response_letter.tex
responses

@@
 \usepackage{soul}
+\usepackage{hyperref}
 
-\newcommand{\referee}[1]{{\item \color{OliveGreen} \emph{{#1}}}}
+\newcommand{\referee}[1]{
+	{\item \color{OliveGreen} \emph{{#1}}}
+	\label{R\therefereeCounter:\arabic{enumi}}
+}
 
@@
 Hence, I think the synchronization among tags is not a big problem that hinder RFID communication.
-Hence, this study is not well motivated a least currently.
-}
+Hence, this study is not well motivated a least currently.}
 
@@
 \response{
-	Fill
+Thanks to these comments, we noticed that we did not highlight the main difference of our work from \emph{bit-level synchronization} in the previous submission.
+Moreover, we thank the reviewer for providing the studies we have not cited in the previous submission.\newline
+As indicated by the reviewer, it is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal---i.e.
+bit-level synchronization.
+However, the objective of our work has the following fundamental differences as compared to these studies:
+\begin{itemize}
+	\item \textbf{Portability:} Bit-level synchronization relies on high-precision clock to obtain microsecond-precision timing measurements.
+Despite, in our work, we used we used the external 32\,KHz clock that can run in low-power standby mode and the \emph{higher-level} commands of the EPC-Gen 2 standard, rather than exploiting its low-level signal properties.
+Therefore, our approach is independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.
+	\item \textbf{Explicit Clock:} Bit-level synchronization does not output an explicit clock value and it does not compensate for the clock drift.
+Rather, it uses offset compensation only and \emph{guard time} to reduce the effect of the clock drift.
+On the contrary, we provided time synchronization algorithms that compensate for the clock drift and provide an \emph{explicit} common time notion to CRFID applications.
+	\item \textbf{Energy Budget:} Due to marginal energy budgets, we proposed synchronization algorithms that require minimum number of computation steps and memory space.
+	\item \textbf{Power-Failures:} Providing synchronization service to the application layer requires to cope with power failures---which has not been investigated so far.
+To address this issue, we provided synchronization algorithms that can recover easily after power failures.
+\end{itemize}
+
+We would like to emphasize that none of these issues have been addressed and studied so far within the context of CRFIDs.
 }
@@
 \response{
-	Fill
+As we indicated in the introduction of the article, there are synchronization techniques that provide synchronization among RFID readers.
+Therefore, if there are multiple readers that cover multiple WISP tags, first the readers should be synchronized with each other and then the tags will be synchronized to the RFID readers.
+We mentioned these issues in {\color{red} Section XXX} of our revised paper.
 }
@@
 \response{
-	Fill
+	See our responses for \hyperref[R1:2]{[R1:2]}.
+	See our responses for \hyperref[R1:3]{R1:3}.
 }

### 2018-05-03 TMC-2017-11-0695-main.tex
TMC major revision
responses

--- 2017-11-01 manuscript_tmc.tex
@@
 However, such techniques require a calibration phase in order to build a temperature--clock frequency table.
+\TODO{Cite **Top-k Queries for Categorized RFID Systems, IEEE/ACM Transactions on Networking (TON), 2017. \\
+	**Multi-category RFID Estimation, IEEE/ACM Transactions on Networking (TON), 2017. \\
+	**Completely Pinpointing the Missing RFID Tags in a Time-efficient Way, IEEE Transactions on Computers (TC), 2015.}
 
@@
 In this study, we used the higher-level properties of the EPC Gen2 standard, rather than exploiting its low-level signal properties.
-By means of this, we could  use the external 32\,KHz clock that runs even in low-power standby mode (see Section \ref{sec:Clock_Hardware}) to obtain timing measurements.} Fig.~\ref{fig:event_based} presents the steps of event-based synchronization.
+By means of this, we could  use the external 32\,KHz clock that runs even in low-power standby mode (see Section \ref{sec:Clock_Hardware}) to obtain timing measurements.)} Fig.~\ref{fig:event_based} presents the steps of event-based synchronization.
 During the command phase, the EPC Gen2 standard allows a maximum of eight successive \emph{BlockWrite} operations.
@@
 
-%\input{appendix_tmc}
+\input{appendix_tmc}
 

### 2018-05-03 response/response_letter.tex
responses

@@
 \response{
-	Fill
+	We believe that we revised our article in accordance with the responses of the reviewers. Please see our responses. 
 }
@@
 \response{
-	Fill
+	Please also see \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}.
 }
@@
 Therefore, our approach is independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.
+However, our synchronization algorithms can also be integrated to bit-level synchronization.
 	\item \textbf{Explicit Clock:} Bit-level synchronization does not output an explicit clock value and it does not compensate for the clock drift.
@@
 \response{
-	See our responses for \hyperref[R1:2]{[R1:2]}.
-	See our responses for \hyperref[R1:3]{R1:3}.
+As we indicated in our previous responses (see \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}), our article has the following novelties over the existing work in the literature:
+\begin{itemize}
+	\item Bit-level synchronization does not compensate for clock drift and it does not provide explicit time notion to the tags.
+In order to output a synchronized clock value, a time synchronization algorithm is required.
+	\item Since the energy budget of the tags are limited, synchronization approaches that require marginal number of computation and memory.
+This is challenging---the main motivation for applying a PI-controller based  synchronization for the first time in this domain.
+	\item Maintaining an explicit synchronized clock despite power failures has never been explored before in the field of CRFIDs.
+We also provided how our synchronization techniques overcome frequent power failures and provide a stable synchronized clock value.
+	\item  We also provided an in-depth theoretical analysis of the proposed synchronization techniques.
+As indicate by the reviewer, the main part of the analysis is from previous work.
+However, the previous analysis that assumes uniform distribution for the clock drifts was not realistic---the reason we changed the analysis to use Brownian motion.
+This modification did not change the proof structure but it significantly changed the individual steps of the proof---that led to different stead-state error expressions.
+\end{itemize}
 }
@@
 \response{
-	Fill
+Since we used high-level properties of EPC Gen2 protocol, 
 }

### 2018-05-04 response/response_letter.tex
reviewer 1

@@
 
-\markright{\footnotesize COMMAG-16-00166: {\sl Safe and Secure Wireless Power Transfer Networks ($\ldots$)}}
+\markright{\footnotesize TMC-2017-11-0695: {\sl On the Synchronization of Computational RFIDs}}
 
@@
 
-\noindent{Dr. Tom Alexander\\
-Associate Editor\\
-IEEE Communications Magazine}
+%\noindent{Marwan Krunz, EIC, and Kevin Almeroth, Associate EIC \\
+%	IEEE Transactions on Mobile Computing
+%}
 
@@
 
-\textbf{COMMAG-16-00166:} {\sl Safe and Secure Wireless Power Transfer Networks: Challenges and Opportunities in RF-Based Systems}
+\textbf{TMC-2017-11-0695:} {\sl On the Synchronization of Computational RFIDs}
 
@@
 
-Dear Dr. Alexander,\\
+Dear Editor,\\
 
@@
 \response{
-	Please also see \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}.
+We thank the reviewer for pointing this issue that led us to emphasize our motivation in the revised article.
+As indicated by the reviewer, framed slotted Aloha protocol performs a notion of synchronization: by adjusting offset and introducing guard times, tags are ensured to communicate at allocated frame boundaries.
+However,
+\begin{itemize}
+	\item this method does not output a synchronized clock value that can be accessed by the tags to timestamp and make inferences about the collected information;
+	\item since a clock is not maintained, more sophisticated applications cannot be enabled.
+\end{itemize}
+
+In order to enable sensor network-like data collection applications, tags should maintain an explicit clock that represents synchronized notion of time so that they can  collect, timestamp and store data which will be be queried later by the readers.
+However, due to frequent power failures and marginal energy budgets of these devices, providing explicit clock synchronization and maintaining a clock value is challenging---please also see our responses \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}.
+We mentioned all these points in our revised article.
 }

### 2018-05-04 response/response_letter.tex
reviewer 2

@@
 \response{ 
-	Fill
+ The concern of this reviewer has been shared by also Reviewer 1. In our revised article, we provided a better motivation and highlighted the novelty of our work better.  \\
+ In the previous responses---please also see \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}---we mentioned that ALOHA-type protocol does not maintain an explicit clock value and does not implement a real synchronization algorithm.
+On the contrary, we provided novel synchronization mechanisms that enables tags to access synchronized time---this also enables sensor network-like data collection applications where tags collect and timestamp sensor readings that can be queried by the readers later.
+However, maintaining an explicit clock and ensuring synchronization is not trivial due to the marginal energy budgets of the tags and frequent power failures.
+These are the main challenges of our work and its novelty over existing studies.
 }
@@
 \response{
-	Fill
+We thank the reviewer for raising this issue.
+We already answered this concern in the previous responses---please also see \hyperref[R1:3]{[R1:3]}.
+We would like to mention that there are already prior work that provides synchronization among the readers.
+Therefore, we can also employ these strategies in order to create a common time notion among the readers.
+In order to provide an explicit synchronized clock value to the WISP tags, an explicit reader-tag synchronization algorithm is required---which is the focus of our work.
 }
@@
 \response{
-Since we used high-level properties of EPC Gen2 protocol, 
+In our article, we defined the \emph{transmission delay} as the time that passes between the start of the broadcast command from the reader and the receipt by the receiver node--- which is the major error source and composed of non-deterministic components.
+Since we used high-level properties of EPC Gen2 protocol, we could not use low-level timestamping mechanisms and our synchronization approaches suffer from big transmission delays.
+The reason is that we do not know when the reader starts transmission after broadcast command is requested from the PC that controls the reader.
+During this process, the reader-side does not implement any timestamping mechanism and the delays are mainly dependent on the reader-side implementation of the EPC protocol.
 }
@@
 \response{
-	Fill
+	We would like to thank the reviewer for his/her encouraging comments.
 }
@@
 \response{
-	Fill
+	We provided a discussion 
 }

### 2018-05-07 response/response_letter.tex
small corrections (testing git commit)

@@
 \response{
-	We believe that we revised our article in accordance with the responses of the reviewers. Please see our responses. 
+	We believe that we revised our article in accordance with the responses of the reviewers. Please refer to our respective responses below. 
 }
@@
 
-\referee{This paper studies the problem of synchronization of computational RFID devices, which is a practically important issue.
-And the authors propose a lightweight synchronization method and implement it on the WISP tags.
-I still have the following concerns that require the authors to seriously consider.
-\\
-\\	
-In the Introduction part, the authors should clarify what type of synchronization they care.
+\referee{In the Introduction part, the authors should clarify what type of synchronization they care.
 There are two types of synchronization of RFID communication: syn.
@@
 The authors should discuss and clarify whether they study the bit-level synchronization mentioned in these references.
-\\
-**Top-k Queries for Categorized RFID Systems, IEEE/ACM Transactions on Networking (TON), 2017. \\
-**Multi-category RFID Estimation, IEEE/ACM Transactions on Networking (TON), 2017. \\
-**Completely Pinpointing the Missing RFID Tags in a Time-efficient Way, IEEE Transactions on Computers (TC), 2015.}
+\begin{itemize}
+\item \textbf{Top-k Queries for Categorized RFID Systems}, IEEE/ACM Transactions on Networking, 2017;
+\item \textbf{Multi-category RFID Estimation}, IEEE/ACM Transactions on Networking, 2017;
+\item \textbf{Completely Pinpointing the Missing RFID Tags in a Time-efficient Way}, IEEE Transactions on Computers, 2015.
+\end{itemize}}
 
@@
 Thanks to these comments, we noticed that we did not highlight the main difference of our work from \emph{bit-level synchronization} in the previous submission.
-Moreover, we thank the reviewer for providing the studies we have not cited in the previous submission.\newline
+Moreover, we thank the reviewer for providing the studies we have not cited in the previous submission---we have added these papers to our related work list.
+
 As indicated by the reviewer, it is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal---i.e.

### 2018-05-07 response/response_letter.tex
Reviewer 3: only comments are left

@@
 1): the authors claim that the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant clock drift in WISP.
-However, as far as I know, WISP uses a regulator to supply a constant 1.8V DC to power the WISP.}
+However, as far as I know, WISP uses a regulator to supply a constant 1.8\,V DC to power the WISP.}
 
@@
 
-\referee{2): section 5.1, the authors observe a 1.89ms transmission delay. What does this delay account for? the in-air time should not be so long.
+\referee{2): section 5.1, the authors observe a 1.89\,ms transmission delay. What does this delay account for? the in-air time should not be so long.
 }
@@
 
-\referee{The authors have investigated the synchronization problem of RFID networks which consists of RFID readers and WISP tags.
-Several challenges of existing hardware platform and systems are identified and discussed.
-An event-based synchronization approach is proposed for solving the problem.
-A real demo system is implemented and used to evaluate the effectiveness of the proposed approach.
-The results show that the error of synchronization can be about 1ms and the energy overhead is about 1.85uJ.
-\\
-\\
-Generally, this paper is well written.
-It is a novel idea to improve the synchronization ability of RFID systems.
-Considering the emerging of internet of things, this work  is attractive.
-The original version of this paper is published in a conference and this version includes sufficient additional contents for the journal.
-Therefore, I recommend this paper to be accepted after some revisions as follows:  \\
-\\
-1) It is better to simplify the descriptions of the challenges and existing works, especially some contents that are not directly related to the synchronization.}
-
-\response{
-	We would like to thank the reviewer for his/her encouraging comments.
-}
-
-\referee{2) It is better to report some results or discuss the performance of the proposed approach in the multi-hop networks, since the authors highlights the advantages of  multi-hop networks several times in the introduction.}
+\referee{It is better to simplify the descriptions of the challenges and existing works, especially some contents that are not directly related to the synchronization.}
 
@@
 
-\referee{3) The authors may consider to discuss a specific scenario that can use the proposed approach.}
+\referee{It is better to report some results or discuss the performance of the proposed approach in the multi-hop networks, since the authors highlights the advantages of  multi-hop networks several times in the introduction.}
 
 \response{
-	We provided a discussion 
+	Fill
+}
+
+\referee{The authors may consider to discuss a specific scenario that can use the proposed approach.}
+
+\response{
+	Fill
 }

### 2018-05-07 response/response_letter.tex
Reviewer 3 drafted: needs text in the paper
Comments to the editor revised

@@
 \begin{editor}
-\referee{The authors need to revise the paper accordingly and prepare for the second review. The reviewers mainly concerns in the following aspects. \\
-1. Clearly explain the motivation, describe clearly about the problem definition, and describe what new challenges address
-in this paper. \\
-2. Address the technical details mentioned by Reviewers 1 and 2. \\
-3. More results to present mentioned by Reviewer 3.
+\referee{The authors need to revise the paper accordingly and prepare for the second review. The reviewers mainly concerns in the following aspects.
+	\begin{enumerate}
+		\item Clearly explain the motivation, describe clearly about the problem definition, and describe what new challenges address in this paper;
+		\item Address the technical details mentioned by Reviewers 1 and 2;
+		\item More results to present mentioned by Reviewer 3.
+\end{enumerate}
 }
@@
 \response{
-	We believe that we revised our article in accordance with the responses of the reviewers. Please refer to our respective responses below. 
+	Thank you once more for coordinating the review process of our paper.
+We hope that we have revised our article in accordance with the responses of the reviewers.
+We therefore believe we have addressed all three points that the Editor raised as accurately as possible.
+We request the Editor to refer to our respective responses below.
 }
@@
 \response{
-	Fill
+	To address the comment of the reviewer we have reduced the length of the description in Section 2 of the paper. \textcolor{red}{Compress text in Section 2}
 }
@@
 \response{
-	Fill
+	Fill \textcolor{red}{Matlab simulations for the multi-hop scenario.}
 }
@@
 \response{
-	Fill
+	Fill \textcolor{red}{Discuss a scenario for our method}
 }

### 2018-05-07 response/response_letter.tex
R1.1 revised

@@
 We thank the reviewer for pointing this issue that led us to emphasize our motivation in the revised article.
-As indicated by the reviewer, framed slotted Aloha protocol performs a notion of synchronization: by adjusting offset and introducing guard times, tags are ensured to communicate at allocated frame boundaries.
+As indicated by the reviewer, framed slotted Aloha protocol indeed performs a form of synchronization: by adjusting offset and introducing guard times, tags are ensured to communicate at the allocated frame boundaries.
 However,
+
 \begin{itemize}
-	\item this method does not output a synchronized clock value that can be accessed by the tags to timestamp and make inferences about the collected information;
-	\item since a clock is not maintained, more sophisticated applications cannot be enabled.
+	\item this method \emph{does not output a synchronized clock value} that can be accessed by the tags to timestamp (and make inference about) the collected information, e.g.
+from the sensors attached to the CRFID;
+	\item in consequence, \emph{applications requiring time notion cannot be enabled}, since a clock is not maintained at the CRFID tags.
 \end{itemize}
 
-In order to enable sensor network-like data collection applications, tags should maintain an explicit clock that represents synchronized notion of time so that they can  collect, timestamp and store data which will be be queried later by the readers.
-However, due to frequent power failures and marginal energy budgets of these devices, providing explicit clock synchronization and maintaining a clock value is challenging---please also see our responses \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}.
-We mentioned all these points in our revised article.
+Unfortunately due to frequent power failures and marginal energy budgets of CRFIDs, providing explicit clock synchronization (that represents \emph{synchronized notion of time}) and maintaining this clock value is challenging.
+We have emphasized these points in our revised article in Section 1.1 and Section 2.1.3.
+We hope that with these answers we have motivated the need for our study sufficiently.
+We kindly refer the reviewer also to our responses \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}.
+\textcolor{red}{Add clarification on the difference between clock synchronization with Aloha and our study in Section 1.1.
+and 2.1.3}
 }

### 2018-05-07 response/response_letter.tex
R1.2 and R1.3 revised

@@
 We request the Editor to refer to our respective responses below.
+\textcolor{red}{Add direct references where each of these three points has been addressed.}
 }
@@
 The authors should discuss and clarify whether they study the bit-level synchronization mentioned in these references.
-\begin{itemize}
-\item \textbf{Top-k Queries for Categorized RFID Systems}, IEEE/ACM Transactions on Networking, 2017;
-\item \textbf{Multi-category RFID Estimation}, IEEE/ACM Transactions on Networking, 2017;
-\item \textbf{Completely Pinpointing the Missing RFID Tags in a Time-efficient Way}, IEEE Transactions on Computers, 2015.
-\end{itemize}}
+}
 
 \response{
-Thanks to these comments, we noticed that we did not highlight the main difference of our work from \emph{bit-level synchronization} in the previous submission.
-Moreover, we thank the reviewer for providing the studies we have not cited in the previous submission---we have added these papers to our related work list.
+We thank the reviewer for this comments and for pointing us to these relevant papers, which we cite in Section 2.1.3 of our revised work.
+Thanks to the comment of the reviewer we have noticed that in the previous version of our submission we did not highlighted the main difference between our work and the works pertaining to the \emph{bit-level synchronization}.
 
-As indicated by the reviewer, it is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal---i.e.
+To clarify this difference: as indicated earlier by the reviewer, it is possible to exploit modulation properties of signals from the reader to the tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal, i.e.
 bit-level synchronization.
 However, the objective of our work has the following fundamental differences as compared to these studies:
+
 \begin{itemize}
 	\item \textbf{Portability:} Bit-level synchronization relies on high-precision clock to obtain microsecond-precision timing measurements.
-Despite, in our work, we used we used the external 32\,KHz clock that can run in low-power standby mode and the \emph{higher-level} commands of the EPC-Gen 2 standard, rather than exploiting its low-level signal properties.
+Despite, in our work, we used the external 32\,kHz clock that can run in low-power standby mode and the \emph{higher-level} commands of the EPC Gen 2 standard, rather than exploiting its low-level signal properties.
 Therefore, our approach is independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.
-However, our synchronization algorithms can also be integrated to bit-level synchronization.
+However, our synchronization algorithms can also be integrated into the bit-level synchronization;
+
 	\item \textbf{Explicit Clock:} Bit-level synchronization does not output an explicit clock value and it does not compensate for the clock drift.
 Rather, it uses offset compensation only and \emph{guard time} to reduce the effect of the clock drift.
-On the contrary, we provided time synchronization algorithms that compensate for the clock drift and provide an \emph{explicit} common time notion to CRFID applications.
-	\item \textbf{Energy Budget:} Due to marginal energy budgets, we proposed synchronization algorithms that require minimum number of computation steps and memory space.
-	\item \textbf{Power-Failures:} Providing synchronization service to the application layer requires to cope with power failures---which has not been investigated so far.
+On the contrary, we provided time synchronization algorithms that compensate for the clock drift and provide an \emph{explicit} common time notion to CRFID applications;
+
+	\item \textbf{Energy Budget:} Due to marginal energy budgets, we proposed synchronization algorithms that require minimum number of computation steps and memory space;
+
+	\item \textbf{Power Failures:} Providing synchronization service to the application layer requires to cope with power failures, which has not been investigated so far.
 To address this issue, we provided synchronization algorithms that can recover easily after power failures.
@@
 We would like to emphasize that none of these issues have been addressed and studied so far within the context of CRFIDs.
+The above discussion has been briefly added to Section 2.1.3.
+\textcolor{red}{Add brief extension on clock synchronization to Section 2.1.3}
 }
@@
 \response{
-As we indicated in the introduction of the article, there are synchronization techniques that provide synchronization among RFID readers.
-Therefore, if there are multiple readers that cover multiple WISP tags, first the readers should be synchronized with each other and then the tags will be synchronized to the RFID readers.
-We mentioned these issues in {\color{red} Section XXX} of our revised paper.
+As we indicated in the introduction of the article (Section 1) there are already existing synchronization techniques that provide synchronization among RFID readers.
+Therefore, if there are multiple readers that cover multiple WISP tags, first the readers should be synchronize with each other and then the tags will be synchronized to the RFID readers.
+This approach will cause any danger to the protocol we designed.
+We mentioned this explicitly in Section 4 of our revised paper.
+\textcolor{red}{Add clarification about single reader synchronization}
 }

### 2018-05-09 response/response_letter.tex
responses to R3

@@
 \response{
-	Fill
+The voltage regulators exhibit instability and inaccuracy with respect to the input voltage; i.e.
+they are not perfect and they have an input voltage transient response.
+Therefore, they have an output voltage variation with respect to the input voltage---leading to clock frequency instabilities.
 }
@@
 \response{
-	To address the comment of the reviewer we have reduced the length of the description in Section 2 of the paper. \textcolor{red}{Compress text in Section 2}
+	This issue has been raised by also reviewers 1 and 2.
+To address the comments of the reviewers, we have reduced the length of the description in Section 2 of the paper.
+Moreover, we provided a better motivation and new challenges (and related work) in the revised paper.
+\textcolor{red}{Compress text in Section 2}
 }
@@
 \response{
-	Fill \textcolor{red}{Discuss a scenario for our method}
+	We thank the reviewer for these comment that led us to provide a better motivation for our study.
+By also considering the comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we provided a specific scenario where the proposed synchronization techniques can be used.
+In the introduction of the article, we depicted the case where several WISP tags are powered using a reader.
+In this scenario, the tags collaboratively timestamp their temperature readings.
+When the reader requests their temperature readings, the tags also transmit their timestamps together with the collected temperature data.
+In order to make inferences on the collected temperature readings, timestamps should be consistent so that the reader can provide a temporal ordering of the readings.
+Therefore, tags require a common time notion to timestamp their readers---creating an demand for explicit time synchronization.
 }

### 2018-05-09 response/response_letter.tex
responses to R3

@@
 \response{
-	Fill \textcolor{red}{Matlab simulations for the multi-hop scenario.}
+There is a misunderstanding regarding the concept of multi-hop networks.
+As of now, CRFID tags cannot operate in a multi-hop manner since they can only communicate directly with the RFID readers.
+In the introduction of the article, we anticipated that after enabling tag-to-tag backscatter communication, multi-hop CRFID networks will emerge and therefore time synchronization will be more important.
+However, as of now, we do not have any network infrastructure to demonstrate neither multi-hop communication nor synchronization.
+The thing that make sense is that if there are several RFID readers and that cover several tags simultaneously, the whole network can be synchronized by first allowing RFID readers to access central time, e.g.
+using internet; then individual tags can be synchronized to the RFID readers.
+As suggested by the reviewer, we provided a discussion for this concern in the revised article.
+
+ \textcolor{red}{add a reader-network-tag sync. discussion}
 }

### 2018-05-10 response/response_letter.tex
R3.1 revised

@@
 The above discussion has been briefly added to Section 2.1.3.
-\textcolor{red}{Add brief extension on clock synchronization to Section 2.1.3}
+\textcolor{red}{Add brief extension on clock synchronization to Section 2.1.3; add references they list explicitly}
 }
@@
 \response{
-	This issue has been raised by also reviewers 1 and 2.
-To address the comments of the reviewers, we have reduced the length of the description in Section 2 of the paper.
-Moreover, we provided a better motivation and new challenges (and related work) in the revised paper.
+	To address this comment of the reviewer, we have removed unnecessary description from Section 2 of our paper.
+At the same time, in this section we have provided a better motivation and listed novelties of our contribution in the context of new related work.
 \textcolor{red}{Compress text in Section 2}

### 2018-05-10 response/response_letter.tex
R3.2 revised

@@
 \response{
-There is a misunderstanding regarding the concept of multi-hop networks.
+We thank the reviewer for this comment.
+However, we feel that there is a misunderstanding regarding the concept of multi-hop networks.
 As of now, CRFID tags cannot operate in a multi-hop manner since they can only communicate directly with the RFID readers.
-In the introduction of the article, we anticipated that after enabling tag-to-tag backscatter communication, multi-hop CRFID networks will emerge and therefore time synchronization will be more important.
-However, as of now, we do not have any network infrastructure to demonstrate neither multi-hop communication nor synchronization.
+In the introduction of our article (Section 1.1), we stated that we \emph{anticipate} that after enabling tag-to-tag backscatter communication, multi-hop CRFID networks will \emph{emerge} and therefore time synchronization will be more important.
+However, as of now, we do not have access to (nor we are not aware of) any network infrastructure to demonstrate neither multi-hop communication nor synchronization.
 The thing that make sense is that if there are several RFID readers and that cover several tags simultaneously, the whole network can be synchronized by first allowing RFID readers to access central time, e.g.
@@
 
- \textcolor{red}{add a reader-network-tag sync. discussion}
+ \textcolor{red}{add a reader-network-tag synchronization discussion in Section }
 }

### 2018-05-10 response/response_letter.tex
R3.2 revised

--- 2018-05-10 response/response_letter.tex
@@
 \response{
-There is a misunderstanding regarding the concept of multi-hop networks.
+We thank the reviewer for this comment.
+However, we feel that there has been a misunderstanding regarding the concept of backscatter (CRFID) multi-hop networks.
 As of now, CRFID tags cannot operate in a multi-hop manner since they can only communicate directly with the RFID readers.
-In the introduction of the article, we anticipated that after enabling tag-to-tag backscatter communication, multi-hop CRFID networks will emerge and therefore time synchronization will be more important.
-However, as of now, we do not have any network infrastructure to demonstrate neither multi-hop communication nor synchronization.
-The thing that make sense is that if there are several RFID readers and that cover several tags simultaneously, the whole network can be synchronized by first allowing RFID readers to access central time, e.g.
-using internet; then individual tags can be synchronized to the RFID readers.
-As suggested by the reviewer, we provided a discussion for this concern in the revised article.
+In the introduction of our article (Section 1.1), we only stated that we \emph{anticipate} that after enabling tag-to-tag backscatter communication, multi-hop CRFID networks will \emph{emerge} and therefore time synchronization will be more important.
+However, as of now, we do not have access to (nor we are not aware of) any network infrastructure to demonstrate neither multi-hop communication nor multi-hop synchronization.
 
- \textcolor{red}{add a reader-network-tag sync. discussion}
+Our work applies to the case when several RFID readers cover large tag population simultaneously.
+Then the whole network can be synchronized by first allowing RFID readers to access central reference time, e.g.
+using the Internet; then individual tags can be synchronized to the individual RFID readers.
+To clarify the connection of our work to the multi-hop networks, we have provided an extra clarifying sentence in Section 1.1.
+
+ \textcolor{red}{add a sentence on reader-network-tag synchronization in Section 1.1}
 }

### 2018-05-10 response/response_letter.tex
R3.3 revised

@@
 \response{
-	We thank the reviewer for these comment that led us to provide a better motivation for our study.
+	We thank the reviewer for this comment that led us to provide a better motivation for our study.
 By also considering the comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we provided a specific scenario where the proposed synchronization techniques can be used.
-In the introduction of the article, we depicted the case where several WISP tags are powered using a reader.
+In the introduction of the article, we depicted the case where several WISPs  are powered by a reader.
 In this scenario, the tags collaboratively timestamp their temperature readings.
 When the reader requests their temperature readings, the tags also transmit their timestamps together with the collected temperature data.
-In order to make inferences on the collected temperature readings, timestamps should be consistent so that the reader can provide a temporal ordering of the readings.
-Therefore, tags require a common time notion to timestamp their readers---creating an demand for explicit time synchronization.
+In order to make inferences on the collected temperature readings, timestamps should be \emph{consistent} so that the reader can provide a temporal ordering of the readings.
+Therefore, tags require a common time notion to timestamp their readers---creating an demand for explicit time synchronization, which we address in our work.
 }

### 2018-05-10 response/response_letter.tex
R3.3 revised

@@
 	We thank the reviewer for this comment that led us to provide a better motivation for our study.
-By also considering the comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we provided a specific scenario where the proposed synchronization techniques can be used.
+Already, in reply to comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we provided a specific scenario where the proposed synchronization techniques can be used.
 In the introduction of the article, we depicted the case where several WISPs  are powered by a reader.

### 2018-05-10 response/response_letter.tex
R3.3 revised

@@
 	We thank the reviewer for this comment that led us to provide a better motivation for our study.
-Already, in reply to comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we provided a specific scenario where the proposed synchronization techniques can be used.
+Already, in reply to comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we commented on cases where the proposed synchronization techniques can be used.
 In the introduction of the article, we depicted the case where several WISPs  are powered by a reader.

### 2018-05-10 response/response_letter.tex
R1.4 revised

@@
 \response{
-As we indicated in our previous responses (see \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}), our article has the following novelties over the existing work in the literature:
-\begin{itemize}
-	\item Bit-level synchronization does not compensate for clock drift and it does not provide explicit time notion to the tags.
-In order to output a synchronized clock value, a time synchronization algorithm is required.
-	\item Since the energy budget of the tags are limited, synchronization approaches that require marginal number of computation and memory.
-This is challenging---the main motivation for applying a PI-controller based synchronization for the first time in this domain.
-	\item Maintaining an explicit synchronized clock despite power failures has never been explored before in the field of CRFIDs.
-We also provided how our synchronization techniques overcome frequent power failures and provide a stable synchronized clock value.
-	\item  We also provided an in-depth theoretical analysis of the proposed synchronization techniques.
-As indicate by the reviewer, the main part of the analysis is from previous work.
-However, the previous analysis that assumes uniform distribution for the clock drifts was not realistic---the reason we changed the analysis to use Brownian motion.
-This modification did not change the proof structure but it significantly changed the individual steps of the proof---that led to different stead-state error expressions.
-\end{itemize}
+As we indicated in our previous responses (see \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}), our article has the following novelties over the previous results:
+\begin{enumerate}
+	\item \textbf{Difference with bit-level synchronization:} Bit-level synchronization does not compensate for clock drift and does not provide explicit time notion to the tags.
+In order to output a synchronized clock value, a time synchronization algorithm is required;
+	\item \textbf{Energy budget of synchronization algorithm:} Since the energy budget of the tags are limited, synchronization approaches require marginal number of computation and memory.
+This is challenging, which was our main motivation for applying a PI-controller based synchronization for the first time in this domain;
+	\item \textbf{Consideration of intermittent power supply:} Maintaining an explicit synchronized clock despite power failures has never been explored before in the field of CRFIDs.
+We also provided how our synchronization technique overcomes frequent power failures and provide a stable synchronized clock value;
+	\item \textbf{Novel theoretical consideration:} We also provided an in-depth theoretical analysis of the proposed synchronization techniques.
+As indicate by the reviewer, the main part of the analysis is indeed from previous work.
+However, the previous analysis that assumes uniform distribution for the clock drifts was not realistic---the reason we changed the analysis to use the Brownian motion.
+This modification did not change the proof structure but it significantly changed the individual steps of the proof---that led to \emph{different steady-state error expressions}.
+\end{enumerate}
 }
@@
 	We thank the reviewer for this comment that led us to provide a better motivation for our study.
-Already, in reply to comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we commented on cases where the proposed synchronization techniques can be used.
-In the introduction of the article, we depicted the case where several WISPs  are powered by a reader.
+Already, in reply to comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we commented on cases where our proposed synchronization technique can be used.
+To recap our answer: in the introduction of our article, we depicted the case where several WISPs  are powered by a reader.
 In this scenario, the tags collaboratively timestamp their temperature readings.

### 2018-05-10 response/response_letter.tex
R1.3 revised

@@
 As we indicated in the introduction of the article (Section 1) there are already existing synchronization techniques that provide synchronization among RFID readers.
-Therefore, if there are multiple readers that cover multiple WISP tags, first the readers should be synchronize with each other and then the tags will be synchronized to the RFID readers.
+Therefore, if there are multiple readers that cover multiple WISP tags, first the readers should be synchronized with each other and then the tags will be synchronized to the RFID readers.
 This approach will cause any danger to the protocol we designed.
 We mentioned this explicitly in Section 4 of our revised paper.
-\textcolor{red}{Add clarification about single reader synchronization}
+\textcolor{red}{Add clarification about single reader synchronization in Section 4}
 }

### 2018-05-10 response/response_letter.tex
R2.1 revised

@@
 
-\referee{In this paper, the authors study the time synchronization problem in low-power backscatter systems, i.e., WISP.
-They point out the unique challenges in designing time synchronization algorithms for these energy-harvesting tags, and propose two synchronization protocols in response to these challenges.
-The proposed protocol is compatible with the default reader-to-tag communication protocol and consumes an ignorable amount of energy.
-The authors evaluate their proposed protocols through real-world experiments.
-\\
-\\
-+: an interesting problem which has not been explored before. \\
-+: the proposed protocols are simple yet look effective. \\
-+: evaluating the proposed protocols with real-world experiments; \\
-\\		
--: the motivation behind is not well explained; \\
--: not sure the proposed protocol will work in large-scale deployment. \\
-\\
-My major concern about this paper is its motivation: I do agree with the authors that the time synchronization is important to the multi-hop backscatter networks, but I don’t think it is necessary for the single-hop network like the WISP networks.
-The ALOHA-type protocol has already synchronized the tags in nature.
-}
+\referee{My major concern about this paper is its motivation: I do agree with the authors that the time synchronization is important to the multi-hop backscatter networks, but I don’t think it is necessary for the single-hop network like the WISP networks.
+The ALOHA-type protocol has already synchronized the tags in nature.}
 
-\response{ 
- The concern of this reviewer has been shared by also Reviewer 1. In our revised article, we provided a better motivation and highlighted the novelty of our work better.  \\
- In the previous responses---please also see \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}---we mentioned that ALOHA-type protocol does not maintain an explicit clock value and does not implement a real synchronization algorithm.
+\response{In our revised article, we provided a better motivation and highlighted the novelty of our work.
+In our earlier response to the same questions raised by reviewer 1---please refer to \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}---we mentioned that ALOHA-type protocol does not maintain an explicit clock value and does not implement a real synchronization algorithm.
 On the contrary, we provided novel synchronization mechanisms that enables tags to access synchronized time---this also enables sensor network-like data collection applications where tags collect and timestamp sensor readings that can be queried by the readers later.
 However, maintaining an explicit clock and ensuring synchronization is not trivial due to the marginal energy budgets of the tags and frequent power failures.
-These are the main challenges of our work and its novelty over existing studies.
-}
+These are the main challenges of our work and our novelty over the existing studies.}
 
-\referee{I’m not sure synchronizing tags through the command initiated by the 
-excitation generator (e.g., RFID reader) is feasible in real-world applications.
+\referee{I’m not sure synchronizing tags through the command initiated by the excitation generator (e.g., RFID reader) is feasible in real-world applications.
 This is because these energy-harvesting sensors can be charged within a very short range.
@@
 
-\response{
-We thank the reviewer for raising this issue.
+\response{We thank the reviewer for raising this issue.
 We already answered this concern in the previous responses---please also see \hyperref[R1:3]{[R1:3]}.
@@
 Therefore, we can also employ these strategies in order to create a common time notion among the readers.
-In order to provide an explicit synchronized clock value to the WISP tags, an explicit reader-tag synchronization algorithm is required---which is the focus of our work.
-}
+In order to provide an explicit synchronized clock value to the WISP tags, an explicit reader-tag synchronization algorithm is required---which is the focus of our work.}
 

### 2018-05-10 response/response_letter.tex
Reviewer 2 revised; response letter almost ready

@@
 
-\response{
-	Thank you once more for coordinating the review process of our paper.
+\response{Thank you once more for coordinating the review process of our paper.
 We hope that we have revised our article in accordance with the responses of the reviewers.
@@
 We request the Editor to refer to our respective responses below.
-\textcolor{red}{Add direct references where each of these three points has been addressed.}
-}
+\textcolor{red}{Add direct references where each of these three points has been addressed.}}
 
@@
 
-\response{
-We thank the reviewer for pointing this issue that led us to emphasize our motivation in the revised article.
+\response{We thank the reviewer for pointing this issue that led us to emphasize our motivation in the revised article.
 As indicated by the reviewer, framed slotted Aloha protocol indeed performs a form of synchronization: by adjusting offset and introducing guard times, tags are ensured to communicate at the allocated frame boundaries.
@@
 \textcolor{red}{Add clarification on the difference between clock synchronization with Aloha and our study in Section 1.1.
-and 2.1.3}
-}
+and 2.1.3}}
 
@@
 The bit-level synchronization is not well investigated yet and attracts much attention from academic communities, e.g., the following three references.
-The authors should discuss and clarify whether they study the bit-level synchronization mentioned in these references.
-}
+The authors should discuss and clarify whether they study the bit-level synchronization mentioned in these references.}
 
-\response{
-We thank the reviewer for this comments and for pointing us to these relevant papers, which we cite in Section 2.1.3 of our revised work.
+\response{We thank the reviewer for this comments and for pointing us to these relevant papers, which we cite in Section 2.1.3 of our revised work.
 Thanks to the comment of the reviewer we have noticed that in the previous version of our submission we did not highlighted the main difference between our work and the works pertaining to the \emph{bit-level synchronization}.
@@
 The above discussion has been briefly added to Section 2.1.3.
-\textcolor{red}{Add brief extension on clock synchronization to Section 2.1.3; add references they list explicitly}
-}
+\textcolor{red}{Add brief extension on clock synchronization to Section 2.1.3; add references they list explicitly}}
 
@@
 
-\response{
-As we indicated in the introduction of the article (Section 1) there are already existing synchronization techniques that provide synchronization among RFID readers.
+\response{As we indicated in the introduction of the article (Section 1) there are already existing synchronization techniques that provide synchronization among RFID readers.
 Therefore, if there are multiple readers that cover multiple WISP tags, first the readers should be synchronized with each other and then the tags will be synchronized to the RFID readers.
@@
 We mentioned this explicitly in Section 4 of our revised paper.
-\textcolor{red}{Add clarification about single reader synchronization in Section 4}
-}
+\textcolor{red}{Add clarification about single reader synchronization in Section 4}}
 
@@
 Brownian motion rather than uniform distribution, I still this significantly reduces the theoretical depth of this paper.
-The authors should show what NEW challenges beyond previous work are addressed by this paper.
-}
+The authors should show what NEW challenges beyond previous work are addressed by this paper.}
 
-\response{
-As we indicated in our previous responses (see \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}), our article has the following novelties over the previous results:
+\response{As we indicated in our previous responses (see \hyperref[R1:1]{[R1:1]} and \hyperref[R1:2]{[R1:2]}), our article has the following novelties over the previous results:
 \begin{enumerate}
@@
 This modification did not change the proof structure but it significantly changed the individual steps of the proof---that led to \emph{different steady-state error expressions}.
-\end{enumerate}
-}
+\end{enumerate}}
 
@@
 Hence if the sensors are deployed sparsely, multiple excitation generators are needed, and it is thus important to synchronize these excitation generators as well.
-However, the authors did not consider this point.
-}
+However, the authors did not consider this point.}
 
 \response{We thank the reviewer for raising this issue.
-We already answered this concern in the previous responses---please also see \hyperref[R1:3]{[R1:3]}.
-We would like to mention that there are already prior work that provides synchronization among the readers.
+We already answered this concern in the previous responses---please refer to \hyperref[R1:3]{[R1:3]}.
+We would like to mention that there is already prior work that provides synchronization among the readers.
 Therefore, we can also employ these strategies in order to create a common time notion among the readers.
-In order to provide an explicit synchronized clock value to the WISP tags, an explicit reader-tag synchronization algorithm is required---which is the focus of our work.}
+However, in order to provide an explicit synchronized clock value to the CRFID tags, an explicit reader-tag synchronization algorithm is required---which is the focus of our work.}
 
-\referee{Some questions about the technical parts: \\
-1): the authors claim that the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant clock drift in WISP.
+\referee{The authors claim that the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant clock drift in WISP.
 However, as far as I know, WISP uses a regulator to supply a constant 1.8\,V DC to power the WISP.}
 
-\response{
-The voltage regulators exhibit instability and inaccuracy with respect to the input voltage; i.e.
+\response{The voltage regulators exhibit instability and inaccuracy with respect to the input voltage; i.e.
 they are not perfect and they have an input voltage transient response.
-Therefore, they have an output voltage variation with respect to the input voltage---leading to clock frequency instabilities.
-}
+Therefore, they have an output voltage variation with respect to the input voltage---leading to clock frequency instabilities.}
 
-\referee{2): section 5.1, the authors observe a 1.89\,ms transmission delay. What does this delay account for? the in-air time should not be so long.
-}
+\referee{Section 5.1, the authors observe a 1.89\,ms transmission delay. What does this delay account for? the in-air time should not be so long.}
 
-\response{
-In our article, we defined the \emph{transmission delay} as the time that passes between the start of the broadcast command from the reader and the receipt by the receiver node--- which is the major error source and composed of non-deterministic components.
+\response{In our article, we defined the \emph{transmission delay} as the time that passes between the start of the broadcast command from the reader and the receipt by the receiver node---which is the major error source and composed of non-deterministic components.
 Since we used high-level properties of EPC Gen2 protocol, we could not use low-level timestamping mechanisms and our synchronization approaches suffer from big transmission delays.
 The reason is that we do not know when the reader starts transmission after broadcast command is requested from the PC that controls the reader.
-During this process, the reader-side does not implement any timestamping mechanism and the delays are mainly dependent on the reader-side implementation of the EPC protocol.
-}
+During this process, the reader-side does not implement any timestamping mechanism and the delays are mainly dependent on the reader-side implementation of the EPC protocol.}
 
@@
 
-\response{
-	To address this comment of the reviewer, we have removed unnecessary description from Section 2 of our paper.
+\response{To address this comment of the reviewer, we have removed unnecessary description from Section 2 of our paper.
 At the same time, in this section we have provided a better motivation and listed novelties of our contribution in the context of new related work.
-\textcolor{red}{Compress text in Section 2}
-}
+\textcolor{red}{Compress text in Section 2}}
 
@@
 
-\response{
-We thank the reviewer for this comment.
+\response{We thank the reviewer for this comment.
 However, we feel that there has been a misunderstanding regarding the concept of backscatter (CRFID) multi-hop networks.
@@
 To clarify the connection of our work to the multi-hop networks, we have provided an extra clarifying sentence in Section 1.1.
-
- \textcolor{red}{add a sentence on reader-network-tag synchronization in Section 1.1}
-}
+\textcolor{red}{Add a sentence on reader-network-tag synchronization in Section 1.1}}
 
@@
 
-\response{
-	We thank the reviewer for this comment that led us to provide a better motivation for our study.
+\response{We thank the reviewer for this comment that led us to provide a better motivation for our study.
 Already, in reply to comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we commented on cases where our proposed synchronization technique can be used.
@@
 In order to make inferences on the collected temperature readings, timestamps should be \emph{consistent} so that the reader can provide a temporal ordering of the readings.
-Therefore, tags require a common time notion to timestamp their readers---creating an demand for explicit time synchronization, which we address in our work.
-}
+Therefore, tags require a common time notion to timestamp their readers---creating an demand for explicit time synchronization, which we address in our work.}
 

### 2018-05-12 response/response_letter.tex
finished

@@
 We therefore believe we have addressed all three points that the Editor raised as accurately as possible.
-We request the Editor to refer to our respective responses below.
-\textcolor{red}{Add direct references where each of these three points has been addressed.}}
+Briefly, in the revised article:
+\begin{itemize}
+	\item We provided better a motivation in Section 1.1, we mentioned our contributions more clearly by also citing the related work suggested by Reviewer 1;
+	\item We addressed all of the technical details, especially the differences of our work from slot and bit-level synchronization in EPC Gen2 standard;
+	\item We solved the misunderstanding about multi-hop CRFID networks by clearly stating that these networks do not exist yet and the focus of our work is current of-the-shelf CRFID and reader platforms.
+\end{itemize}
+
+We request the Editor to refer to our respective responses below; in particular \hyperref[R1:1]{[R1:1]}, \hyperref[R1:2]{[R1:2]}, \hyperref[R1:3]{[R1:3]}, \hyperref[R2:1]{[R2:1]} and \hyperref[R3:2]{[R3:2]}.
+}
 
@@
 Unfortunately due to frequent power failures and marginal energy budgets of CRFIDs, providing explicit clock synchronization (that represents \emph{synchronized notion of time}) and maintaining this clock value is challenging.
-We have emphasized these points in our revised article in Section 1.1 and Section 2.1.3.
+We have emphasized these points in our revised article in Section 1.1, 1.2 and Section 2.2.
 We hope that with these answers we have motivated the need for our study sufficiently.
-We kindly refer the reviewer also to our responses \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}.
-\textcolor{red}{Add clarification on the difference between clock synchronization with Aloha and our study in Section 1.1.
-and 2.1.3}}
+We kindly refer the reviewer also to our responses \hyperref[R1:2]{[R1:2]} and \hyperref[R1:4]{[R1:4]}.}
 
@@
 
-\response{We thank the reviewer for this comments and for pointing us to these relevant papers, which we cite in Section 2.1.3 of our revised work.
+\response{We thank the reviewer for this comments and for pointing us to these relevant papers, which we cite in Section 2.2.2 of our revised work.
 Thanks to the comment of the reviewer we have noticed that in the previous version of our submission we did not highlighted the main difference between our work and the works pertaining to the \emph{bit-level synchronization}.
@@
 
-We would like to emphasize that none of these issues have been addressed and studied so far within the context of CRFIDs.
-The above discussion has been briefly added to Section 2.1.3.
-\textcolor{red}{Add brief extension on clock synchronization to Section 2.1.3; add references they list explicitly}}
+We would like to emphasize that none of these issues have been addressed and studied so far within the context of CRFIDs. The above discussion has been briefly added to Section 2.2.2.}
 
@@
 This approach will cause any danger to the protocol we designed.
-We mentioned this explicitly in Section 4 of our revised paper.
-\textcolor{red}{Add clarification about single reader synchronization in Section 4}}
+We mentioned this explicitly in Section 1.1 and 7.5 of our revised paper.}
 
@@
 
-\response{To address this comment of the reviewer, we have removed unnecessary description from Section 2 of our paper.
-At the same time, in this section we have provided a better motivation and listed novelties of our contribution in the context of new related work.
-\textcolor{red}{Compress text in Section 2}}
+\response{To address this comment of the reviewer, we simplified the descriptions in Section 2.
+At the same time, we listed novelties of our contribution in the context of new related work in Section 2.2.2  and in Section 1 we have provided a better motivation.}
 
@@
 using the Internet; then individual tags can be synchronized to the individual RFID readers.
-To clarify the connection of our work to the multi-hop networks, we have provided an extra clarifying sentence in Section 1.1.
-\textcolor{red}{Add a sentence on reader-network-tag synchronization in Section 1.1}}
+To clarify the connection of our work to the multi-hop networks, we have provided an extra clarifying sentence in Section 1.1.}
 
@@
 Already, in reply to comments \hyperref[R1:1]{[R1:1]} and \hyperref[R1:4]{[R1:4]}, we commented on cases where our proposed synchronization technique can be used.
-To recap our answer: in the introduction of our article, we depicted the case where several WISPs  are powered by a reader.
+To recap our answer: in Section 1.1 of our article, we depicted a greenhouse monitoring scneario where several WISPs  are powered by two readers.
 In this scenario, the tags collaboratively timestamp their temperature readings.
 When the reader requests their temperature readings, the tags also transmit their timestamps together with the collected temperature data.
-In order to make inferences on the collected temperature readings, timestamps should be \emph{consistent} so that the reader can provide a temporal ordering of the readings.
+In order to make inferences on the collected temperature readings, timestamps should be \emph{consistent} so that any reader can provide a temporal ordering of the readings.
 Therefore, tags require a common time notion to timestamp their readers---creating an demand for explicit time synchronization, which we address in our work.}

### 2018-05-12 TMC-2017-11-0695-main.tex
finished

@@
 As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}.
-However, recent advancements in the wireless communication enable tag-to-tag communication, e.g.~\cite{liu2013ambient}.
-We anticipate that these multi-hop networks will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.
-In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences at the CRFID tag side.
+\add{Therefore, CRFID tags cannot
+operate in a multi-hop manner.} However, recent advancements in the wireless communication enabled tag-to-tag communication, e.g.~\cite{liu2013ambient}.
+We anticipate that \add{this communication capability will lead up multi-hop CRFID networks that} will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.
+
+\add{Since we are not aware of any network infrastructure to demonstrate neither multi-hop communication nor multi-hop synchronization, our focus in this article is the networks where several readers cover a tag population simultaneously.
+A representative greenhouse monitoring application where several tags are powered by two readers is depicted in Figure~\ref{fig:motivation}.
+In this application, the tags sense the environment and timestamp their temperature readings.
+When the readers request their temperature readings, the tags transmit their timestamps together with the collected temperature data.
+In order to make inferences on the collected temperature readings, timestamps should be consistent so that the readers can provide a temporal ordering of the readings.
+Therefore, tags require a common time notion to timestamp their readers---creating an demand for explicit time synchronization.
+In order to achieve synchronization, first readers should acquire the global time, .e.g.
+by accessing time servers using the Internet; then individual tags should be synchronized to the individual readers.}
+
+%We conjecture that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences at the CRFID tag side.
 Therefore, establishing a common time notion among the tags, rather than synchronizing RFID readers only, will be a crucial service.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\columnwidth]{figures/scenario.pdf}
+	\caption{\label{fig:motivation}\color{blue}A representative greenhouse monitoring application using CRFIDs.
+Two readers are deployed to provide power to the individual tags and to collect their periodic sensor readings.}
+\end{figure}
 
@@
 \label{sec:Motivation}
-
-However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service. The reason is mainly twofold:
+\add{The EPC Gen2 uses slotted Aloha protocol that performs a form of synchronization between the reader and the tags: by adjusting clock offset differences and introducing guard times, tags are ensured to communicate at the allocated slot boundaries.
+However, this method does not compensate for clock drifts and output a synchronized clock value that can be accessed by the tags to timestamp (and make inference about) the collected information, e.g.
+from the sensors attached to tags.
+In consequence, applications requiring time notion cannot be enabled, since a clock should be maintained at each tag---making synchronization service mandatory.
+However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service.
+The reason is mainly twofold:}
 
@@
 	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
+	\item \add{Our time synchronization algorithms compensate for the clock drift and provide access to an \emph{explicit clock} for CRFID applications.
+Moreover, our implementations are independent of how EPC Gen2 is implemented ---providing a level of portability across different platforms.}
 	%
@@
 
-\subsubsection{Wireless Communication}
+\subsubsection{Synchronization Error}
 
@@
 However, such techniques require a calibration phase in order to build a temperature--clock frequency table.
-\TODO{Cite **Top-k Queries for Categorized RFID Systems, IEEE/ACM Transactions on Networking (TON), 2017. \\
-	**Multi-category RFID Estimation, IEEE/ACM Transactions on Networking (TON), 2017. \\
-	**Completely Pinpointing the Missing RFID Tags in a Time-efficient Way, IEEE Transactions on Computers (TC), 2015.}
 
@@
 Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}.
-Since WISP tags are unable to communicate with their neighboring nodes directly using backscatter communication as of now, \emph{tag-tag synchronization} is impossible.
+Since WISP tags are unable to communicate with their neighboring nodes directly using backscatter communication as of now, \emph{tag-tag synchronization} is not feasible.
 
@@
 
+\add{\subsubsection{Slot and Bit-level Synchronization}
+EPC Gen2 requires slot-level synchronization to implement slotted Aloha protocol.
+This is achieved by adjusting clock offsets upon reception from the reader and introducing guard times to ensure that tags communicate at the
+allocated frame boundaries. It is also possible to exploit modulation
+properties of signals from the reader to the tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal, i.e.
+bit-level synchronization~\cite{liu_tcomp_2015,liu_ton_2017,liu2_ton_2017}.
+However, neither slot nor bit level synchronization compensate for clock drifts and output a synchronized clock value that can be accessed by the tags.
+}
+
 \subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term.
 On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift.
-Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.
+\add{Even though embedded platforms are equipped with voltage regulators, the output of these regulators also exhibit instability and inaccuracy with respect to the input voltage; i.e.
+they are not perfect and they have an input voltage transient response.
+Therefore, they have an output
+voltage variation with respect to the input voltage---still leading to clock frequency instabilities.} Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.
 
-\subsubsection{Frequent loss of synchronization state} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g.
+\subsubsection{Frequent power failures} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g.
 the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up.
@@
 
-\subsubsection{Computation and memory overhead sensitivity} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p.
+\subsubsection{Energy and resource constraints} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p.
 44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}.
@@
 \subsubsection{The Lack of broadcast primitive} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader.
-Unfortunately, in the current EPC Gen2 and LLRP standard, we were unable to find any mechanism to send a global timestamp that is received by all tags inside the range of the RFID reader to establish synchronization in one step.
+%Unfortunately, in the current EPC Gen2 and LLRP standard, we were unable to find any mechanism to send a global timestamp that is received by all tags inside the range of the RFID reader to establish synchronization in one step.
 
@@
 Therefore, we require additional steps that will allow tags to obtain the actual reference clock value.
-However, to the best of our knowledge, there is no such command in the EPC Gen 2 standard that allows to send the explicit RFID reader time to the tag.
+\add{Moreover, if the tags are covered by different readers, the first step should be first synchronizing the readers to the real-time.
+e.g.
+by allowing them to access time servers using internet.}
+%However, to the best of our knowledge, there is no such command in the EPC Gen 2 standard that allows to send the explicit RFID reader time to the tag.
 Second, WISP-Sync requires knowledge about $\tau$, which can be RFID reader dependent and requires a sniffer setup described in Section~\ref{sec:sniffer} to measure.

### 2018-05-13 TMC-2017-11-0695-main.tex
Section 1.1 of the paper revised

@@
 As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}.
-\add{Therefore, CRFID tags cannot
-operate in a multi-hop manner.} However, recent advancements in the wireless communication enabled tag-to-tag communication, e.g.~\cite{liu2013ambient}.
+\add{Therefore, CRFID tags cannot operate in a multi-hop manner.} However, recent advancements in the wireless communication enabled tag-to-tag communication, e.g.~\cite{liu2013ambient}.
 We anticipate that \add{this communication capability will lead up multi-hop CRFID networks that} will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.
 
-\add{Since we are not aware of any network infrastructure to demonstrate neither multi-hop communication nor multi-hop synchronization, our focus in this article is the networks where several readers cover a tag population simultaneously.
-A representative greenhouse monitoring application where several tags are powered by two readers is depicted in Figure~\ref{fig:motivation}.
+\add{Since we are not aware of any network infrastructure demonstrating neither multi-hop communication nor multi-hop synchronization, our focus in this article are the CRFID networks where several readers cover a tag population simultaneously in a single hop.
+A representative application is the greenhouse monitoring where several tags are powered by multiple readers, see Figure~\ref{fig:motivation}.
 In this application, the tags sense the environment and timestamp their temperature readings.
@@
 In order to make inferences on the collected temperature readings, timestamps should be consistent so that the readers can provide a temporal ordering of the readings.
-Therefore, tags require a common time notion to timestamp their readers---creating an demand for explicit time synchronization.
-In order to achieve synchronization, first readers should acquire the global time, .e.g.
+Therefore, tags require a common time notion to timestamp their readers---creating a demand for explicit time synchronization.
+In order to achieve synchronization, readers should first acquire the global time, e.g.
 by accessing time servers using the Internet; then individual tags should be synchronized to the individual readers.}
@@
 	\centering
-	\includegraphics[width=0.7\columnwidth]{figures/scenario.pdf}
+	\includegraphics[width=0.5\columnwidth]{figures/scenario.pdf}
 	\caption{\label{fig:motivation}\color{blue}A representative greenhouse monitoring application using CRFIDs.

### 2018-05-13 TMC-2017-11-0695-main.tex
Section 1.2 of the paper revised

@@
 \label{sec:Motivation}
-\add{The EPC Gen2 uses slotted Aloha protocol that performs a form of synchronization between the reader and the tags: by adjusting clock offset differences and introducing guard times, tags are ensured to communicate at the allocated slot boundaries.
+
+\add{The EPC Gen2 protocol uses slotted Aloha that performs a form of synchronization between the reader and the tags: by adjusting clock offset differences and introducing guard times, tags are ensured to communicate at the allocated slot boundaries.
 However, this method does not compensate for clock drifts and output a synchronized clock value that can be accessed by the tags to timestamp (and make inference about) the collected information, e.g.

### 2018-05-13 TMC-2017-11-0695-main.tex
Section 2.2 of the paper revised

@@
 	\item \add{Our time synchronization algorithms compensate for the clock drift and provide access to an \emph{explicit clock} for CRFID applications.
-Moreover, our implementations are independent of how EPC Gen2 is implemented ---providing a level of portability across different platforms.}
+Moreover, our implementations are independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.}
 	%
@@
 
-\add{\subsubsection{Slot and Bit-level Synchronization}
-EPC Gen2 requires slot-level synchronization to implement slotted Aloha protocol.
+\add{\subsubsection{Slot and bit-level synchronization} EPC Gen2 requires a slot-level synchronization to implement slotted Aloha protocol.
 This is achieved by adjusting clock offsets upon reception from the reader and introducing guard times to ensure that tags communicate at the
-allocated frame boundaries. It is also possible to exploit modulation
-properties of signals from the reader to the tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal, i.e.
+allocated frame boundaries.
+It is also possible to exploit modulation properties of signals from the reader to the tags (such as bit timing in FM0 modulation) in order to get the timing measurements of the regular patterns in the signal, i.e.
 bit-level synchronization~\cite{liu_tcomp_2015,liu_ton_2017,liu2_ton_2017}.
-However, neither slot nor bit level synchronization compensate for clock drifts and output a synchronized clock value that can be accessed by the tags.
-}
+However, neither slot nor bit level synchronization compensate for clock drifts and output a synchronized clock value that can be accessed by the tags.}
 
@@
 they are not perfect and they have an input voltage transient response.
-Therefore, they have an output
-voltage variation with respect to the input voltage---still leading to clock frequency instabilities.} Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.
+Therefore, they have an output voltage variation with respect to the input voltage---still leading to clock frequency instabilities.} Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.
 

### 2018-05-13 TMC-2017-11-0695-main.tex
Section 2.2 of the paper revised

@@
 	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
-	\item \add{Our time synchronization algorithms compensate for the clock drift and provide access to an \emph{explicit clock} for CRFID applications.
+	\item \add{Our time synchronization algorithms compensate for the clock drift and enable the first step to enable an \emph{explicit clock} for CRFID applications.
 Moreover, our implementations are independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.}
@@
 Therefore, we require additional steps that will allow tags to obtain the actual reference clock value.
-\add{Moreover, if the tags are covered by different readers, the first step should be first synchronizing the readers to the real-time.
+\add{Moreover, if the tags are covered by different readers, the first step should be to first synchronize the readers to the real-time.
 e.g.

### 2018-05-14 response/response_letter.tex
intro to the response letter revised (address changed)

@@
 Delft University of Technology \\ 
-Mekelweg 4, 2628 CD, Delft, The Netherlands \\
+Van Mourik Broekmanweg 6\\
+2628\,XE Delft, The Netherlands \\
 E-mail: p.pawelczak@tudelft.nl\\
@@
 
-We would like to thank for handling our paper and providing a rapid review results. We are also grateful to the Reviewers for their substantial effort spent for evaluating our work. \\
+We would like to thank for handling our submission and providing a rapid review results. We are also grateful to the reviewers for their substantial effort spent for evaluating our work. \\
 
-With this letter we provide our response to all comments of the Reviewers and the revised version of our paper.
+With this letter we provide our response to all comments of the reviewers and the revised version of our paper.
 We look forward to hearing from you at the earliest and we would be happy to provide further clarifications and revisions if necessary.

### 2018-05-14 response/response_letter.tex
Revision of response to the editor

@@
 We hope that we have revised our article in accordance with the responses of the reviewers.
-We therefore believe we have addressed all three points that the Editor raised as accurately as possible.
+We believe we have addressed all three points that the editor raised as accurately as possible.
 Briefly, in the revised article:
 \begin{itemize}
-	\item We provided better a motivation in Section 1.1, we mentioned our contributions more clearly by also citing the related work suggested by Reviewer 1;
-	\item We addressed all of the technical details, especially the differences of our work from slot and bit-level synchronization in EPC Gen2 standard;
-	\item We solved the misunderstanding about multi-hop CRFID networks by clearly stating that these networks do not exist yet and the focus of our work is current of-the-shelf CRFID and reader platforms.
+	\item We provided better a motivation in Section 1.1, we mentioned our contributions more clearly by also citing the related work suggested by Reviewer 1 in Section 2.2.2;
+	\item We addressed all of the technical details, especially the differences of our work from slot and bit-level synchronization in EPC Gen2 standard in Section 2.2.2;
+	\item We clarified the misunderstanding about multi-hop CRFID networks by clearly stating in Section 1.1.
+that these networks do not exist yet and the focus of our work is current of-the-shelf CRFID and reader platforms.
 \end{itemize}
 
-We request the Editor to refer to our respective responses below; in particular \hyperref[R1:1]{[R1:1]}, \hyperref[R1:2]{[R1:2]}, \hyperref[R1:3]{[R1:3]}, \hyperref[R2:1]{[R2:1]} and \hyperref[R3:2]{[R3:2]}.
-}
+We request the editor to refer to our respective responses below; in particular \hyperref[R1:1]{[R1:1]}, \hyperref[R1:2]{[R1:2]}, \hyperref[R1:3]{[R1:3]}, \hyperref[R2:1]{[R2:1]} and \hyperref[R3:2]{[R3:2]}.
+Additions with respect to the previous version of our submission are marked in blue text in our latest paper version.}
 

### 2018-05-14 response/response_letter.tex
Revision of response to reviewe 1 (minor edits)

@@
 	\item \textbf{Portability:} Bit-level synchronization relies on high-precision clock to obtain microsecond-precision timing measurements.
-Despite, in our work, we used the external 32\,kHz clock that can run in low-power standby mode and the \emph{higher-level} commands of the EPC Gen 2 standard, rather than exploiting its low-level signal properties.
+Despite, in our work, we used the external 32\,kHz clock that can run in low-power standby mode and the \emph{higher-level} commands of the EPC Gen2 standard, rather than exploiting its low-level signal properties.
 Therefore, our approach is independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.
@@
 
-We would like to emphasize that none of these issues have been addressed and studied so far within the context of CRFIDs. The above discussion has been briefly added to Section 2.2.2.}
+We would like to emphasize that none of these issues have been addressed and studied so far within the context of CRFIDs. Again, the above discussion has been briefly added to Section 2.2.2.}
 
@@
 This approach will cause any danger to the protocol we designed.
-We mentioned this explicitly in Section 1.1 and 7.5 of our revised paper.}
+We mentioned this explicitly in Section 1.1 and Section 7.5 of our revised paper.}
 
@@
 	\item \textbf{Novel theoretical consideration:} We also provided an in-depth theoretical analysis of the proposed synchronization techniques.
-As indicate by the reviewer, the main part of the analysis is indeed from previous work.
+As indicate by the reviewer, the main part of the analysis indeed originates from the previous work.
 However, the previous analysis that assumes uniform distribution for the clock drifts was not realistic---the reason we changed the analysis to use the Brownian motion.
-This modification did not change the proof structure but it significantly changed the individual steps of the proof---that led to \emph{different steady-state error expressions}.
+This modification did not change the proof structure but it changed the individual steps of the proof---that led to \emph{different steady-state error expressions}.
 \end{enumerate}}

### 2018-05-14 response/response_letter.tex
Revision of response to reviewer 3 (minor edits)
Revision of response R3.1

@@
 
-\response{To address this comment of the reviewer, we simplified the descriptions in Section 2.
-At the same time, we listed novelties of our contribution in the context of new related work in Section 2.2.2  and in Section 1 we have provided a better motivation.}
+\response{To address this comment of the reviewer, we modified the description in Section 2.
+That is, we have listed novelties of our contribution in the context of new related work in Section 2.2.2.
+Accompanying this revision, in Section 1 we have provided a better motivation.}
 
@@
 using the Internet; then individual tags can be synchronized to the individual RFID readers.
-To clarify the connection of our work to the multi-hop networks, we have provided an extra clarifying sentence in Section 1.1.}
+To clarify the connection of our work to the multi-hop networks, we have provided an extra clarifying sentences in Section 1.1.}
 

### 2018-05-15 response/response_letter.tex
fix typos

@@
 \begin{itemize}
-	\item We provided better a motivation in Section 1.1, we mentioned our contributions more clearly by also citing the related work suggested by Reviewer 1 in Section 2.2.2;
+	\item We provided a better motivation in Section 1.1, we mentioned our contributions more clearly by also citing the related work suggested by Reviewer 1 in Section 2.2.2;
 	\item We addressed all of the technical details, especially the differences of our work from slot and bit-level synchronization in EPC Gen2 standard in Section 2.2.2;
-	\item We clarified the misunderstanding about multi-hop CRFID networks by clearly stating in Section 1.1.
-that these networks do not exist yet and the focus of our work is current of-the-shelf CRFID and reader platforms.
+	\item We clarified the misunderstanding about multi-hop CRFID networks by clearly stating in Section 1.1 that these networks do not exist yet and the focus of our work is on the current of-the-shelf CRFID and reader platforms.
 \end{itemize}

### 2018-05-15 response/response_letter.tex
fix typos

@@
 
-	\item \textbf{Power Failures:} Providing synchronization service to the application layer requires to cope with power failures, which has not been investigated so far.
+	\item \textbf{Power Failures:} Providing synchronization service to the application layer requires coping with power failures, which has not been investigated so far.
 To address this issue, we provided synchronization algorithms that can recover easily after power failures.
@@
 Therefore, if there are multiple readers that cover multiple WISP tags, first the readers should be synchronized with each other and then the tags will be synchronized to the RFID readers.
-This approach will cause any danger to the protocol we designed.
+This approach will not cause any danger to the protocol we designed.
 We mentioned this explicitly in Section 1.1 and Section 7.5 of our revised paper.}

### 2018-05-15 response/response_letter.tex
Revision of response R1.2 and typo in the intro

@@
 
-in the name of all Authors---Dr. Przemys{\l}aw Pawe{\l}czak
+in the name of all authors---Dr. Przemys{\l}aw Pawe{\l}czak
 
@@
 \begin{itemize}
-	\item \textbf{Portability:} Bit-level synchronization relies on high-precision clock to obtain microsecond-precision timing measurements.
-Despite, in our work, we used the external 32\,kHz clock that can run in low-power standby mode and the \emph{higher-level} commands of the EPC Gen2 standard, rather than exploiting its low-level signal properties.
+	\item \textbf{Portability:} Bit-level synchronization relies on a high-precision clock to obtain microsecond-precision timing measurements.
+However, rather than exploiting low-level signal properties, in our work we rely on (i) external 32\,kHz clock that can run in low-power standby mode and (ii) the \emph{higher-level} commands of the EPC Gen2 standard.
 Therefore, our approach is independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.
-However, our synchronization algorithms can also be integrated into the bit-level synchronization;
+We emphasize that our synchronization algorithms can also be integrated into the bit-level synchronization;
 

### 2018-05-17 TMC-2017-11-0695-main.tex
figure
Figure 1 size reduced
typo
latest changes

@@
 \usepackage{url}
-\usepackage{hyperref}
+%\usepackage{hyperref}
 \usepackage{stmaryrd}
@@
 	\centering
-	\includegraphics[width=0.5\columnwidth]{figures/scenario.pdf}
+	\includegraphics[width=0.75\columnwidth]{figures/scenario.pdf}
 	\caption{\label{fig:motivation}\color{blue}A representative greenhouse monitoring application using CRFIDs.
@@
 We note that WISP-side implementations throughout the article are done using C and assembly language.
-All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\href{https://goo.gl/RuWIZ2}{https://goo.gl/RuWIZ2}.
+All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\url{https://goo.gl/RuWIZ2}.
 
@@
 \end{eqnarray}
-which is derived in Section~\ref{sec:app_b} of the Appendix.
+which is derived in Section A.2 of the Appendix.
 
@@
 	\includegraphics[width=\columnwidth]{figures/sim-theory.eps}
-	\caption{\label{fig:theory-vs-simulations} The plot of the cumulative distribution function (CDF) of $\gamma$ in \eqref{eq:sync_performance} and the CDF of the steady-state $\gamma$ obtained during numerical simulations of \eqref{eq:update_rule} (by implementing \eqref{eq:theta} and \eqref{eq:delta_recursive} in Appendix).
+	\caption{\label{fig:theory-vs-simulations} The plot of the cumulative distribution function (CDF) of $\gamma$ in \eqref{eq:sync_performance} and the CDF of the steady-state $\gamma$ obtained during numerical simulations of \eqref{eq:update_rule} (by implementing (15) and (16) in Appendix).
 We used $\tau=1$, $\beta= 0.0001$, $\sigma_{\tau}=0.00005$ and our plots depict the theoretical and simulation results with respect to $\sigma_\eta = 0.0002$ and $\sigma_\eta = 0.0008.$ It can be observed that \eqref{eq:sync_performance} is consistent with the simulations and bigger uncertainty leads to bigger synchronization error.}
@@
 Observe that increasing $\tau$ increases $\sigma^2_{\gamma}$.
-Then, we set $\sigma_{\tau}=0.4097$,	$\tau=1$ and $\beta \in (0,2)$ due to \eqref{eq:step_size_bounds-pi}.
+Then, we set $\sigma_{\tau}=0.4097$,	$\tau=1$ and $\beta \in (0,2)$ due to (19).
 Using different values for $\sigma_\eta$, we plotted how $\sigma^2_{\gamma}$ changes with respect to the $\beta$ (right hand side).
@@
 First, we explored how $\beta$ affects the performance of the algorithm.
-By substituting $\tau$ into \eqref{eq:step_size_bounds-pi}, the convergence condition becomes $0<\beta<0.000284$ in our case.
+By substituting $\tau$ into (19), the convergence condition becomes $0<\beta<0.000284$ in our case.
 We present the synchronization error with different $\beta$ values in Fig.~\ref{fig:beta}.
@@
 Therefore, we require additional steps that will allow tags to obtain the actual reference clock value.
-\add{Moreover, if the tags are covered by different readers, the first step should be to first synchronize the readers to the real-time.
-e.g.
+\add{Moreover, if the tags are covered by different readers, the first step should be synchronizing the readers to the real-time, e.g.
 by allowing them to access time servers using internet.}
@@
 
-\input{appendix_tmc}
+%\input{appendix_tmc}
 

### 2018-05-17 TMC-appendix.tex
latest changes

--- 2017-11-01 appendix_tmc.tex
@@
-\newpage\null\thispagestyle{empty}\newpage
+\documentclass[10pt,journal,compsoc]{IEEEtran}
+
+\ifCLASSOPTIONcompsoc
+  % IEEE Computer Society needs nocompress option
+  % requires cite.sty v4.0 or later (November 2003)
+  \usepackage[nocompress]{cite}
+\else
+  % normal IEEE
+  \usepackage{cite}
+\fi
+
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+
+\newcommand{\expected}[1]{E\!\left\llbracket #1 \right\rrbracket}
+
+\addtocounter{equation}{7}
+
+\begin{document}
+
+\title{On the Synchronization of Computational RFIDs}
+
+\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, 
+	Henko~Aantjes, 
+	Przemys{\l}a~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
+	and~Amjad~Yousef~Majid
+	%
+	\IEEEcompsocitemizethanks{
+		%\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}. \protect\\
+		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).
+	}
+%	\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}
+}
+
+\maketitle
+
+\IEEEdisplaynontitleabstractindextext
+
+\IEEEpeerreviewmaketitle
+
 \appendices
-\section{Analysis of Algorithm~\ref{alg:sync}}
+\section{Analysis of Algorithm 1}
 
-We provide the synchronization conditions of Algorithm~\ref{alg:sync} and show that the synchronization will be established eventually.
+We provide the synchronization conditions of Algorithm 1 and show that the synchronization will be established eventually.
 We need to mention that we follow similar calculation steps to the ones presented in the studies~\cite{pi2015,Yildirim:Gradient:2016}.
@@
 \subsection{Proof of Convergence}
-Consider \eqref{eq:gamma_t_1} and without loss of 
+Consider (5) and without loss of 
 generality we assume that 
@@
 uncertainty of the event period (as 
-indicated in Section \ref{subsec:event_period_uncertainty}). For ease of 
+indicated in Section 6.1). For ease of 
 calculations we assume that $\varepsilon\sim\mathcal{N}(0,\sigma^2_{\tau})$. 
@@
 \end{eqnarray}
-Using \eqref{eq:gamma_t_1}, \eqref{eq:drift_relation} and 
+Using (5), \eqref{eq:drift_relation} and 
 \eqref{eq:relative_drift_estimation_error} we get:
@@
 \nonumber\\
-&\overset{\eqref{eq:update_rule},\eqref{eq:drift}}{=}& 
+&\overset{(6),\eqref{eq:drift}}{=}& 
 \Delta(h)+\int_{t_h}^{t_{h+1}}\eta(t)dt-\beta\Theta(h+1) \nonumber \\
@@
 Without loss of generality, assume that $t_{h+1}-t_h\approxeq\tau$. Therefore, 
-the asymptotic estimation error of Algorithm \ref{alg:sync} is given by
+the asymptotic estimation error of Algorithm 1 is given by
 \begin{eqnarray}
@@
 \end{eqnarray}
-which results in \eqref{eq:sync_performance}.
+which results in (7).
+
+\bibliographystyle{IEEEtran}
+\bibliography{IEEEabrv,references}
+
+\end{document}

### 2018-06-29 TMC-2017-11-0695.R1-main.tex
minor revision

--- 2018-05-17 TMC-2017-11-0695-main.tex
@@
 	\IEEEcompsocitemizethanks{
-		\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}. \protect\\
+		\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}.
 		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).
+		\IEEEcompsocthanksitem Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m is also with the Computer Engineering Department of Ege University, Izmir, Turkey.
 	}
@@
 As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}.
-\add{Therefore, CRFID tags cannot operate in a multi-hop manner.} However, recent advancements in the wireless communication enabled tag-to-tag communication, e.g.~\cite{liu2013ambient}.
-We anticipate that \add{this communication capability will lead up multi-hop CRFID networks that} will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.
+Therefore, CRFID tags cannot operate in a multi-hop manner.
+However, recent advancements in the wireless communication enabled tag-to-tag communication, e.g.~\cite{liu2013ambient}.
+We anticipate that this communication capability will lead up multi-hop CRFID networks that will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts.
 
-\add{Since we are not aware of any network infrastructure demonstrating neither multi-hop communication nor multi-hop synchronization, our focus in this article are the CRFID networks where several readers cover a tag population simultaneously in a single hop.
+Since we are not aware of any network infrastructure demonstrating neither multi-hop communication nor multi-hop synchronization, our focus in this article are the CRFID networks where several readers cover a tag population simultaneously in a single hop.
 A representative application is the greenhouse monitoring where several tags are powered by multiple readers, see Figure~\ref{fig:motivation}.
@@
 In order to achieve synchronization, readers should first acquire the global time, e.g.
-by accessing time servers using the Internet; then individual tags should be synchronized to the individual readers.}
+by accessing time servers using the Internet; then individual tags should be synchronized to the individual readers.
 
@@
 	\includegraphics[width=0.75\columnwidth]{figures/scenario.pdf}
-	\caption{\label{fig:motivation}\color{blue}A representative greenhouse monitoring application using CRFIDs.
+	\caption{\label{fig:motivation}A representative greenhouse monitoring application using CRFIDs.
 Two readers are deployed to provide power to the individual tags and to collect their periodic sensor readings.}
@@
 
-\add{The EPC Gen2 protocol uses slotted Aloha that performs a form of synchronization between the reader and the tags: by adjusting clock offset differences and introducing guard times, tags are ensured to communicate at the allocated slot boundaries.
+The EPC Gen2 protocol uses slotted Aloha that performs a form of synchronization between the reader and the tags: by adjusting clock offset differences and introducing guard times, tags are ensured to communicate at the allocated slot boundaries.
 However, this method does not compensate for clock drifts and output a synchronized clock value that can be accessed by the tags to timestamp (and make inference about) the collected information, e.g.
@@
 However, the characteristics of CRFID systems expose fundamentally different challenges than battery-powered sensor networks to implement the synchronization service.
-The reason is mainly twofold:}
+The reason is mainly twofold:
 
@@
 	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
-	\item \add{Our time synchronization algorithms compensate for the clock drift and enable the first step to enable an \emph{explicit clock} for CRFID applications.
-Moreover, our implementations are independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.}
+	\item Our time synchronization algorithms compensate for the clock drift and enable the first step to enable an \emph{explicit clock} for CRFID applications.
+Moreover, our implementations are independent of how EPC Gen2 is implemented---providing a level of portability across different platforms.
 	%
@@
 
-\add{\subsubsection{Slot and bit-level synchronization} EPC Gen2 requires a slot-level synchronization to implement slotted Aloha protocol.
+\subsubsection{Slot and bit-level synchronization} EPC Gen2 requires a slot-level synchronization to implement slotted Aloha protocol.
 This is achieved by adjusting clock offsets upon reception from the reader and introducing guard times to ensure that tags communicate at the
@@
 bit-level synchronization~\cite{liu_tcomp_2015,liu_ton_2017,liu2_ton_2017}.
-However, neither slot nor bit level synchronization compensate for clock drifts and output a synchronized clock value that can be accessed by the tags.}
+However, neither slot nor bit level synchronization compensate for clock drifts and output a synchronized clock value that can be accessed by the tags.
 
@@
 On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift.
-\add{Even though embedded platforms are equipped with voltage regulators, the output of these regulators also exhibit instability and inaccuracy with respect to the input voltage; i.e.
+Even though embedded platforms are equipped with voltage regulators, the output of these regulators also exhibit instability and inaccuracy with respect to the input voltage; i.e.
 they are not perfect and they have an input voltage transient response.
-Therefore, they have an output voltage variation with respect to the input voltage---still leading to clock frequency instabilities.} Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.
+Therefore, they have an output voltage variation with respect to the input voltage---still leading to clock frequency instabilities.
+Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level, see Fig.~\ref{fig:frequency_dc_rf}.
 
@@
 Therefore, we require additional steps that will allow tags to obtain the actual reference clock value.
-\add{Moreover, if the tags are covered by different readers, the first step should be synchronizing the readers to the real-time, e.g.
-by allowing them to access time servers using internet.}
+Moreover, if the tags are covered by different readers, the first step should be synchronizing the readers to the real-time, e.g.
+by allowing them to access time servers using internet.
 %However, to the best of our knowledge, there is no such command in the EPC Gen 2 standard that allows to send the explicit RFID reader time to the tag.
@@
 
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/ksy.jpeg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
-%received BSc(Eng), M.Sc. and Ph.D. degrees from Ege University, İzmir, 
-%Turkey in 2003, 2006 and 2012 respectively.
-He is  currently with Embedded Software group of TU Delft.
-His research interests are embedded systems, distributed algorithms,  wireless sensor networks, energy harvesting systems and computational RFIDs.
-%\end{IEEEbiography}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/ksy.jpeg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
+received his Ph.D. degree in Computer Engineering from Ege University 
+in 2012.
+He was a visiting scholar at the Department of Information Engineering, University of Padova in 2013.
+Between 2015-2017, he was a post-doctoral researcher at Embedded Software Group, Delft University of Technology.
+Currently, he is an assistant professor at the Department of Computer Engineering, Ege University and also a research affiliate with the Embedded Software Group at Delft University of Technology.
+His research interests are embedded systems and software, distributed algorithms,  wireless sensor networks, energy harvesting systems and computational RFIDs.
+\end{IEEEbiography}
+
+% if you will not have a photo at all:
+\begin{IEEEbiographynophoto}{Henko Aantjes}
+Biography text here.
+\end{IEEEbiographynophoto}
 %
-%% if you will not have a photo at all:
-%\begin{IEEEbiographynophoto}{Henko Aantjes}
-%Biography text here.
-%\end{IEEEbiographynophoto}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.jpg}}]{}
+Biography text here.	
+\end{IEEEbiography}
 %
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.jpg}}]{}
-%Biography text here.	
-%\end{IEEEbiography}
-%
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
-%Biography text here.
-%\end{IEEEbiography}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
+Biography text here.
+\end{IEEEbiography}
 %

### 2018-06-29 response/TMC-2017-11-0695.R1-response.tex
minor revision

\documentclass[10pt]{article}
\usepackage{amsmath,amsfonts,amssymb,graphicx,color,psfrag}
\usepackage[utf8]{inputenc}

\usepackage{vmargin}
\setmarginsrb{2.25cm}{2.25cm}{2.25cm}{2.25cm}
         {12pt}{20pt}{12pt}{36pt}

\usepackage{soul}
\usepackage{hyperref}

\newcommand{\referee}[1]{
	{\item \color{OliveGreen} \emph{{#1}}}
	\label{R\therefereeCounter:\arabic{enumi}}
}

\definecolor{OliveGreen}{rgb}{0.1,0.2,0.5}

\newcommand{\response}[1]{{\color{blue} #1}}

\newenvironment{editor}{%
\textbf{\large Comments by the Editor}
\begin{enumerate}%
\renewcommand{\labelenumi}{\textbf{[E:\,\arabic{enumi}]}} %
}{ \end{enumerate}}

\newcounter{refereeCounter}

\newenvironment{responses}{%
\refstepcounter{refereeCounter}%
\textbf{\large Comments by Referee \therefereeCounter}
\begin{enumerate}%
\renewcommand{\labelenumi}{\textbf{[R\therefereeCounter :\,\arabic{enumi}]}} %
}{\end{enumerate}}

\setlength{\parindent}{0pt}

\begin{document}

\pagestyle{myheadings}
\thispagestyle{empty}

\markright{\footnotesize TMC-2017-11-0695.R1: {\sl On the Synchronization of Computational RFIDs}}

\headsep 0.5cm

\bigskip\bigskip

%\noindent{Marwan Krunz, EIC, and Kevin Almeroth, Associate EIC \\
%	IEEE Transactions on Mobile Computing
%}

\bigskip\bigskip

\begin{flushright}
Dr. Przemys{\l}aw Pawe{\l}czak \\
Embedded Software Group \\ 
Delft University of Technology \\ 
Van Mourik Broekmanweg 6\\
2628\,XE Delft, The Netherlands \\
E-mail: p.pawelczak@tudelft.nl\\
Phone: +31 15 278 7491\\
\end{flushright}

\vspace*{2cm}

\today
\medskip

\textbf{TMC-2017-11-0695.R1:} {\sl On the Synchronization of Computational RFIDs}

\bigskip

Dear Editor,\\

We would like to thank for handling our submission and providing a rapid review results. We are also grateful to the reviewers for their substantial effort spent for evaluating our work. \\

With this letter we provide our response to all comments of the reviewers and the revised version of our paper. We look forward to hearing from you at the earliest and we would be happy to provide further clarifications if necessary.

With best regards, \\

in the name of all authors---Dr. Przemys{\l}aw Pawe{\l}czak

\pagebreak

\begin{editor}
\referee{The authors need to go through a minor revision to address the concern raised by reviewer 3 about the motivation.
}

\response{Thank you once more for coordinating the review process of our paper. We believe that there is a misunderstanding that led to a concern raised by the reviewer 3. We addressed this concern in our responce \hyperref[R3:1]{[R3:1]}.
}

\end{editor}

\begin{responses}

\referee{Accept With No Changes.}

\response{We are pleased to receive this decision and thank the reviewer for his/her time spent during the evaluation.}

\end{responses}

\begin{responses}

\referee{Accept With No Changes.}

\response{We are pleased to receive this decision and thank the reviewer for his/her time spent during the evaluation.}

\end{responses}

\begin{responses}

\referee{It's good to see the authors addressed my concerns. While I'm satisfied with most of the responses, the motivation is still not clear to me. The authors claim that in greenhouse monitoring scenario, the nodes need to be synchronized for data reporting. Why does ALOHA not work here? I know the node did not provide timestamp in ALOHA protocol. However, the reader could rely on its own clock to determine the timestamp of each node access. You should pay more attention to this.}

\response{As we indicated in Section 2.2.2 of our article, slotted ALOHA protocol in EPC Gen2 does not compensate for the clock drifts and output an explicit clock value that can be accessed by the tags. Therefore, while the nodes are collecting data and storing the data in their local buffers, they cannot timestamp the collected data using synchronized notion of time---they cannot obtain actual (value,timestamp) pairs. When the data is queried by the RFID reader, the local clock of the reader does not represent the actual data collection time. Therefore, the reader could not rely on its own clock to determine the timestamp values. Time synchronization is required for obtaining the data collection timestamps precisely---the tags access the synchronized time notion, obtain (data, timestamp) pairs and store these pairs in their local buffers; thereafter these values can be queried by the RFID reader at any time so that the temporal ordering if the received pairs can be performed. }

\end{responses}

\end{document}

### 2018-07-01 TMC-2017-11-0695.R1-main.tex
add Amjad biography

@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
-Biography text here.
+obtained his MSc.
+in Computer science from Delft University of Technology in 2015.
+Currently, he is a PhD.
+candidate at Embedded Software Group at Delft University.
+His research focuses on intermittent software architecture, (computational) RFID systems, backscatter technology, and wireless sensor networks.
 \end{IEEEbiography}

### 2018-07-10 TMC-2017-11-0695.R1-main.tex
Henko Aantjes bio added (first version), front footer with addressess changed, revisions of bios

--- 2018-06-29 TMC-2017-11-0695.R1-main.tex
@@
 	Henko~Aantjes, 
-	Przemys{\l}a~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
+	Przemys{\l}aw~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
 	and~Amjad~Yousef~Majid
@@
 		\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}.
-		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).
-		\IEEEcompsocthanksitem Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m is also with the Computer Engineering Department of Ege University, Izmir, Turkey.
+		\IEEEcompsocthanksitem Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m, Przemys{\l}aw Pawe{\l}czak and Amjad Yousef Majid are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl).
+		\IEEEcompsocthanksitem Henko Aantjes is with the Brighsight BV, Brassersplein 2, 2612\,CT, Delft, The Netherlands (e-mail: aantjes@brightsight.com)
+		\IEEEcompsocthanksitem Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m is also affilated with the Computer Engineering Department of Ege University, Izmir, Turkey.
 	}
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/ksy.jpeg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
-received his Ph.D. degree in Computer Engineering from Ege University 
-in 2012.
+received his Ph.D.
+degree in Computer Engineering from Ege University in 2012.
 He was a visiting scholar at the Department of Information Engineering, University of Padova in 2013.
-Between 2015-2017, he was a post-doctoral researcher at Embedded Software Group, Delft University of Technology.
+Between 2015 and 2017, he was a post-doctoral researcher at Embedded Software Group, Delft University of Technology.
 Currently, he is an assistant professor at the Department of Computer Engineering, Ege University and also a research affiliate with the Embedded Software Group at Delft University of Technology.
@@
 % if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{Henko Aantjes}
-Biography text here.
-\end{IEEEbiographynophoto}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/henko.jpg}}]{Henko Aantjes}
+Henko Aantjes received his MSc degree in 201X.
+Currently he is a security evaluator at Brightsight BV, Delft, The Netherlands.
+His research interests include computational RFIDs, secure computation, and contactless payments.
+\end{IEEEbiography}
 %
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
-Biography text here.
+obtained his MSc degree in Computer Science from Delft University of Technology in 2015.
+Currently, he is a PhD candidate at the Embedded Software Group at Delft University of Technology.
+His research focuses on intermittent software architecture, (computational) RFIDs, backscatter technology, and wireless sensor networks.
 \end{IEEEbiography}

### 2018-07-10 TMC-2017-11-0695.R1-main.tex
bio of all authors revised

@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/ksy.jpeg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/sinan.jpg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
 received his Ph.D.
-degree in Computer Engineering from Ege University in 2012.
+degree in Computer Engineering from Ege University, Izmir, Turkey, in 2012.
 He was a visiting scholar at the Department of Information Engineering, University of Padova in 2013.
-Between 2015 and 2017, he was a post-doctoral researcher at Embedded Software Group, Delft University of Technology.
+Between 2015 and 2017, he was a postdoctoral researcher at Embedded Software Group, Delft University of Technology, Delft, The Netherlands.
 Currently, he is an assistant professor at the Department of Computer Engineering, Ege University and also a research affiliate with the Embedded Software Group at Delft University of Technology.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/henko.jpg}}]{Henko Aantjes}
-Henko Aantjes received his MSc degree in 201X.
+Henko Aantjes received his MSc degree in 2016 from Delft University of Technology, Delft, The Netherlands.
 Currently he is a security evaluator at Brightsight BV, Delft, The Netherlands.
@@
 %
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.jpg}}]{}
-Biography text here.	
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.jpg}}]{Przemys{\l}aw~Pawe{\l}czak} (S'03-M'10) received the MSc degree from Wroc{\l}aw University of Technology, Poland, in 2004 and the PhD degree from Delft University of Technology, The Netherlands in 2009.
+ Between 2009 and 2011 he was a postdoctoral researcher at the Cognitive Reconfigurable Embedded Systems Lab, University of California, Los Angeles.
+Between 2012 and 2013 he was a research staff member at Fraunhofer Institute for Telecommunications, Heinrich Hertz Institute, Berlin, Germany.
+Since 2013 he is a (tenured) assistant professor at the Embedded Software Group of Delft University of Technology.
+His research interests include wireless networking, RFID systems and intermittently-powered embedded devices.
 \end{IEEEbiography}
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
-obtained his MSc degree in Computer Science from Delft University of Technology in 2015.
-Currently, he is a PhD candidate at the Embedded Software Group at Delft University of Technology.
-His research focuses on intermittent software architecture, (computational) RFIDs, backscatter technology, and wireless sensor networks.
+obtained his MSc degree in 2015 from Delft University of Technology, Delft, The Netherlands.
+Currently, he is a PhD candidate at the Embedded Software Group at the same university.
+His research focuses on intermittent software architecture, (computational) RFIDs, and backscatter communication.
 \end{IEEEbiography}

### 2018-07-10 response/TMC-2017-11-0695.R1-response.tex
Response letter of 0695.R1 revised

@@
 \setlength{\parindent}{0pt}
+
+%Count from three - ignore two other reviewers 
+\stepcounter{refereeCounter}
+\stepcounter{refereeCounter}
 
@@
 
-We would like to thank for handling our submission and providing a rapid review results. We are also grateful to the reviewers for their substantial effort spent for evaluating our work. \\
+We would like to thank for handling our revised submission and providing a rapid review results. We are also once more grateful to the reviewers for their effort spent for evaluating our work. \\
 
-With this letter we provide our response to all comments of the reviewers and the revised version of our paper.
+With this letter we provide our response to the final comment, together with the revised version of our paper.
 We look forward to hearing from you at the earliest and we would be happy to provide further clarifications if necessary.
@@
 \begin{editor}
-\referee{The authors need to go through a minor revision to address the concern raised by reviewer 3 about the motivation.
-}
+\referee{The authors need to go through a minor revision to address the concern raised by reviewer 3 about the motivation.}
 
-\response{Thank you once more for coordinating the review process of our paper.
-We believe that there is a misunderstanding that led to a concern raised by the reviewer 3.
-We addressed this concern in our responce \hyperref[R3:1]{[R3:1]}.
-}
+\response{Thank you once more for coordinating the review process of our paper. We have addressed the final concern of Reviewer 3 in our response given below.}
 
@@
 
-\referee{Accept With No Changes.}
-
-\response{We are pleased to receive this decision and thank the reviewer for his/her time spent during the evaluation.}
-
-\end{responses}
-
-\begin{responses}
-
-\referee{Accept With No Changes.}
-
-\response{We are pleased to receive this decision and thank the reviewer for his/her time spent during the evaluation.}
-
-\end{responses}
-
-\begin{responses}
-
-\referee{It's good to see the authors addressed my concerns.
-While I'm satisfied with most of the responses, the motivation is still not clear to me.
+\referee{While I'm satisfied with most of the responses, the motivation is still not clear to me.
 The authors claim that in greenhouse monitoring scenario, the nodes need to be synchronized for data reporting.
@@
 
-\response{As we indicated in Section 2.2.2 of our article, slotted ALOHA protocol in EPC Gen2 does not compensate for the clock drifts and output an explicit clock value that can be accessed by the tags.
-Therefore, while the nodes are collecting data and storing the data in their local buffers, they cannot timestamp the collected data using synchronized notion of time---they cannot obtain actual (value,timestamp) pairs.
+\response{Thank you once more for giving us the opportunity to address this comment.
+As indicated in Section 2.2.2 of our submission, slotted ALOHA protocol in the EPC Gen2 does not compensate for the clock drifts and output an explicit clock value that can be accessed by the RFID tags.
+Therefore, while the nodes are collecting data and storing the data in their local buffers, they \emph{cannot timestamp} the collected data using synchronized notion of time---in other words, they cannot obtain actual \texttt{(value, timestamp)} pairs.
 When the data is queried by the RFID reader, the local clock of the reader does not represent the actual data collection time.
 Therefore, the reader could not rely on its own clock to determine the timestamp values.
-Time synchronization is required for obtaining the data collection timestamps precisely---the tags access the synchronized time notion, obtain (data, timestamp) pairs and store these pairs in their local buffers; thereafter these values can be queried by the RFID reader at any time so that the temporal ordering if the received pairs can be performed.
+Time synchronization is required for obtaining the data collection timestamps precisely---the tags access the synchronized time notion, obtain \texttt{(value, timestamp)} pairs and store these pairs in their local buffers; thereafter these values can be queried by the RFID reader at any time so that the temporal ordering of the received pairs can be performed.
 }

### 2018-08-31 ieee-tmc-latex/article.tex
final repo

--- 2018-07-10 TMC-2017-11-0695.R1-main.tex
@@
 \newcommand{\replace}[2]{\remove{#1} \add{#2}}
+
+\newcommand{\myitem}[1]{\noindent\textbf{{#1}}}
 
@@
 	\centering
-	\includegraphics[width=0.75\columnwidth]{figures/scenario.pdf}
+	\includegraphics[width=0.70\columnwidth]{figures/scenario.png}
 	\caption{\label{fig:motivation}A representative greenhouse monitoring application using CRFIDs.
@@
 
-\begin{itemize}
-	\item \textbf{Challenge I:} CRFID systems should perform computations efficiently using \emph{marginal} and intermittent RF power that leads to frequent loss of computational state, e.g.
+\myitem{Challenge I:} CRFID systems should perform computations efficiently using \emph{marginal} and intermittent RF power that leads to frequent loss of computational state, e.g.
 when the RFID reader moves away from the CRFID tag \cite{ransford:2008}.
@@
 
-	\item \textbf{Challenge II:} The continuously varying voltage supply introduces severe hardware instability, e.g.
+\myitem{Challenge II:} The continuously varying voltage supply introduces severe hardware instability, e.g.
 varying oscillator frequencies in short-term affects the stability of the clocks and degrades the accuracy sensing.
@@
 \cite{Schmid:2009}, are not applicable to compensate for the voltage instabilities in CRFIDs due to their computational complexity.
-
-\end{itemize}
 
@@
 	%
-	\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
+	\item Our testbed evaluations show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
 \end{itemize}
@@
 We note that WISP-side implementations throughout the article are done using C and assembly language.
-All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\url{https://goo.gl/RuWIZ2}.
+All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\url{https://goo.gl/aiqukc}.
 
@@
 We also present the synchronization of multiple tags in Section~\ref{sec:multi-wisp}.
-Finally, we provide our conclusions and future research directions in Section~\ref{sec:Conclusion}.
+Finally, we provide our conclusions and future work in Section~\ref{sec:Conclusion}.
 
@@
 
-\subsubsection{Clock Hardware}
-
-In WSNs, each sensor node is equipped with a built-in clock that is implemented as a counter register clocked by a low-cost external crystal oscillator.
+\myitem{Clock Hardware:} In WSNs, each sensor node is equipped with a built-in clock that is implemented as a counter register clocked by a low-cost external crystal oscillator.
 At each oscillator pulse, i.e.
@@
 
-\subsubsection{Synchronization Error}
-
-Sensor nodes exchange their clock information periodically to compute a \emph{software clock} that represents the synchronized notion of time.
+\myitem{Synchronization Error:} Sensor nodes exchange their clock information periodically to compute a \emph{software clock} that represents the synchronized notion of time.
 A software clock is composed of an offset and a frequency that hold the value and speed difference between the corresponding built-in clock and the reference time, respectively.
@@
 
-\subsubsection{State-of-the-art Synchronization Techniques}
-
-The most common synchronization mechanism is to propagate the time information of a particular \emph{reference node} to let receiver nodes synchronize themselves to the received reference time information.
+\myitem{State-of-the-art Synchronization Techniques:} The most common synchronization mechanism is to propagate the time information of a particular \emph{reference node} to let receiver nodes synchronize themselves to the received reference time information.
 Commonly, \emph{least-squares regression} is employed by several practical synchronization protocols to adjust the offset and the frequency of the software clocks ~\cite{Maroti2004,Lenzen:2015,TATS:2016}.
@@
 
-\subsubsection{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication.
+\myitem{Single-hop reader-tag architecture:} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication.
 Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}.
@@
 
-\subsubsection{Slot and bit-level synchronization} EPC Gen2 requires a slot-level synchronization to implement slotted Aloha protocol.
+\myitem{Slot and bit-level synchronization:} EPC Gen2 requires a slot-level synchronization to implement slotted Aloha protocol.
 This is achieved by adjusting clock offsets upon reception from the reader and introducing guard times to ensure that tags communicate at the
@@
 
-\subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term.
+\myitem{Continuously varying voltage level:} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term.
 On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift.
@@
 
-\subsubsection{Frequent power failures} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g.
+\myitem{Frequent power failures:} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g.
 the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up.
@@
 
-\subsubsection{Energy and resource constraints} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p.
+\myitem{Energy and resource constraints:} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p.
 44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}.
@@
 
-\subsubsection{Limitations of the EPC Gen 2 standard} The WISP firmware implements the EPC Gen 2 standard~\cite{epc_gen2} which increases the compatibility with the existing RFID systems.
+\myitem{Limitations of the EPC Gen 2 standard:} The WISP firmware implements the EPC Gen 2 standard~\cite{epc_gen2} which increases the compatibility with the existing RFID systems.
 However, the standard introduces limitations, e.g.
@@
 
-\section{WISP Hardware and CRFID Networks: A Brief Look}
+\section{WISP and CRFID Networks: A Brief Look}
 \label{sec:Clock_Hardware}
@@
 	\node[anchor=south west, inner sep=0] at 
-	(0,0){\includegraphics[width=\columnwidth]{figures/testbed_resized.jpg}};
+	(0,0){\includegraphics[width=\columnwidth]{figures/testbed_resized.png}};
 
@@
 
+\begin{figure*}
+	\centering
+	\includegraphics[width=1.4\columnwidth] {figures/relative_clock_frequency.eps}
+	\caption{\label{fig:frequency}The frequencies of the toggle events (obtained using the same steps presented in Fig.\ref{fig:frequency_dc_rf}) of two RF-powered WISP tags, placed at 5 cm distance from the RFID reader.
+On the left the event frequencies ($f_1$ and $f_2$) are presented, and their relative clock frequency ($\frac{f_1}{f_2}-1$) is presented on the right.
+We measured the mean and standard deviation of $f_1$ as 1.1793 and 0.00026049 and those of $f_2$ as 1.1705 and  0.00068239, respectively.}
+\end{figure*}
+
 \subsection{\label{sec:sniffer}A Sniffer for the CRFID System}
@@
 We will present these measurements in the next sections.
-\section{System Model, Notation and Preliminaries}
+
+\section{System Model and Preliminaries}
 \label{sec:clock_model}
@@
 We will use these models to prove the correctness of our synchronization algorithms and to derive mathematical expressions that represent their synchronization performance.
-\begin{figure*}
-	\centering
-	\includegraphics[width=1.4\columnwidth] {figures/relative_clock_frequency.eps}
-	\caption{\label{fig:frequency}The frequencies of the toggle events (obtained using the same steps presented in Fig.\ref{fig:frequency_dc_rf}) of two RF-powered WISP tags, placed at 5 cm distance from the RFID reader.
-On the left the individual event frequencies ($f_1$ and $f_2$) are presented, and their relative clock frequency ($\frac{f_1}{f_2}-1$) is presented on the right.
-We measured the mean and standard deviation of $f_1$ as 1.1793 and 0.00026049 and those of $f_2$ as 1.1705 and  0.00068239, respectively.}
-\end{figure*}
+
 In this framework, we denote a single WISP tag by $w$ and the RFID reader by $r$.
@@
 By collecting (time) information from the RFID reader, the tag $w$ runs a \emph{clock synchronization algorithm} to calculate a \emph{software clock}, denoted by $S_{w}(t)$, whose value represents the synchronized notion of time.
-The objective of the synchronization algorithm is minimizing the \emph{synchronization error} with respect to the reader $r$ at any time $t$, which is defined as
+The objective of the synchronization algorithm is minimizing the \emph{synchronization error} with respect to the reader $r$ at any time $t$, defined as
 %
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/Sender-Receiver.pdf}
+	\includegraphics[width=\columnwidth]{figures/Sender-Receiver.png}
 	\caption{\label{fig:sender_receiver} The message exchange among the host computer, the RFID reader and the tag for sender-receiver synchronization.
@@
 
-\subsection{Transmission Delay and Jitter Measurements}
-
 \begin{figure}
@@
 
-In order to characterize the transmission delay $\Delta t$, we placed the WISP tag 20 cm away from the antenna using the setup presented in Fig.~\ref{fig:crfid_testbed} and we sniffed the communication between the RFID reader and the WISP tag during the communication scenario of Fig.~\ref{fig:sender_receiver}.
+\myitem{Transmission Delay and Jitter Measurements:} In order to characterize the transmission delay $\Delta t$, we placed the WISP tag 20 cm away from the antenna using the setup presented in Fig.~\ref{fig:crfid_testbed} and we sniffed the communication between the RFID reader and the WISP tag during the communication scenario of Fig.~\ref{fig:sender_receiver}.
 Fig.~\ref{fig:transmission_delay} presents a summary of our measurements (with 300 samples gathered).
@@
 
-\subsection{Collection of Clock Values for Off-line Processing}
-
-After characterizing the variation of the transmission delay, the next step was the collection of the clock values from the RFID Reader and the WISP tag in order to try synchronization mechanisms offline, which allowed to evaluate the performance of sender-receiver synchronization mechanism in CRFID networks.
+\myitem{Collection of Clock Values for Off-line Processing:} After characterizing the variation of the transmission delay, the next step was the collection of the clock values from the RFID Reader and the WISP tag in order to try synchronization mechanisms offline, which allowed to evaluate the performance of sender-receiver synchronization mechanism in CRFID networks.
 We collected $(C_w(t_2),C_r(t_1))$ pairs for offline processing by controlling the RFID reader to send a \emph{Read} command to the tag and by programming the WISP tag so that it backscatters $C_w(t_2)$ upon receiving this command.
@@
 
-\subsubsection{Host computer computation} We are unaware of any EPC Gen2 command that will allow to send the \emph{FistSeenTimestamp} to the tag.
+\myitem{Host computer computation:} We are unaware of any EPC Gen2 command that will allow to send the \emph{FistSeenTimestamp} to the tag.
 Hence, even though we were able to collect $(C_w(t),C_r(t))$ pairs for offline processing, it is not possible for the tag to collect $C_r(t)$ and synchronize itself to the reader.
@@
 
-\subsubsection{The Lack of broadcast primitive} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader.
+\myitem{The Lack of broadcast primitive:} Since the RFID reader assigns the \emph{FirstSeenTimestamp} for each tag, the synchronization steps in Fig.~\ref{fig:sender_receiver} should be repeated for each tag to obtain synchronization in the communication domain of the reader.
 %Unfortunately, in the current EPC Gen2 and LLRP standard, we were unable to find any mechanism to send a global timestamp that is received by all tags inside the range of the RFID reader to establish synchronization in one step.
 
-\subsubsection{Big and reader dependent transmission delays} As depicted previously, we observed that transmission delays in our testbed were on the order of milliseconds, see Fig.~\ref{fig:least-squares}, whereas they are reported as a few microseconds in conventional WSN platforms.
-This is one of the crucial limitations since the smaller the transmission delays are, the better the synchronization.
+\myitem{Big and reader dependent transmission delays:} As depicted previously, we observed that transmission delays in our testbed were on the order of milliseconds, see Fig.~\ref{fig:least-squares}, whereas they are reported as a few microseconds in WSN platforms.
+This is a crucial limitation since the smaller the transmission delays are, the better the synchronization.
 
@@
 	\centering
-	\includegraphics[width=\columnwidth]{figures/Event-Based.pdf}
+	\includegraphics[width=\columnwidth]{figures/Event-Based.png}
 	\caption{\label{fig:event_based}Event-Based synchronization steps: The tag timestamps successive \emph{BlockWrite} events and adjusts its software clock.}
@@
 
-\subsection{Event-Based Synchronization Inspired By PI Controllers}
+\subsection{Event-Based Synchronization Using PI Controllers}
 
@@
 We refer again the reader to the Appendix for the detailed proofs of the convergence and the asymptotic estimation error.
-Moreover, in Section~\ref{sec:adaptibility}, we discuss the impact of the parameters $\tau$ and $\beta$ on the synchronization accuracy.
-
-\subsubsection{The Synchronization Algorithm}
+%Moreover, in Section~\ref{sec:adaptibility}, we discuss the impact of the parameters $\tau$ and $\beta$ on the synchronization accuracy.
 
@@
 
-Based on this model we introduce Algorithm \ref{alg:sync}, namely \emph{WISP-Sync}, that synchronizes the speed of the software clock to the reference clock's speed using the update rule defined in \eqref{eq:update_rule}.
+\myitem{The Synchronization Algorithm:} Based on this model we introduce Algorithm \ref{alg:sync}, namely \emph{WISP-Sync}, that synchronizes the speed of the software clock to the reference clock's speed using the update rule defined in \eqref{eq:update_rule}.
 The steps of this algorithm can be explained as follows.
@@
 After receiving successive events, $\hat{f}_w^r$ will converge to its desired value eventually, as proven in the Appendix.
-
-\subsubsection{State Recovery}
-
-Due to intermittent RF power, the tag is subject to frequent power losses, which will lead to loss of the synchronization state.
-In voltage-aware systems~\cite{Buettner:2011:Dewdrop,Ransford:2011:Mementos,Lucia:2015:Execution,Colin:2016:Debugger}, the general idea is to save the system state when the input voltage is below a pre-defined threshold $V_{\min}$ and then recover it back when the input voltage is above a pre-defined threshold $V_{\text{op}}$.
-As shown in Algorithm~\ref{alg:sync}, the only part of the system state pertaining to synchronization is the variable $\hat{f}_w^r$.
-Therefore, if this variable is also saved to FRAM (when there is a power loss and recovered when there is sufficient voltage) the synchronization will resume from the point of power loss, that will prevent loss of synchronization and lead to improved synchronization accuracy.
-These steps are summarized in Lines 6--9 of Algorithm~\ref{alg:sync}.
-It should be noted that these steps require hardware support so that the voltage level is monitored continuously and the corresponding voltage level indicator events are generated.
-Another approach can be keeping $\hat{f}_w^r$ in non-volatile memory as a \emph{persistent variable} so that the corresponding FRAM memory location of $\hat{f}_w^r$ is updated immediately when this variable is updated---eliminating the need for Lines 6--9 of Algorithm~\ref{alg:sync}, and in turn extra voltage measurement hardware, but introducing more FRAM access overhead.
-
-\subsubsection{Computation Complexity\label{subsec:complexity}}
-
-It can be observed that WISP-Sync employs only very simple arithmetic operations at each step, e.g.
-Lines 3--5 are composed of only two subtraction, two addition and two multiplication operations.
-Moreover, the \emph{state of synchronization}, i.e.
-the parameters that are required to be saved in the non-volatile memory, is represented by only the variable $\hat{f}_w^r$.
-As a consequence, WISP-Sync addresses \emph{Challenge I} in Section~\ref{sec:Motivation} in terms of computation and memory overhead.
-
-\subsubsection{Adaptability} 
-\label{sec:adaptibility}
 
@@
 
-In Section \ref{subsec:event_period_uncertainty} we presented our measurements pertaining to the event period and showed 
+\myitem{State Recovery:} Due to intermittent RF power, the tag is subject to frequent power losses, which will lead to loss of the synchronization state.
+In voltage-aware systems~\cite{Buettner:2011:Dewdrop,Ransford:2011:Mementos,Lucia:2015:Execution,Colin:2016:Debugger}, the general idea is to save the system state when the input voltage is below a pre-defined threshold $V_{\min}$ and then recover it back when the input voltage is above a pre-defined threshold $V_{\text{op}}$.
+As shown in Algorithm~\ref{alg:sync}, the only part of the system state pertaining to synchronization is the variable $\hat{f}_w^r$.
+Therefore, if this variable is also saved to FRAM (when there is a power loss and recovered when there is sufficient voltage) the synchronization will resume from the point of power loss, that will prevent loss of synchronization and lead to improved synchronization accuracy.
+These steps are summarized in Lines 6--9 of Algorithm~\ref{alg:sync}.
+It should be noted that these steps require hardware support so that the voltage level is monitored continuously and the corresponding voltage level indicator events are generated.
+Another approach can be keeping $\hat{f}_w^r$ in non-volatile memory as a \emph{persistent variable} so that the corresponding FRAM memory location of $\hat{f}_w^r$ is updated immediately when this variable is updated---eliminating the need for Lines 6--9 of Algorithm~\ref{alg:sync}, and in turn extra voltage measurement hardware, but introducing more FRAM access overhead.
+
+\myitem{Computation Complexity:} It can be observed that WISP-Sync employs only very simple arithmetic operations at each step, e.g.
+Lines 3--5 are composed of only two subtraction, two addition and two multiplication operations.
+Moreover, the \emph{state of synchronization}, i.e.
+the parameters that are required to be saved in the non-volatile memory, is represented by only the variable $\hat{f}_w^r$.
+As a consequence, WISP-Sync addresses \emph{Challenge I} in Section~\ref{sec:Motivation} in terms of computation and memory overhead.
+
+\myitem{Adaptability:} In Section \ref{subsec:event_period_uncertainty} we presented our measurements pertaining to the event period and showed 
 that $\sigma_{\tau}=0.4097$.
@@
 	\caption{\label{fig:sync_error}Event-based synchronization performance when the tag is powered through a constant voltage source and through RF power.
-Mean synchronization errors without and with WISP-Sync were almost 127 and 16 clock ticks under stable voltage, and 175 and 22 clock ticks with RF power harvesting (represented by the straight lines in this figure), respectively.}
+Mean synchronization errors without and with WISP-Sync were almost 127 and 16 clock ticks under stable voltage, and 175 and 22 clock ticks with RF power harvesting (represented by the straight lines), respectively.}
 \end{figure}
@@
 We repeated this experiment ten times to eliminate quantization effects during measurements.
-We observed an average energy of $\approx1.85 \mu$J with a standard deviation of 0.038\,$\mu$J is spent by the tag in order to perform the aforementioned computations.
+We observed that the tag spends an average energy of $\approx1.85 \mu$J with a standard deviation of 0.038\,$\mu$J to perform the aforementioned computations.
 
@@
 
-\section{The Synchronization of Multiple WISP Tags Using the WISP-Sync}
+\section{The Synchronization of Multiple WISPs}
 \label{sec:multi-wisp}
@@
 	\node[anchor=south west, inner sep=0] at 
-	(0,0){\includegraphics[width=0.8\columnwidth]{figures/3-wisps_resized.jpg}};
+	(0,0){\includegraphics[width=0.8\columnwidth]{figures/3-wisps_resized.png}};
 
@@
 the maximum pairwise difference between the port clear times) when we considered all of the tags; (ii) the red line depicts when only synchronized tags are considered.
-From this figure it can be concluded that synchronization led to a factor of approximately four times better prediction on average among the tags: we observed an average synchronization error of 1\,ms while it was 4\,ms when one of the tags was unsynchronized.
+As seen from this figure, synchronization led to a factor of approximately four times better prediction on average among the tags: we observed an average synchronization error of 1\,ms while it was 4\,ms when one of the tags was unsynchronized.
 
-\section{Conclusions and Future Research Directions}
+\section{Conclusions and Future Work}
 \label{sec:Conclusion}
@@
 
-\begin{itemize}
-	\item \textbf{Network-wide synchronization:} We studied event-based synchronization among many tags.
+\myitem{Network-wide synchronization:} We studied event-based synchronization among many tags.
 However, synchronization of these tags where they can access the explicit clock of the RFID reader and also synchronization of the whole CRFID network composed of several RFID readers and tags is still an issue, due to the single-hop nature of backscatter communication.
@@
 
-	\item \textbf{Power loss and recovery:} It is crucial to save the synchronization status to non-volatile memory just before the power loss, but writing to non-volatile memory also consumes energy.
-Hence, when to save the synchronization status without extra hardware support is worth exploring~\cite{Buettner:2011:Dewdrop,Ransford:2011:Mementos,Lucia:2015:Execution,Colin:2016:Debugger}.
+\myitem{Power loss and recovery:} It is crucial to save the synchronization status to non-volatile memory just before the power loss~\cite{Ransford:2011:Mementos}, but this operation introduces energy overhead.
+Hence, when to save the synchronization status (without extra hardware support) is worth exploring.
 
-	\item \textbf{Voltage-frequency relations:} It might be interesting to explore the voltage-clock frequency relationship since1 tags are subject to varying voltage source when they are powered with only RF energy harvesting from the RFID reader.
+\myitem{Voltage-frequency relations:} It might be interesting to explore the voltage-clock frequency relationship since1 tags are subject to varying voltage source when they are powered with only RF energy harvesting from the RFID reader.
 We anticipate that voltage level should be incorporated to the establishment of the software clock so that voltage dependent instability of clock frequency can be compensated.
@@
 
-	\item \textbf{Synchronization in other IPD platforms:} Synchronization in other IPD platforms, e.g. that use ambient backscatter communication \cite{ambient_backscatter}, is worth exploring as well. 
-\end{itemize}
+\myitem{Synchronization in other IPD platforms:} Synchronization in other IPD platforms, e.g. that use ambient backscatter communication \cite{ambient_backscatter}, is worth exploring as well.
 
@@
 
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/sinan.jpg}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/sinan.png}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
 received his Ph.D.
 degree in Computer Engineering from Ege University, Izmir, Turkey, in 2012.
-He was a visiting scholar at the Department of Information Engineering, University of Padova in 2013.
-Between 2015 and 2017, he was a postdoctoral researcher at Embedded Software Group, Delft University of Technology, Delft, The Netherlands.
+He was a post-doc at the Department of Information Engineering, University of Padova in 2013 and at Embedded Software Group, Delft University of Technology between 2015 and 2017.
 Currently, he is an assistant professor at the Department of Computer Engineering, Ege University and also a research affiliate with the Embedded Software Group at Delft University of Technology.
-His research interests are embedded systems and software, distributed algorithms, wireless sensor networks, energy harvesting systems and computational RFIDs.
+His research interests are embedded systems and software, distributed algorithms and wireless networking.
 \end{IEEEbiography}
@@
 % if you will not have a photo at all:
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/henko.jpg}}]{Henko Aantjes}
-Henko Aantjes received his MSc degree in 2016 from Delft University of Technology, Delft, The Netherlands.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/henko.png}}]{Henko Aantjes}
+received his MSc degree in 2016 from Delft University of Technology, Delft, The Netherlands.
 Currently he is a security evaluator at Brightsight BV, Delft, The Netherlands.
@@
 %
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.jpg}}]{Przemys{\l}aw~Pawe{\l}czak} (S'03-M'10) received the MSc degree from Wroc{\l}aw University of Technology, Poland, in 2004 and the PhD degree from Delft University of Technology, The Netherlands in 2009.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/przemek.png}}]{Przemys{\l}aw~Pawe{\l}czak} (S'03-M'10) received the MSc degree from Wroc{\l}aw University of Technology, Poland, in 2004 and the PhD degree from Delft University of Technology, The Netherlands in 2009.
  Between 2009 and 2011 he was a postdoctoral researcher at the Cognitive Reconfigurable Embedded Systems Lab, University of California, Los Angeles.
@@
 %
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.jpg}}]{Amjad Yousef Majid}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/amjad.png}}]{Amjad Yousef Majid}
 obtained his MSc degree in 2015 from Delft University of Technology, Delft, The Netherlands.

### 2018-08-31 source codes/data/matlab/blockwrite_delay.m

@@ -1,0 +1,27 @@
+clear all;
+close all;
+fileID = fopen('testbed data/blockwrite_delay.txt','r');
+[A] = fscanf(fileID, ['%d'])
+A = A/10000; % from 10MHZ sample rate to milliseconds
+
+Diff = [];
+m = 1;
+while (m+8)<numel(A)
+    
+    X = A(m);
+    Y = A(m+7);
+    diff = Y-X;
+    m = m+8;
+    Diff = [Diff diff];
+end
+hFig = figure(1)
+h = histogram(Diff,'Normalization','probability');
+h.NumBins = 20;
+
+[muhat,sigmahat,muci,sigmaci] = normfit(Diff,0.01)
+xlabel('Delay (ms)')
+ylabel('Normalized num. of occurence')
+set(gca,'FontSize',18)
+set(hFig, 'Position', [0 0 600 300])
+grid on
+

### 2018-08-31 source codes/data/matlab/integral_gain.m

@@ -1,0 +1,77 @@
+clear all;
+fid = fopen('testbed data/stable_voltage.txt','r');
+
+
+Period = 7118.0;   % Period of the event
+Elapsed = 0.0;  % The hardware time of the next event
+
+alpha1 = 1.0;     % Integral gain
+beta1 = 0.0002;
+
+alpha2 = 1.0;     % Integral gain
+beta2 = 0.0001;
+
+alpha3 = 1.0;     % Integral gain
+beta3 = 0.00005;
+
+m=1;
+
+
+while(~feof(fid))
+  A = fgetl(fid);
+  HC = hex2dec(A(1:4));     % Read hardware clock 
+ 
+  Elapsed = HC;  % Expected next event time
+    
+  for k = 6:6
+    HC = hex2dec(A(k*4+1:k*4+4));   % The hardware time of current event
+    if(HC < Elapsed)
+        Elapsed = HC + 2^16 - Elapsed;
+    else
+        Elapsed = HC - Elapsed;
+    end
+    
+    E(m) = Period - Elapsed;             % The error between expected and actual time
+    
+    E1(m) = alpha1*Elapsed - Period;
+    alpha1 = alpha1 - beta1*E1(m)     
+    
+    E2(m) = alpha2*Elapsed - Period;
+    alpha2 = alpha2 - beta2*E2(m)     
+    
+    E3(m) = alpha3*Elapsed - Period;
+    alpha3 = alpha3 - beta3*E3(m)    
+    
+    m=m+1;
+       
+    end
+end
+
+E = abs(E);
+E1 = abs(E1);
+E2 = abs(E2);
+E3 = abs(E3);
+
+hFig = figure(1)
+plot(E1,'r--','LineWidth',1.5);
+hold on
+plot(E2,'b','LineWidth',1.5);
+hold on
+plot(E3,':','LineWidth',2);
+
+
+
+
+xlim([0 50]);
+ylim([0 80]);
+
+xlabel('Samples')
+ylabel('\gamma (clock ticks)');
+%ylabel(sprintf('$\gamma$ \n (clock ticks)'))
+legend('beta=0.0002', 'beta=0.0001','beta=0.00005')
+set(gca,'FontSize',18)
+set(hFig, 'Position', [0 0 600 300])
+grid on
+
+% Save Figure
+print(hFig,'integral_gain','-depsc2')

### 2018-08-31 source codes/edb/energy.m

@@ -1,0 +1,12 @@
+clear all;
+close all;
+
+measurements = [2.2147-1.9319 2.2111-1.9312 2.2198-1.9400 2.2242-1.9407 ...
+    2.2234-1.9451 2.2285-1.9472 2.1914-1.9101 2.2293-1.9472 2.2293-1.9472 ...
+    2.2293-1.9472 2.2227-1.9465 2.1928-1.9115 2.2227-1.9465 2.2162-1.9421]
+
+capacitor = 47*10e-7
+
+joule = 1/2*capacitor*(measurements.^2)
+mean(joule)
+std(joule)

### 2018-09-07 ieee-tmc-latex/article.tex
image licence

@@
 	\caption{\label{fig:motivation}A representative greenhouse monitoring application using CRFIDs.
-Two readers are deployed to provide power to the individual tags and to collect their periodic sensor readings.}
+Two readers are deployed to provide power to the individual tags and to collect their periodic sensor readings.
+(Image by Joi Ito, source~\cite{greenhouse-image} licensed under CC BY 2.0.) }
 \end{figure}
@@
 
-\myitem{Frequent power failures:} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g.
-the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up.
-However, saving computational state is also an energy consuming task~\cite{blisp}.
+\myitem{Frequent power failures:} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss.
+Therefore, they need to save the synchronization state, e.g.
+the clock offset and relative frequency, into the non-volatile memory to recover when they harvest sufficient energy to start up.
+However, this is also an energy consuming task~\cite{blisp}.
 
@@
 To program the WISP tag, we used MSP430 Flash Emulation Tool (FET), in combination with TI Code Composer Studio (CCS), attached to the host.
-All of our experimental evaluation presented in the next sections were conducted using this setup at a university office with human presence.
+Our experimental evaluations presented hereafter were conducted using this setup at a university office with human presence.
 
@@
 By means of this, we could  use the external 32\,KHz clock that runs even in low-power standby mode (see Section \ref{sec:Clock_Hardware}) to obtain timing measurements.)} Fig.~\ref{fig:event_based} presents the steps of event-based synchronization.
-During the command phase, the EPC Gen2 standard allows a maximum of eight successive \emph{BlockWrite} operations.
+During the command phase, the EPC Gen2 allows a maximum of eight successive \emph{BlockWrite} operations.
 It is desirable to use the first and the last \emph{BlockWrite} events for synchronization since it is better to compensate frequency differences observed in longer time intervals to adjust the software clock.
@@
 According to our measurements, the \emph{event period} was distributed with a mean of 226.76\,ms and standard deviation of 0.41\,ms; respectively.
-We would like to mention that these values are dependent on the RFID Reader implementation of the EPC Gen2 protocol.
+We would like to mention that these values are dependent on the RFID Reader implementation of the EPC Gen2.
 
@@
 
-Let $t_0$ denote the first event reception time at the tag. Using the pre-measured event period $\tau$ of the RFID reader, the tag can estimate the next event reception time at time $t_1$ as
+Let $t_0$ denote the first event reception time at the tag. Using the pre-measured event period $\tau$ of the reader, the tag can estimate the next event reception time at time $t_1$ as
 \begin{eqnarray}
@@
 He was a post-doc at the Department of Information Engineering, University of Padova in 2013 and at Embedded Software Group, Delft University of Technology between 2015 and 2017.
-Currently, he is an assistant professor at the Department of Computer Engineering, Ege University and also a research affiliate with the Embedded Software Group at Delft University of Technology.
-His research interests are embedded systems and software, distributed algorithms and wireless networking.
+Currently, he is an assistant professor at the Department of Computer Engineering, Ege University and also a member of the Embedded Software Group at Delft University of Technology.
+His research interests are embedded systems, wireless networks and distributed algorithms.
 \end{IEEEbiography}

### 2018-09-09 ieee-tmc-latex/article.tex

@@
 We note that WISP-side implementations throughout the article are done using C and assembly language.
-All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\url{https://goo.gl/aiqukc}.
+All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\url{https://github.com/TUDSSL/wispsync}.
 

### 2018-09-09 ieee-tmc-latex/supplemental.tex

--- 2018-05-17 TMC-appendix.tex
@@
 
-\title{On the Synchronization of Computational RFIDs}
+\title{On the Synchronization of Computational RFIDs \\
+(Supplemental Material)}
 
@@
 		%\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}. \protect\\
-		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).
+	\IEEEcompsocthanksitem Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m, Przemys{\l}aw Pawe{\l}czak and Amjad Yousef Majid are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl).
+	\IEEEcompsocthanksitem Henko Aantjes is with the Brighsight BV, Brassersplein 2, 2612\,CT, Delft, The Netherlands (e-mail: aantjes@brightsight.com)
+	\IEEEcompsocthanksitem Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m is also affilated with the Computer Engineering Department of Ege University, Izmir, Turkey.
 	}

### 2018-09-09 ieee-tmc-latex/article.tex
typo and extra clarification on the code

--- 2018-09-07 ieee-tmc-latex/article.tex
@@
 	Henko~Aantjes, 
-	Przemys{\l}aw~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
+	Przemys{\l}aw~Pawe{\l}czak,~\IEEEmembership{Member,~IEEE,} 
 	and~Amjad~Yousef~Majid
@@
 We note that WISP-side implementations throughout the article are done using C and assembly language.
-All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\url{https://goo.gl/aiqukc}.
+All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available upon request or via~\url{https://github.com/TUDSSL/wispsync}.
 

### 2018-09-09 matlab/blockwrite_delay.m
final

@@ -1,27 +1,0 @@
-clear all;
-close all;
-fileID = fopen('testbed data/blockwrite_delay.txt','r');
-[A] = fscanf(fileID, ['%d'])
-A = A/10000; % from 10MHZ sample rate to milliseconds
-
-Diff = [];
-m = 1;
-while (m+8)<numel(A)
-    
-    X = A(m);
-    Y = A(m+7);
-    diff = Y-X;
-    m = m+8;
-    Diff = [Diff diff];
-end
-hFig = figure(1)
-h = histogram(Diff,'Normalization','probability');
-h.NumBins = 20;
-
-[muhat,sigmahat,muci,sigmaci] = normfit(Diff,0.01)
-xlabel('Delay (ms)')
-ylabel('Normalized num. of occurence')
-set(gca,'FontSize',18)
-set(hFig, 'Position', [0 0 600 300])
-grid on
-

### 2018-09-09 matlab/frequency_testbed.m

@@ -1,150 +1,0 @@
-function frequency_testbed
-    clear all;
-    close all;
-
-    % Open the comma seperated data file - [Timestamp Port]
-    fileID = fopen('testbed data/w1_DC.csv');
-    file = textscan(fileID,'%f %d', 'Delimiter', ',');
-    fclose(fileID);
-
-    wisp1_DC_t  = cell2mat(file(1,1));
-    
-    fileID = fopen('testbed data/w2_DC.csv');
-    file = textscan(fileID,'%f %d', 'Delimiter', ',');
-    fclose(fileID);
-    
-    wisp2_DC_t  = cell2mat(file(1,1));
-
-    fileID = fopen('testbed data/w1_RF.csv');
-    file = textscan(fileID,'%f %d', 'Delimiter', ',');
-    fclose(fileID);
-
-    wisp1_RF_t  = cell2mat(file(1,1));
-    
-    fileID = fopen('testbed data/w2_RF.csv');
-    file = textscan(fileID,'%f %d', 'Delimiter', ',');
-    fclose(fileID);
-
-    wisp2_RF_t  = cell2mat(file(1,1));
-    %------------------------------------------------------
-
-    drift_dc = get_drift(wisp1_DC_t);
-    drift_rf = get_drift(wisp1_RF_t);
-    
-    freq_dc = get_freq(wisp1_DC_t);
-    freq_rf = get_freq(wisp1_RF_t);
-    
-    freq_dc_2 = get_freq(wisp2_DC_t);
-    freq_rf_2 = get_freq(wisp2_RF_t);
-    
-    %------------------------------------------------------
-    hFig = figure('DefaultAxesFontSize',18);
-    subplot(1,2,2);
-    plot(1:size(drift_dc),drift_dc,'DisplayName','DC Voltage','LineWidth',1.5);
-    %hold on
-    %plot(1:size(drift_rf),drift_rf,'DisplayName','RF Power','LineWidth',1.5);
-    xlabel('Time (second)')
-    ylabel('Drift (microsec.)')
-    xlim([1 100])   
-    %set(gca,'YTick',[1:10]); 
-    grid on
-    %legend('show')
-    %set(hFig, 'Position', [0 0 600 300])
-    
-    subplot(1,2,1);
-    %hFig = figure('DefaultAxesFontSize',18);
-    plot(1:size(freq_dc),freq_dc,'DisplayName','DC Voltage','LineWidth',1.5);
[…]

### 2018-09-09 matlab/integral_gain.m

@@ -1,77 +1,0 @@
-clear all;
-fid = fopen('testbed data/stable_voltage.txt','r');
-
-
-Period = 7118.0;   % Period of the event
-Elapsed = 0.0;  % The hardware time of the next event
-
-alpha1 = 1.0;     % Integral gain
-beta1 = 0.0002;
-
-alpha2 = 1.0;     % Integral gain
-beta2 = 0.0001;
-
-alpha3 = 1.0;     % Integral gain
-beta3 = 0.00005;
-
-m=1;
-
-
-while(~feof(fid))
-  A = fgetl(fid);
-  HC = hex2dec(A(1:4));     % Read hardware clock 
- 
-  Elapsed = HC;  % Expected next event time
-    
-  for k = 6:6
-    HC = hex2dec(A(k*4+1:k*4+4));   % The hardware time of current event
-    if(HC < Elapsed)
-        Elapsed = HC + 2^16 - Elapsed;
-    else
-        Elapsed = HC - Elapsed;
-    end
-    
-    E(m) = Period - Elapsed;             % The error between expected and actual time
-    
-    E1(m) = alpha1*Elapsed - Period;
-    alpha1 = alpha1 - beta1*E1(m)     
-    
-    E2(m) = alpha2*Elapsed - Period;
-    alpha2 = alpha2 - beta2*E2(m)     
-    
-    E3(m) = alpha3*Elapsed - Period;
-    alpha3 = alpha3 - beta3*E3(m)    
-    
-    m=m+1;
-       
-    end
-end
-
-E = abs(E);
-E1 = abs(E1);
-E2 = abs(E2);
-E3 = abs(E3);
-
-hFig = figure(1)
-plot(E1,'r--','LineWidth',1.5);
-hold on
-plot(E2,'b','LineWidth',1.5);
-hold on
-plot(E3,':','LineWidth',2);
-
-
-
-
-xlim([0 50]);
-ylim([0 80]);
-
-xlabel('Samples')
-ylabel('\gamma (clock ticks)');
-%ylabel(sprintf('$\gamma$ \n (clock ticks)'))
-legend('beta=0.0002', 'beta=0.0001','beta=0.00005')
-set(gca,'FontSize',18)
-set(hFig, 'Position', [0 0 600 300])
-grid on
-
-% Save Figure
-print(hFig,'integral_gain','-depsc2')

### 2018-09-09 matlab/least_squares.m

@@ -1,31 +1,0 @@
-clear all;
-close all;
-
-[X Y] = textread('testbed data/least_squares.txt', '%d %d', 'delimiter', ' ')
-
-mean_delay = 59; % the mean transmission delay we measured
-
-Error = [];
-m = 9;
-while (m+8)<numel(X)
-    
-    A = X(m:m+7);
-    B = Y(m:m+7);
-    p = polyfit(A,B,1);
-    error = Y(m+8)-polyval(p,X(m+8))
-    m = m+8;
-    Error = [Error error];
-end
-Error(Error > 10) = [];
-Error(Error < -10) = [];
-
-hFig = figure(1)
-
-histogram(Error);
-
-xlabel('Synchronization error (clock ticks)')
-ylabel('Num. of occurence')
-set(gca,'FontSize',18)
-ylim([0 90]);
-set(hFig, 'Position', [0 0 600 300])
-grid on

### 2018-09-09 matlab/multiwisp.m

@@ -1,128 +1,0 @@
-function simulations
-    clear all;
-    close all;
-    
-    [x_all y_all x_2 y_2]= calculate_sync_error();
-    
-    hFig = figure('DefaultAxesFontSize',18);
-    plot(x_all,y_all,':','DisplayName','Max. \gamma (3 Wisps)','LineWidth',2);
-    hold on 
-    plot(x_2,y_2,'DisplayName','Max. \gamma (2 Wisps)','LineWidth',1.5);
-    hold on 
-    plot(x_all,repmat(mean(y_all),numel(x_all),1),'--','DisplayName','Avg. \gamma (3 Wisps)','LineWidth',1);
-    hold on
-    plot(x_2,repmat(mean(y_2),numel(x_2),1),'-.','DisplayName','Avg. \gamma (2 Wisps)','LineWidth',1);
-    xlabel('Time (second)')
-    ylabel('\gamma (ms)')
-    %set(gca,'YTick',[1:10]);
-    grid on
-    legend('show')
-    set(hFig, 'Position', [0 0 600 300])
-    ylim([0 10])
-    xlim([4 max(x_all)])
-    
-    print(hFig,'multiwisp-experiments','-depsc2')
-end
-
-function [x1 y1 x2 y2] = calculate_sync_error()
-    % Open the comma seperated data file - [Timestamp Ch1 Ch2 Ch3]
-    fileID = fopen('testbed data/3w.csv');
-    file = textscan(fileID,'%f %d %d %d', 'Delimiter', ',');
-    fclose(fileID);
-
-    timestamps = 1000.0*cell2mat(file(1,1)); % in ms
-    wisp1 = cell2mat(file(1,2));
-    wisp2 = cell2mat(file(1,3));
-    wisp3 = cell2mat(file(1,4));
-
-    % reset on and off times
-    starts = zeros(3,1);
-    ends =  zeros(3,1);
-
-    x1 = [];
-    y1 = [];
-    
-    x2 = [];
-    y2 = [];
-
-    for i = 1:size(timestamps)
-        
-        
-
-        % save on and off times of the ports
-        if(starts(1) == 0 && wisp1(i) == 1)
-            starts(1) = timestamps(i);
-        elseif (starts(1) > 0 && ends(1) == 0 && wisp1(i) == 0)
-            ends(1) = timestamps(i);
-        end
-
-        if(starts(2) == 0 && wisp2(i) == 1)
-            starts(2) = timestamps(i);
[…]

### 2018-09-09 matlab/power_loss.m

@@ -1,78 +1,0 @@
-clear all;
-close all;
-fid = fopen('testbed data/stable_voltage.txt','r');
-
-Period = 7086.0;   % Period of the event
-Elapsed = 0.0;  % The hardware time of the next event
-
-alpha = 1.0;     % Integral gain
-alpha2 = 1.0;     % Integral gain
-beta = 0.0001;
-sample = 0;
-
-m=1;
-
-while(~feof(fid))
-  A = fgetl(fid);
-  HC = hex2dec(A(1:4));     % Read hardware clock 
- 
-  Elapsed = HC;  % Expected next event time
-  sample = sample + 1;
-  if((10<sample && sample<20) ||(30<sample && sample<40))
-      E1(m) = 0;
-      m=m+1;
-      alpha2 = 1.0;
-      continue
-  end
-   
-  for k = 6:6
-      
-    HC = hex2dec(A(k*4+1:k*4+4));   % The hardware time of current event
-    
-    if(HC < Elapsed)
-        Elapsed = HC + 2^16 - Elapsed;
-    else
-        Elapsed = HC - Elapsed;
-    end
-        
-    E1(m) = alpha*Elapsed - Period;
-    E2(m) = alpha2*Elapsed - Period;
-    alpha = alpha - beta*E1(m) ;    
-    alpha2 = alpha2 - beta*E2(m) ;    
-    
-    m=m+1;
-    
-  end
-end
-
-E1 = abs(E1);
-E1(E1 > 65000) = [];
-mean(E1)
-E2 = abs(E2);
-E2(E2 > 65000) = [];
-mean(E2)
-
-
-hFig = figure(1);
-
-plot(E2,'r-.','LineWidth',1.7),
-hold on
-plot(E1,'b');
-hold on;
-text(11,5,'Power Loss','FontSize',13)
-hold on
-text(30.5,5,'Power Loss','FontSize',13)
-
-
-xlim([0 50]);
-ylim([0 120]);
-
-xlabel('Samples');
-ylabel('\gamma (clock ticks)');
-h_legend = legend('w/o State Recovery', 'State Recovery');
-set(h_legend,'FontSize',15);
-grid on
-set(gca,'FontSize',18);
-set(hFig, 'Position', [0 0 600 300]);
-
-print(hFig,'power_loss','-depsc2')

### 2018-09-09 matlab/sender_receiver_uncertainty.m

@@ -1,31 +1,0 @@
-clear all;
-close all;
-fileID = fopen('testbed data/sender_receiver_uncertainty.txt','r');
-[A] = fscanf(fileID, ['%d']);
-A = A/10000; % from 10MHZ sample rate to milliseconds
-
-hFig = figure(1);
-
-h = histogram(A,'Normalization','probability');
-h.NumBins = 50;
-xlim([1.80 2.4]);
-
-hold on 
-A(A > 2.1) = [];
-A(A < 1.8) = [];
-
-[muhat,sigmahat,muci,sigmaci] = normfit(A,0.01)
-pd = fitdist(A,'Normal');
-x_values = 1.4:0.001:2.4;
-y = pdf(pd,x_values);
-y = y ./h.NumBins;
-plot(x_values,y,'LineWidth',2)
-
-ylim([0 0.5]);
-
-xlabel('Delay (ms)')
-ylabel('Normalized num. of occurence')
-legend('Frequency of the transmission delay', 'Normal fit function')
-set(gca,'FontSize',18)
-set(hFig, 'Position', [0 0 600 300])
-grid on

### 2018-09-09 matlab/simulations.m

@@ -1,170 +1,0 @@
-function plot()
-
-    clear all;
-    close all;
-    
-
-    
-    %----------------------------------------------------
-    % theory simulations consistency
-    %----------------------------------------------------
-    eta =0.0002;
-    d = 0.00005;
-    [x2,y2,error2] = theory_simulations(eta,d);
-    eta = 0.0008;
-    d = 0.00005;
-    [x3,y3,error3] = theory_simulations(eta,d);
-    
-    hFig = figure(1);
-    histogram(error2,'Normalization','cdf','DisplayName','Simulations \sigma_\eta=0.0002');
-    hold on
-    plot(x2,y2,'LineWidth',2,'DisplayName','Theory \sigma_\eta=0.0002');
-    hold on 
-    histogram(error3,'Normalization','cdf','DisplayName','Simulations \sigma_\eta=0.0008');
-    hold on
-    plot(x3,y3,'LineWidth',2,'DisplayName','Theory \sigma_\eta=0.0008');
-    hold on 
-    grid on
-    set(gca,'FontSize',16);
-    xlabel('\gamma');
-    ylabel('CDF');
-    xlim([-3 2]);
-    legend('show')
-    set(hFig, 'Position', [0 0 600 300])    
-    %Save Figure
-    %print(hFig,'sim-theory','-depsc2')
-    
-    eta = 8e-04;
-    d = 41*(1e-5);
-    eta_ss = eta^2;
-    d_ss = d^2;
-    %----------------------------------------------------
-    % period vs variance
-    %----------------------------------------------------
-    hFig = figure(2);
-    subplot(1,2,1);
-    alpha = 1;
-    period=[1:1:25];
-    variance = get_variance(eta_ss,d_ss,period,alpha);
-    plot(period,variance,'LineWidth',2);
-    grid on
-    set(gca,'FontSize',16);
-    xlabel({'Communication period \tau (second)'});
-    ylabel({'\sigma^2_\gamma (second^2)'});
-    xlim([1 25]);
-    ylim([0 15e-5])
-    %set(gca,'XTick',[0:10:100]);
-    %set(hFig, 'Position', [0 0 600 300]);
-    
-    % Save Figure
-    print(hFig,'period-vs-variance','-depsc2')
-
-    %----------------------------------------------------
-    % alpha vs variance
-    %----------------------------------------------------
[…]

### 2018-09-09 matlab/synchronization_accuracy.m

@@ -1,124 +1,0 @@
-clear all;
-close all;
-fid = fopen('testbed data/stable_voltage.txt','r');
-
-Period = 7086.0;   % Period of the event
-Elapsed = 0.0;  % The hardware time of the next event
-
-alpha = 1.0;     % Integral gain
-beta = 0.0001;
-
-m=1;
-
-while(~feof(fid))
-  A = fgetl(fid);
-  HC = hex2dec(A(1:4));     % Read hardware clock 
- 
-  Elapsed = HC;  % Expected next event time
-   
-  for k = 6:6
-      
-    HC = hex2dec(A(k*4+1:k*4+4));   % The hardware time of current event
-    
-    if(HC < Elapsed)
-        Elapsed = HC + 2^16 - Elapsed;
-    else
-        Elapsed = HC - Elapsed;
-    end
-    
-    E(m) = Period - Elapsed;             % The error between expected and actual time
-    
-    E1(m) = alpha*Elapsed - Period;
-    alpha = alpha - beta*E1(m) ;    
-    
-    m=m+1;
-    
-    end
-end
-
-E = abs(E);
-E1 = abs(E1);
-E1(E1 > 65000) = [];
-m1 = mean(E)
-m2 = mean(E1)
-
-hFig = figure(1);
-plot(E,'r--o','LineWidth',1);
-hold on
-plot(E1,'r-.','LineWidth',1);
-hold on
-
-fid = fopen('testbed data/unstable_voltage.txt','r');
-
-
-Period = 7118.0;   % Period of the event
-Elapsed = 0.0;  % The hardware time of the next event
-
-alpha = 1.0;     % Integral gain
-beta = 0.0001;
-
-m=1;
-
-while(~feof(fid))
-  A = fgetl(fid);
-  HC = hex2dec(A(1:4));     % Read hardware clock 
- 
-  Elapsed = HC;  % Expected next event time
-   
-  for k = 6:6
-      
-    HC = hex2dec(A(k*4+1:k*4+4));   % The hardware time of current event
-    
-    if(HC < Elapsed)
-        Elapsed = HC + 2^16 - Elapsed;
-    else
-        Elapsed = HC - Elapsed;
-    end
-    
-    E(m) = Period - Elapsed;             % The error between expected and actual time
-    
-    E1(m) = alpha*Elapsed - Period;
[…]
